---
title: "exploratory_covertype"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load data

```{r}
#load packages
library(ggplot2) #for plots
library(dplyr) #for working with dataframes
library(tidyr) #for working with datasets
library(lme4) #for mixed effects models
library(ggpubr) #for adding significance asterisks to graph

#read in data
data <- read.csv("06_cleandata/combined_plot_data.csv")

#check that variables are in appropriate format
str(data)

#change character columns to factors
data <- data %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     area = as.factor(area))

#filter data to only include within field data
data <- data %>%
  filter(plot_id != "d")

```

# Compare vegetation variables between grassy and bare

\% bare:

p=0.00057 but in wrong direction (more bare_perc in grassy sites) –\> likely due to dry (sprayed) veg, flooding and moss subtracting from bare_perc in bare sites

```{r}
ggplot(data, aes(x = cover_type, y = bare_perc)) + 
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Cover") +
  ylab("% Bare") +
  stat_compare_means(
    method = "t.test",        # or "wilcox.test"
    label = "p.format",       # shows formatted p-value
    size = 3.7,
    vjust = 0
  )

```

\% bare + flood+ bryo + dry

no significant (p=0.077) difference between cover type when I include dry rooted, flooding and bryophyte with bare

```{r}
ggplot(data, aes(x = cover_type, y = bare_flood_dry_bryo_perc)) + 
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Cover") +
  ylab("% Bare + Flood + Dry + Bryo") +
  stat_compare_means(
    method = "t.test",        # or "wilcox.test"
    label = "p.format",       # shows formatted p-value
    size = 3.7,
    vjust = 0
  )

```

\% grass: p=1e-11

```{r}
ggplot(data, aes(x=cover_type, y = grass_perc)) + 
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Cover") +
  ylab("% Grass") +
  stat_compare_means(
    method = "t.test",        # or "wilcox.test"
    label = "p.format",       # shows formatted p-value
    size = 3.7,
    vjust = 0
  )
```

\% grass by area

```{r}
ggplot(data, aes(x = area, fill=cover_type, y = grass_perc)) + 
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Area") +
  ylab("% Grass") +
  stat_compare_means(
    method = "t.test",        # or "wilcox.test"
    label = "p.format",       # shows formatted p-value
    size = 3.7,
    vjust = 0
  )
```

\% forb: p=3.7e-07

```{r}
ggplot(data, aes(x = cover_type, y = forb_perc)) + 
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Cover") +
  ylab("% Forb") +
  stat_compare_means(
    method = "t.test",        # or "wilcox.test"
    label = "p.format",       # shows formatted p-value
    size = 3.7,
    vjust = 0
  )
```

\% grass + forb: p=\<2e-16

```{r}
ggplot(data, aes(x = cover_type, y = grass_forb_perc)) + 
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Cover") +
  ylab("% grass + forb") +
  stat_compare_means(
    method = "t.test",        # or "wilcox.test"
    label = "p.format",       # shows formatted p-value
    size = 3.7,
    vjust = 0
  )
```

\% grass + forb by area

```{r}
ggplot(data, aes(x = area, fill = cover_type, y = grass_forb_perc)) + 
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Area") +
  ylab("% grass + forb") +
  stat_compare_means(
    method = "t.test",        # or "wilcox.test"
    label = "p.format",       # shows formatted p-value
    size = 3.7,
    vjust = 0
  )
```

grass h: p=1.8e-09

```{r}
ggplot(data, aes(x = cover_type, y = grass_h)) + 
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Cover") +
  ylab("Grass H") +
  stat_compare_means(
    method = "t.test",        # or "wilcox.test"
    label = "p.format",       # shows formatted p-value
    size = 3.7,
    vjust = 0
  )
```

forb h: p = 0.00057

```{r}
ggplot(data, aes(x = cover_type, y = forb_h)) + 
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Cover") +
  ylab("Forb H") +
  stat_compare_means(
    method = "t.test",        # or "wilcox.test"
    label = "p.format",       # shows formatted p-value
    size = 3.7,
    vjust = 0
  )
```

\% flood: p = 0.063

```{r}
ggplot(data, aes(x = round, fill = cover_type, y = flood_perc)) + 
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Round") +
  ylab("% Flood") +
  stat_compare_means(
    method = "t.test",        # or "wilcox.test"
    label = "p.format",       # shows formatted p-value
    size = 3.7,
    vjust = 0
  )
```

# Compare variables of interest between bare and grassy

num nsq 

```{r}
ggplot(data, aes(x=cover_type, y = nest_searching_queens))+ 
  geom_boxplot() +
  theme_classic(base_size = 15)+
  xlab("Cover")+
  ylab("NSQ abundance") +
  stat_compare_means(
    aes(group = cover_type),  # group by field.type
    method = "t.test",  # or "wilcox.test" if you're doing a non-parametric test
    label = "p.format",  # Display p-value 
    size = 3.7  # Adjust this value to make the asterisks larger
    ) 

```

NSQ broken up by season type

```{r}
ggplot(data, aes(fill=cover_type,x=bee_season_type, y = nest_searching_queens))+ 
  geom_boxplot() +
  theme_classic(base_size = 15)+
  xlab("Season")+
  ylab("NSQ abundance") +
  stat_compare_means(
    aes(group = cover_type),  # group by field.type
    method = "t.test",  # or "wilcox.test" if you're doing a non-parametric test
    label = "p.format",  # Display p-value 
    size = 3.7)  # Adjust this value to make the asterisks larger
```

num workers

```{r}
ggplot(data, aes(x=cover_type, y = workers))+ 
  geom_boxplot() +
  theme_classic(base_size = 15)+
  xlab("Cover")+
  ylab("Worker abundance") +
  stat_compare_means(
    aes(group = cover_type),  # group by field.type
    method = "t.test",  # or "wilcox.test" if you're doing a non-parametric test
    label = "p.format",  # Display p-value 
    size = 3.7)  # Adjust this value to make the asterisks larger

```

Workers broken up by season type

```{r}
ggplot(data, aes(fill=cover_type,x=bee_season_type, y = workers))+ 
  geom_boxplot() +
  theme_classic(base_size = 15)+
  xlab("Season")+
  ylab("Worker abundance") +
  stat_compare_means(
    aes(group = cover_type),  # group by field.type
    method = "t.test",  # or "wilcox.test" if you're doing a non-parametric test
    label = "p.format",  # Display p-value 
    size = 3.7)  # Adjust this value to make the asterisks larger
```

num burrows

```{r}
ggplot(data, aes(x=cover_type, y = num_burrows))+ 
  geom_boxplot() +
  theme_classic(base_size = 15)+
  xlab("Cover")+
  ylab("Burrow abundance") +
  stat_compare_means(
    aes(group = cover_type),  # group by field.type
    method = "t.test",  # or "wilcox.test" if you're doing a non-parametric test
    label = "p.format",  # Display p-value 
    size = 3.7)  # Adjust this value to make the asterisks larger
```

inactive burrows

```{r}
ggplot(data, aes(x=cover_type, y = num_inactive_burrows))+ 
  geom_boxplot() +
  theme_classic(base_size = 15)+
  xlab("Cover")+
  ylab("Inactive burrow abundance") +
  stat_compare_means(
    aes(group = cover_type),  # group by field.type
    method = "t.test",  # or "wilcox.test" if you're doing a non-parametric test
    label = "p.format",  # Display p-value 
    size = 3.7)  # Adjust this value to make the asterisks larger
```
Num burrows with runways (voles)
```{r}
ggplot(data, aes(x=cover_type, y = num_runway_burrows))+ 
  geom_boxplot() +
  theme_classic(base_size = 15)+
  xlab("Cover")+
  ylab("Runway burrow abundance") +
  stat_compare_means(
    aes(group = cover_type),  # group by field.type
    method = "t.test",  # or "wilcox.test" if you're doing a non-parametric test
    label = "p.format",  # Display p-value 
    size = 3.7)  # Adjust this value to make the asterisks larger
```


floral abundance (relevant to bees)

```{r}
ggplot(data, aes(x=cover_type, y = bombus_floral_abun))+ 
  geom_boxplot() +
  theme_classic(base_size = 15)+
  xlab("Cover")+
  ylab("Bombus-relevant floral abundance") +
  stat_compare_means(
    aes(group = cover_type),  # group by field.type
    method = "t.test",  # or "wilcox.test" if you're doing a non-parametric test
    label = "p.format",  # Display p-value 
    size = 3.7)  # Adjust this value to make the aste