---
title: "ch2_extra"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load data

```{r}
#load packages
library(ggplot2) #for plots
library(dplyr) #for working with dataframes
library(tidyr) #for working with datasets
library(lme4) #for mixed effects models
library(ggpubr) #for adding significance asterisks to graph
library(stringr)
library(lubridate)

#read in data
data <- read.csv("06_cleandata/combined_plot_data.csv")
rodent <- read.csv("06_cleandata/rodent_clean.csv")

#change character columns to factors
data <- data %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     area = as.factor(area))

#add column for ditch vs field
##in overall data
data <- data %>%
  mutate(field_ditch = case_when(
    str_detect(plot_id, "p1|p2") ~ "field",
    str_detect(plot_id, "d") ~ "ditch",
    TRUE ~ NA_character_
  ))

##in rodent data
rodent <- rodent %>%
  mutate(field_ditch = case_when(
    str_detect(plot_id, "p1|p2") ~ "field",
    str_detect(plot_id, "d") ~ "ditch",
    TRUE ~ NA_character_
  ))
```

# Bumble bee floral info

## Load data

```{r}
bee <- read.csv("06_cleandata/bee_clean.csv")
##add ditch vs field column
bee <- bee %>%
  mutate(field_ditch = case_when(
    str_detect(plot_id, "p1|p2") ~ "field",
    str_detect(plot_id, "d") ~ "ditch",
    TRUE ~ NA_character_
  ))


#add column for julian date
bee$julian.date <- yday(bee$date)

#filter to queens and workers
queens <- bee %>%
  filter(caste == "q", flower_species != "" & !is.na(flower_species))
workers <- bee %>%
  filter(caste == "w", flower_species != "" & !is.na(flower_species))
```

## Flower species used by bees

### Overall

Queens

```{r}
queens %>%
  count(flower_species, name = "count") %>%
  mutate(flower_species = reorder(flower_species, -count)) %>%
  ggplot(aes(flower_species, count)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = count), vjust = -0.15, size = 3) +
  theme_classic(base_size = 15) +
  xlab("Floral species") +
  ylab("Queen count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

Workers

```{r}
workers %>%
  count(flower_species, name = "count") %>%
  mutate(flower_species = reorder(flower_species, -count)) %>%
  ggplot(aes(flower_species, count)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = count), vjust = -0.15, size = 3) +
  theme_classic(base_size = 15) +
  xlab("Floral species") +
  ylab("Worker count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### Separated by plot type (ditch vs field)

Queens

```{r}
queens %>%
  count(flower_species, field_ditch, name = "count") %>%
  mutate(adjusted_count = case_when( ##adjust count for sampling effort (2x as many field plots as ditch plots)
    field_ditch == "field" ~ count * 1,
    field_ditch == "ditch" ~ count * 2
  )) %>%
  mutate(flower_species = reorder(flower_species, -adjusted_count)) %>%
  ggplot(aes(flower_species, adjusted_count, fill = field_ditch)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = adjusted_count), 
            position = position_dodge(width = 0.9),
            vjust = -0.15, 
            size = 3) +
  theme_classic(base_size = 15) +
  xlab("Floral species") +
  ylab("Queen count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

Workers

```{r}
workers %>%
  count(flower_species, field_ditch, name = "count") %>%
  mutate(adjusted_count = case_when( ##adjust count for sampling effort (2x as many field plots as ditch plots)
    field_ditch == "field" ~ count * 1,
    field_ditch == "ditch" ~ count * 2
  )) %>%
  mutate(flower_species = reorder(flower_species, -adjusted_count)) %>%
  ggplot(aes(flower_species, adjusted_count, fill = field_ditch)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = adjusted_count), 
            position = position_dodge(width = 0.9),
            vjust = -0.15, 
            size = 3) +
  theme_classic(base_size = 15) +
  xlab("Floral species") +
  ylab("Workers count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

Plot \# foraging bees in ditch vs field across time

```{r}

# Step 1: Filter and summarize data
foraging_summary <- bee %>%
  filter(behaviour == "foraging") %>%
  count(julian.date, field_ditch, name = "n_foraging")

# Step 2: Plot
ggplot(foraging_summary, aes(x = julian.date, y = n_foraging, color = field_ditch)) +
  geom_line(size = 1.2) +
  labs(
    title = "# Bombus Foraging vs. Julian Date",
    x = "Julian Date",
    y = "# Foraging Bombus",
    color = "Plot Type"
  ) +
  theme_classic()
```

Queens only

```{r}
# Step 1: Filter and summarize data
foraging_queen_summary <- queens %>%
  filter(behaviour == "foraging") %>%
  count(julian.date, field_ditch, name = "n_foraging")

# Step 2: Plot
ggplot(foraging_queen_summary, aes(x = julian.date, y = n_foraging, color = field_ditch)) +
  geom_line(size = 1.2) +
  labs(
    title = "# Bombus Foraging vs. Julian Date",
    x = "Julian Date",
    y = "# Foraging Queens",
    color = "Plot Type"
  ) +
  theme_classic()
```

Workers only

```{r}
# Step 1: Filter and summarize data
foraging_worker_summary <- workers %>%
  filter(behaviour == "foraging") %>%
  count(julian.date, field_ditch, name = "n_foraging")

# Step 2: Plot
ggplot(foraging_worker_summary, aes(x = julian.date, y = n_foraging, color = field_ditch)) +
  geom_line(size = 1.2) +
  labs(
    title = "# Bombus Foraging vs. Julian Date",
    x = "Julian Date",
    y = "# Foraging workers",
    color = "Plot Type"
  ) +
  theme_classic()
```

Plot number foraging vs time, by caste

```{r}
# Step 1: Filter and summarize data
foraging_caste_summary <- bee %>%
  filter(behaviour == "foraging") %>%
  filter(caste != "")%>%
  count(julian.date, caste, name = "n_foraging")

# Step 2: Plot
ggplot(foraging_caste_summary, aes(x = julian.date, y = n_foraging, color = caste)) +
  geom_line(size = 1.2) +
  labs(
    title = "# Bombus Foraging vs. Julian Date",
    x = "Julian Date",
    y = "# Foraging Bombus",
    color = "Caste"
  ) +
  theme_classic()
```

Floral availability of VACO over time

```{r}
#load data
floral <- read.csv("06_cleandata/floral_clean.csv")

#add column for julian date
floral$julian.date <- yday(floral$date)

# Step 1: Filter and summarize data
vaco_summary <- floral %>%
  filter(flower_code == "vaco") %>%
  group_by(julian.date) %>%
  summarise(vaco_abun = sum(abundance), .groups = "drop") 

# Step 2: Plot
ggplot(vaco_summary, aes(x = julian.date, y = vaco_abun)) +
  geom_line(size = 1.2) +
  labs(
    title = "VACO abundance vs Julian Date",
    x = "Julian Date",
    y =  "VACO abundance"
  ) +
  theme_classic()
```

Number of bees foraging on vaco over time, by caste

```{r}
# Step 1: Filter and summarize data
vacoforaging_caste_summary <- bee %>%
  filter(behaviour == "foraging") %>%
  filter(flower_code == "vaco") %>%
  filter(caste != "")%>%
  count(julian.date, caste, name = "n_foraging_vaco")

# Step 2: Plot
ggplot(vacoforaging_caste_summary, aes(x = julian.date, y = n_foraging_vaco, color = caste)) +
  geom_line(size = 1.2) +
  labs(
    title = "# Bombus Foraging on VACO vs. Julian Date",
    x = "Julian Date",
    y = "# Bombus foraging on vaco",
    color = "Caste"
  ) +
  theme_classic()
```

Compare number of queens vs males vs workers foraging on vaco

```{r}
# Step 1: Filter and summarize data
vacoforaging_caste_summary <- bee %>%
  filter(behaviour == "foraging") %>%
  filter(flower_code == "vaco") %>%
  filter(caste != "")%>%
  count(caste, name = "n_foraging_vaco")

# Step 2: Plot
ggplot(vacoforaging_caste_summary, aes(x = caste, y = n_foraging_vaco, fill = caste)) +
  geom_bar(stat = "identity") +
  labs(
    title = "# Bombus Foraging on VACO vs. Caste",
    x = "Caste",
    y = "# Bombus foraging on vaco"
  ) +
  theme_classic()
```

# Number of individuals observed 

## Bees

```{r}
#load bee data
bee <- read.csv("06_cleandata/bee_clean.csv")

#add column for ditch vs field
bee <- bee %>%
  mutate(field_ditch = case_when(
    str_detect(plot_id, "p1|p2") ~ "field",
    str_detect(plot_id, "d") ~ "ditch",
    TRUE ~ NA_character_
  ))

##counts per caste
bee_count_summary <- bee %>%
  #change blanks to unknown
  mutate(caste = ifelse(caste == "" , "unknown", caste)) %>%
  # Group by species and caste
  group_by(caste) %>%
  summarise(n = n(), .groups = "drop") %>%
  # Add a total row
  bind_rows(tibble(caste = "overall", n = sum(.$n)))

##counts per plot type 
bee_count_summary_plot <- bee %>%
  # Group by plot type 
  group_by(field_ditch) %>%
  summarise(n = n(), .groups = "drop") %>%
  # Add a total row
  bind_rows(tibble(field_ditch = "overall", n = sum(.$n)))
  
#number of species
bee_species_summary <- bee %>%
  #change blanks to unknown
  mutate(caste = ifelse(caste == "" , "unknown", caste)) %>%
  mutate(bee_species = ifelse(bee_species == "" , "unknown", bee_species)) %>%
  # Group by species and caste
  group_by(bee_species, caste) %>%
  summarise(n = n(), .groups = "drop") %>%
  # Pivot caste levels into columns
  pivot_wider(
    names_from = caste,
    values_from = n,
    values_fill = 0  # fill missing combinations with 0
  ) %>%
  # Add a total count column
  mutate(n_overall = rowSums(across(-bee_species)))

#number of species per plot type
bee_species_summary_plot <- bee %>%
  #change blanks to unknown
  mutate(bee_species = ifelse(bee_species == "" , "unknown", bee_species)) %>%
  # Group by species and caste
  group_by(bee_species, field_ditch) %>%
  summarise(n = n(), .groups = "drop") %>%
  # Pivot caste levels into columns
  pivot_wider(
    names_from = field_ditch,
    values_from = n,
    values_fill = 0  # fill missing combinations with 0
  ) %>%
  # Add a total count column
  mutate(n_overall = rowSums(across(-bee_species)))

```

## Rodents

```{r}
#load bee data
rodent <- read.csv("06_cleandata/rodent_clean.csv")

#number of individuals
total_rodent <- count(rodent) ##1037 across all sites
#average per plot
rodent_plot_summary <- rodent %>%
  # Group by plot and site
  group_by(site_id, plot_id) %>%
  summarise(n = n(), .groups = "drop") %>%
  # Pivot caste levels into columns
  pivot_wider(
    names_from = site_id,
    values_from = n,
    values_fill = 0  # fill missing combinations with 0
  )
```

## Floral species

```{r}
floral <- read.csv("06_cleandata/floral_clean.csv")

beefloral <- read.csv("06_cleandata/useful_flowers.csv")

floral_bombus <- floral %>%
  filter(flower_species %in% beefloral$flower_species |
           str_detect(flower_species, "brassica_") |
           str_detect(flower_species, "rhododendron_") |
           str_detect(flower_species, "rorippa_") |
           str_detect(flower_species, "rubus_") |
           str_detect(flower_species, "rosa_") |
           str_detect(flower_species, "calystegia_") |
           str_detect(flower_species, "prunus_") |
           str_detect(flower_species, "phacelia_") 
  )

#count number of floral species:
length(unique(floral_bombus$flower_species)) ##53
```
