---
title: "ch2_exploratory"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

Load and clean data

```{r}
#load packages
library(tidyverse) #data manipulation, visualization etc. (dplyr, ggplot2, tidyr)
library(DHARMa) #diagnostic tools for mixed regression models
library(glmmTMB) #fits generalized linear mixed models, including zero-inflated and overdispersed models
library(car) #tools for regression diagnostics and hypothesis tests
library(emmeans)   # for estimated marginal means (predictions)

#read in data
data <- read.csv("06_cleandata/combined_plot_data.csv")

#change character columns to factors
surveys <- data %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     bee_season_type = as.factor(bee_season_type),
     area = as.factor(area),
     cover_type = as.factor(cover_type))

#add plot type column and make it a factor
surveys <- surveys %>% 
  mutate(plot_type = if_else(plot_id == "d", "ditch", "field"))
surveys <- surveys %>%
   mutate(
     plot_type = as.factor(plot_type))

#scale continuous predictors
surveys$julian.date_scaled <- as.numeric(scale(surveys$julian.date))
surveys$bare_perc_scaled <- as.numeric(scale(surveys$bare_perc))
surveys$grass_perc_scaled <- as.numeric(scale(surveys$grass_perc))
surveys$forb_perc_scaled <- as.numeric(scale(surveys$forb_perc))
surveys$grass_forb_perc_scaled <- as.numeric(scale(surveys$grass_forb_perc))
surveys$grass_h_scaled <- as.numeric(scale(surveys$grass_h))
surveys$forb_h_scaled <- as.numeric(scale(surveys$forb_h))
surveys$floral_perc_scaled <- as.numeric(scale(surveys$floral_perc))
surveys$grass_forb_floral_perc_scaled <- as.numeric(scale(surveys$grass_forb_floral_perc))
surveys$bombus_floral_perc_scaled <- as.numeric(scale(surveys$bombus_floral_perc))
surveys$flood_perc_scaled <- as.numeric(scale(surveys$flood_perc))
surveys$dry_root_perc_scaled <- as.numeric(scale(surveys$dry_root_perc))
surveys$bryo_perc_scaled <- as.numeric(scale(surveys$bryo_perc))
surveys$bare_flood_dry_bryo_perc_scaled <- as.numeric(scale(surveys$bare_flood_dry_bryo_perc))
surveys$floral_richness_scaled <- as.numeric(scale(surveys$floral_richness))
surveys$bombus_floral_richness_scaled <- as.numeric(scale(surveys$bombus_floral_richness))
surveys$bombus_floral_abun_scaled <- as.numeric(scale(surveys$bombus_floral_abun))
surveys$nest_searching_queens_scaled <- as.numeric(scale(surveys$nest_searching_queens))
surveys$workers_scaled <- as.numeric(scale(surveys$workers))
surveys$forage_vaco_scaled <- as.numeric(scale(surveys$forage_vaco))
surveys$num_burrows_scaled <- as.numeric(scale(surveys$num_burrows))
surveys$num_inactive_burrows_scaled <- as.numeric(scale(surveys$num_inactive_burrows))
surveys$num_active_dm_scaled <- as.numeric(scale(surveys$num_active_dm))
surveys$num_dm_size_scaled <- as.numeric(scale(surveys$num_dm_size))
surveys$num_inactive_dm_scaled <- as.numeric(scale(surveys$num_inactive_dm))
```

Look at composition of ditches vs within field plots composition at grassy vs bare sites

-   \% grass and forb

LM

```{r}
#adjust % grass and forb so that all values are positive
surveys$grass_forb_perc_adj <- surveys$grass_forb_perc + 0.001

#model with gamma distribution
ditchVSfield_veg_full <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled +
                              (1|area/site_id/plot_id),
                            family = Gamma(link = "log"), 
                            data = surveys)

ditchVSfield_veg_noArea <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled +
                              (1|site_id/plot_id),
                            family = Gamma(link = "log"), 
                            data = surveys)

ditchVSfield_veg_plotOnly <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled +
                              (1|plot_id),
                            family = Gamma(link = "log"), 
                            data = surveys)

ditchVSfield_veg_siteOnly <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled +
                              (1|site_id),
                            family = Gamma(link = "log"), 
                            data = surveys)

ditchVSfield_veg_siteArea <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled +
                              (1|area/site_id),
                            family = Gamma(link = "log"), 
                            data = surveys)

ditchVSfield_veg_areaPlot <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled +
                              (1|area/plot_id),
                            family = Gamma(link = "log"), 
                            data = surveys)

ditchVSfield_veg_noRE <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled,
                            family = Gamma(link = "log"), 
                            data = surveys)

#compare models
anova(ditchVSfield_veg_full, ditchVSfield_veg_noArea, ditchVSfield_veg_plotOnly, ditchVSfield_veg_siteOnly, ditchVSfield_veg_siteArea,
      ditchVSfield_veg_areaPlot, ditchVSfield_veg_noRE)
##siteArea is best
```

Check diagnostics

```{r}
#simulate residuals
residuals_ditchVSfield_veg_siteArea <- simulateResiduals(fittedModel = ditchVSfield_veg_siteArea, plot = TRUE)

##results indicate plot is not a good fit.
```

Try log-transforming the data (log(x+1) and running a Gaussian model

```{r}
##Log transform and add one to account for zeros.
surveys$log_grass_forb <- log(surveys$grass_forb_perc + 1)

##Model with gaussian distribution
ditchVSfield_veg_full <- glmmTMB(log_grass_forb ~ cover_type*plot_type + julian.date_scaled +
                              (1|area/site_id/plot_id),
                            family = gaussian(), 
                            data = surveys)
ditchVSfield_veg_noArea <- glmmTMB(log_grass_forb ~ cover_type*plot_type + julian.date_scaled +
                              (1|site_id/plot_id),
                            family = gaussian(), 
                            data = surveys)

ditchVSfield_veg_plotOnly <- glmmTMB(log_grass_forb ~ cover_type*plot_type + julian.date_scaled +
                              (1|plot_id),
                            family = gaussian(), 
                            data = surveys)

ditchVSfield_veg_siteOnly <- glmmTMB(log_grass_forb ~ cover_type*plot_type + julian.date_scaled +
                              (1|site_id),
                            family = gaussian(), 
                            data = surveys)
ditchVSfield_veg_siteArea <- glmmTMB(log_grass_forb ~ cover_type*plot_type + julian.date_scaled +
                              (1|area/site_id),
                            family = gaussian(), 
                            data = surveys)

ditchVSfield_veg_areaPlot <- glmmTMB(log_grass_forb ~ cover_type*plot_type + julian.date_scaled +
                              (1|area/plot_id),
                            family = gaussian(), 
                            data = surveys)

ditchVSfield_veg_noRE <- glmmTMB(log_grass_forb ~ cover_type*plot_type + julian.date_scaled,
                            family = gaussian(), 
                            data = surveys)

#compare models
anova(ditchVSfield_veg_full, ditchVSfield_veg_noArea, ditchVSfield_veg_plotOnly, ditchVSfield_veg_siteOnly, ditchVSfield_veg_siteArea,
      ditchVSfield_veg_areaPlot, ditchVSfield_veg_noRE)
##siteArea is best
```

Check diagnostics

```{r}
#simulate residuals
residuals_ditchVSfield_veg_siteArea <- simulateResiduals(fittedModel = ditchVSfield_veg_siteArea, plot = TRUE)

##model fit is good enough.  
```

Plot results

```{r}
summary(ditchVSfield_veg_siteArea)
```

Plot

```{r}
ggplot(surveys, aes(y = grass_forb_perc_scaled, x = plot_type, color = cover_type)) +
  geom_boxplot() +
  labs(x = "Plot type",
       y = "% grass and forb") +
  theme_classic(base_size = 23)
```

Floral abundance model

```{r}
#Load packages needed for CLM
library(ordinal)
library(MASS)

surveys$bombus_floral_abun <- factor(surveys$bombus_floral_abun, ordered = TRUE)

floral_full <- clmm(bombus_floral_abun ~ cover_type*plot_type +
                         julian.date_scaled*plot_type + 
                         (1|area/site_id/plot_id),
                       link = "logit", 
                       data = surveys)

floral_noArea <- clmm(bombus_floral_abun ~ cover_type*plot_type +
                         julian.date_scaled*plot_type + 
                         (1|site_id/plot_id),
                       link = "logit", 
                       data = surveys)

#floral_plotOnly <- clmm(bombus_floral_abun ~ plot_type +
 #                        julian.date_scaled + 
  #                       (1|plot_id),
   #                    link = "logit", 
    #                   data = surveys)
##can't use since there are only 2 levels

floral_siteOnly <- clmm(bombus_floral_abun ~ cover_type*plot_type +
                         julian.date_scaled*plot_type + 
                         (1|site_id),
                       link = "logit", 
                       data = surveys)

floral_AreaPlot <- clmm(bombus_floral_abun ~ cover_type*plot_type +
                         julian.date_scaled*plot_type + 
                         (1|area/plot_id),
                       link = "logit", 
                       data = surveys)
floral_noRE <- polr(bombus_floral_abun ~ cover_type*plot_type +
                         julian.date_scaled*plot_type,
                       method = "logistic", 
                       data = surveys)

#compare models AIC
AIC(floral_full, floral_noArea, floral_siteOnly, floral_AreaPlot, floral_noRE)
##shows that model without random effect has best fit
```

Check output

```{r}
summary(floral_noRE)
```

Plot floral abundance vs cover type and plot type

```{r}
ggplot(surveys, aes(y = bombus_floral_abun_scaled, x = plot_type, color = cover_type)) +
  geom_boxplot() +
  labs(x = "Plot type",
       y = "Floral abun") +
  theme_classic(base_size = 23)
```

Plot floral abundance vs season
