---
title: "glmCH2_clean"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load and clean data

```{r}
#load packages
library(tidyverse) #data manipulation, visualization etc. (dplyr, ggplot2, tidyr)
library(DHARMa) #diagnostic tools for mixed regression models
library(glmmTMB) #fits generalized linear mixed models, including zero-inflated and overdispersed models
library(car) #tools for regression diagnostics and hypothesis tests
library(emmeans)   # for estimated marginal means (predictions)

#read in data
data <- read.csv("06_cleandata/combined_plot_data.csv")

#change character columns to factors
surveys <- data %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     bee_season_type = as.factor(bee_season_type),
     area = as.factor(area))

#add plot type column and make it a factor
surveys <- surveys %>% 
  mutate(plot_type = if_else(plot_id == "d", "ditch", "field"))
surveys <- surveys %>%
   mutate(
     plot_type = as.factor(plot_type))

#scale continuous predictors
surveys$julian.date_scaled <- as.numeric(scale(surveys$julian.date))
surveys$bare_perc_scaled <- as.numeric(scale(surveys$bare_perc))
surveys$grass_perc_scaled <- as.numeric(scale(surveys$grass_perc))
surveys$forb_perc_scaled <- as.numeric(scale(surveys$forb_perc))
surveys$grass_forb_perc_scaled <- as.numeric(scale(surveys$grass_forb_perc))
surveys$grass_h_scaled <- as.numeric(scale(surveys$grass_h))
surveys$forb_h_scaled <- as.numeric(scale(surveys$forb_h))
surveys$floral_perc_scaled <- as.numeric(scale(surveys$floral_perc))
surveys$grass_forb_floral_perc_scaled <- as.numeric(scale(surveys$grass_forb_floral_perc))
surveys$bombus_floral_perc_scaled <- as.numeric(scale(surveys$bombus_floral_perc))
surveys$flood_perc_scaled <- as.numeric(scale(surveys$flood_perc))
surveys$dry_root_perc_scaled <- as.numeric(scale(surveys$dry_root_perc))
surveys$bryo_perc_scaled <- as.numeric(scale(surveys$bryo_perc))
surveys$bare_flood_dry_bryo_perc_scaled <- as.numeric(scale(surveys$bare_flood_dry_bryo_perc))
surveys$floral_richness_scaled <- as.numeric(scale(surveys$floral_richness))
surveys$bombus_floral_richness_scaled <- as.numeric(scale(surveys$bombus_floral_richness))
surveys$bombus_floral_abun_scaled <- as.numeric(scale(surveys$bombus_floral_abun))
surveys$nest_searching_queens_scaled <- as.numeric(scale(surveys$nest_searching_queens))
surveys$workers_scaled <- as.numeric(scale(surveys$workers))
surveys$forage_vaco_scaled <- as.numeric(scale(surveys$forage_vaco))
surveys$num_burrows_scaled <- as.numeric(scale(surveys$num_burrows))
surveys$num_inactive_burrows_scaled <- as.numeric(scale(surveys$num_inactive_burrows))
surveys$num_active_dm_scaled <- as.numeric(scale(surveys$num_active_dm))
surveys$num_dm_size_scaled <- as.numeric(scale(surveys$num_dm_size))
surveys$num_inactive_dm_scaled <- as.numeric(scale(surveys$num_inactive_dm))
```

# Choosing which vegetation variables to include as predictors

1)  Check for multicollinearity among vegetation variables.

-   All variables have VIF \<5 (no multicollinearity), so i can include all these predictors in my model

```{r}
vif(lm(nest_searching_queens ~ bare_perc_scaled + 
         grass_forb_perc_scaled + 
        grass_h_scaled + 
         forb_h_scaled + 
         bombus_floral_perc_scaled +
         flood_perc_scaled +
         dry_root_perc_scaled +
         bombus_floral_richness_scaled +
         bombus_floral_abun_scaled + bee_season_type, 
        data = surveys))
```

2)  Use PCA to determine which vegetation variables to include in my models. I don't want to include all of the predictors, since some will be correlated with each other and therefore dilute their effects of the response variables. I am using PCA so I can pick one variable that drives each of the main 3 components, and use these as the predictors in my model.

```{r}
pca_res <- prcomp(surveys[, c("bare_perc_scaled",
                              "grass_forb_perc_scaled",
                              "grass_h_scaled",
                              "forb_h_scaled",
                              "bombus_floral_perc_scaled",
                              "flood_perc_scaled",
                              "dry_root_perc_scaled",
                              "bombus_floral_richness_scaled",
                              "bombus_floral_abun_scaled")], scale. = TRUE)
summary(pca_res)

#see which variable drives each of the principle components
print(pca_res$rotation)
```

-   PCA Results:

    PC1: grass_h_scaled (-0.50),forb_h_scaled (-0.48),grass_forb_perc_scaled (-0.42),bombus_floral_richness_scaled (-0.43),bombus_floral_abun_scaled (-0.22)

    PC2: bare_perc_scaled (0.51),bombus_floral_abun_scaled (0.52),bombus_floral_perc_scaled (0.40), dry_root_perc_scaled (-0.40)

    Based on these PCA results and my research question, I am selecting to include the following predictors in my model:

    -   grass_forb_perc_scaled (loads strongly on PCA1)

    -   bombus_floral_abun_scaled (loads strongly on PCA1 &2)

3)  Check pairwise correlations between this reduced set of variables

-   correlation among all the pairs are all near zero --\> predictors are all sufficiently independent! (R\<\<0.7)

```{r}
cor(surveys[, c("grass_forb_perc_scaled", "bombus_floral_abun_scaled")])
```

# Nest-searching queens

## [vs plot type (ditch vs field)]{.underline}

### Steps

1)  Plotting a zero-inflated poisson model with multiple different random effect structures to see which is best. I used a zero-inflated poisson model since #NSQ is count data with a lot of zeros.

-   site-only was best according to maximum likelihood tests

```{r}
nsq_glm_plotType <- glmmTMB(nest_searching_queens ~ plot_type +
                              bee_season_type +
                              (1|area/site_id/plot_id),
                            ziformula = ~plot_type +
                              bee_season_type +
                              (1|area/site_id/plot_id),  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

nsq_glm_plotType_noArea <- glmmTMB(nest_searching_queens ~ plot_type +
                              bee_season_type +
                              (1|site_id/plot_id),
                            ziformula = ~plot_type +
                              bee_season_type +
                              (1|site_id/plot_id),  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

nsq_glm_plotType_plotOnly <- glmmTMB(nest_searching_queens ~ plot_type +
                              bee_season_type +
                              (1|plot_id),
                            ziformula = ~plot_type +
                              bee_season_type +
                              (1|plot_id),  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

nsq_glm_plotType_siteOnly <- glmmTMB(nest_searching_queens ~ plot_type +
                                       bee_season_type +
                                       (1|site_id),
                            ziformula = ~plot_type +
                              bee_season_type +
                              (1|site_id),  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

nsq_glm_plotType_areaSite <- glmmTMB(nest_searching_queens ~ plot_type + bee_season_type +
                              (1|area/site_id),
                            ziformula = ~plot_type +
                              bee_season_type +
                              (1|area/site_id),  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

#Compare models
anova(nsq_glm_plotType,nsq_glm_plotType_noArea,nsq_glm_plotType_plotOnly, nsq_glm_plotType_siteOnly,nsq_glm_plotType_areaSite)
##siteOnly is best
```

2)  Check diagnostics of site-only model

-   look good

```{r}
#simulate residuals
residuals_nsq_glm_plotType_siteOnly <- simulateResiduals(fittedModel = nsq_glm_plotType_siteOnly, plot = TRUE)
```

3)  Plot results

```{r}
summary(nsq_glm_plotType_siteOnly)
```

### Plots

Plot type

Conditional model:

```{r}
# Create newdata with discrete factor levels for plot_type
nsq_data_plot <- data.frame(
  bee_season_type = "nestsearching",
  plot_type = factor(c("field", "ditch"), levels = levels(surveys$plot_type))
)

# Predictions for conditional model (count)
pred_nsq_plot_cond <- predict(nsq_glm_plotType_siteOnly, 
                              newdata = nsq_data_plot, 
                              type = "conditional", 
                              se.fit = TRUE,
                              re.form = NA)

# Predictions for zero-inflation model
pred_nsq_plot_zi <- predict(nsq_glm_plotType_siteOnly, 
                            newdata = nsq_data_plot, 
                            type = "zprob", 
                            se.fit = TRUE,
                            re.form = NA)

# Combine predictions with newdata
pred_nsq_plot_df <- cbind(nsq_data_plot,
                 cond_fit = pred_nsq_plot_cond$fit,
                 cond_se = pred_nsq_plot_cond$se.fit,
                 zi_fit = pred_nsq_plot_zi$fit,
                 zi_se = pred_nsq_plot_zi$se.fit)

# Plot conditional predictions
ggplot(pred_nsq_plot_df, aes(x = plot_type, y = cond_fit)) +
  geom_point(size=3) +
  geom_errorbar(aes(ymin = cond_fit - 1.96*cond_se, ymax = cond_fit + 1.96*cond_se), width = 0.2, position = position_dodge(0.6))+
  labs(title = "Conditional Model Predictions (Log-scale)", y = "Predicted NSQ count (log-scale)") +
  theme_classic()
```

Zero-inflation model:

```{r}
ggplot(pred_nsq_plot_df, aes(x = plot_type, y = zi_fit)) +
  geom_point(size=3) +
  geom_errorbar(aes(ymin = zi_fit - 1.96*zi_se, ymax = zi_fit + 1.96*zi_se), width = 0.2, position = position_dodge(0.6))+
  labs(title = "Zero-inflation Model Predictions (Presence/Absence)", y = "Predicted Probability of Structural Zero") +
  theme_classic()
```

Season

Conditional:

```{r}
# Create newdata with discrete factor levels for plot_type
nsq_data_season <- data.frame(
  bee_season_type = factor(c("nestsearching", "worker"), levels = levels(surveys$bee_season_type)),
  plot_type = "field")

# Predictions for conditional model (count)
pred_nsq_season_cond <- predict(nsq_glm_plotType_siteOnly, 
                              newdata = nsq_data_season, 
                              type = "conditional", 
                              se.fit = TRUE,
                              re.form = NA)

# Predictions for zero-inflation model
pred_nsq_season_zi <- predict(nsq_glm_plotType_siteOnly, 
                            newdata = nsq_data_season, 
                            type = "zprob", 
                            se.fit = TRUE,
                            re.form = NA)

# Combine predictions with newdata
pred_nsq_season_df <- cbind(nsq_data_season,
                 cond_fit = pred_nsq_season_cond$fit,
                 cond_se = pred_nsq_season_cond$se.fit,
                 zi_fit = pred_nsq_season_zi$fit,
                 zi_se = pred_nsq_season_zi$se.fit)

# Plot conditional predictions
ggplot(pred_nsq_season_df, aes(x = bee_season_type, y = cond_fit)) +
  geom_point(size=3) +
  geom_errorbar(aes(ymin = cond_fit - 1.96*cond_se, ymax = cond_fit + 1.96*cond_se), width = 0.2, position = position_dodge(0.6))+
  labs(title = "Conditional Model Predictions (Log-scale)", y = "Predicted NSQ count (log-scale)") +
  theme_classic()
```

Zero-inflation:

```{r}
ggplot(pred_nsq_season_df, aes(x = bee_season_type, y = zi_fit)) +
  geom_point(size=3) +
  geom_errorbar(aes(ymin = zi_fit - 1.96*zi_se, ymax = zi_fit + 1.96*zi_se), width = 0.2, position = position_dodge(0.6))+
  labs(title = "Zero-inflation Model Predictions (Presence/Absence)", y = "Predicted Probability of Structural Zero") +
  theme_classic()
```

### Findings

Conditional model (plots NSQ abundance at sites where there at NSQ)

-   Significantly higher abundance of NSQ in ditch plots than in field plots.

Zero-inflation model (predicts presence/absence of NSQ at sites)

-   Significantly more zeros in worker season than in nest-searching season.

## [vs vegetation predictors]{.underline}

Plot model with interaction between vegetation variables and plot type to investigate whether the effects of vegetation cover and floral abundance on queen abundance/presence change depending on plot type (ditch vs field).

### Steps

1)  Try different random effect structures.

-   Model without random effect fit data the best.

```{r}
nsq_glm_veg_full <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled*plot_type+
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type +
                              (1|area/site_id/plot_id),
                            ziformula = ~grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type +
                              (1|area/site_id/plot_id),  # zero-inflation model
                            family = poisson, 
                            data = surveys)

nsq_glm_veg_noArea <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type+
                              bee_season_type+
                              (1|site_id/plot_id),
                            ziformula = ~grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type +
                              (1|site_id/plot_id),  # zero-inflation model
                            family = poisson, 
                            data = surveys)

nsq_glm_veg_justSite <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type+
                              (1|site_id),
                            ziformula = ~grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type+
                              bee_season_type +
                              (1|site_id),  # zero-inflation model
                            family = poisson,
                            data = surveys)

nsq_glm_veg_noRE <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type,
                            ziformula = ~grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type,  # zero-inflation model
                            family = poisson,
                            data = surveys)

#compare models
anova(nsq_glm_veg_noRE, nsq_glm_veg_justSite, nsq_glm_veg_noArea, nsq_glm_veg_full)
##no random effect model is best, followed by just site model
```

2)  Check diagnostics of model

-   look good

```{r}
#simulate residuals
residuals_nsq_glm_veg_noRE <- simulateResiduals(fittedModel = nsq_glm_veg_noRE, plot = TRUE)
```

3)  Plot results

```{r}
summary(nsq_glm_veg_noRE)
```

### Plots

Grass and forb % cover

Conditional model:

```{r}
# Define a sequence of values for grass_forb_perc_scaled across its observed range
grass_seq <- seq(
  from = min(surveys$grass_forb_perc_scaled, na.rm = TRUE),
  to = max(surveys$grass_forb_perc_scaled, na.rm = TRUE),
  length.out = 50
)

#create new data gri
nsq_data_veg<- expand.grid(
  grass_forb_perc_scaled = grass_seq,
  plot_type = unique(surveys$plot_type),
  bombus_floral_abun_scaled = mean(surveys$bombus_floral_abun_scaled, na.rm = TRUE),
  bee_season_type = "nestsearching")


# Get predicted conditional probabilities and standard errors
pred_nsq_veg_cond <- predict(nsq_glm_veg_noRE,
                            newdata = nsq_data_veg,
                            type = "conditional",
                            se.fit = TRUE)

# Combine predictions with data frame
pred_nsq_rodent_df <- cbind(nsq_data_veg,
                 cond_fit = pred_nsq_veg_cond$fit,
                 cond_se = pred_nsq_veg_cond$se.fit)

ggplot(pred_nsq_rodent_df, aes(x = grass_forb_perc_scaled, y = cond_fit, color = plot_type)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = cond_fit - 1.96*cond_se, ymax = cond_fit + 1.96*cond_se), alpha = 0.2) +
  labs(title = "Conditional Model Predictions (Log-scale)",
       x = "Grass and Forb % (scaled)",
       y = "Predicted NSQ count (log-scale)") +
  theme_classic()
```

Zero-inflated model:

```{r}
# Get predicted zero probabilities and standard errors
pred_nsq_veg_zi <- predict(nsq_glm_veg_noRE,
                            newdata = nsq_data_veg,
                            type = "zprob",
                            se.fit = TRUE) 

# Combine predictions with data frame
pred_nsq_veg_df <- cbind(nsq_data_veg,
                 zi_fit = pred_nsq_veg_zi$fit,
                 zi_se = pred_nsq_veg_zi$se.fit)

ggplot(pred_nsq_veg_df, aes(x = grass_forb_perc_scaled, y = zi_fit, color = plot_type)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = zi_fit - 1.96*zi_se, ymax = zi_fit + 1.96*zi_se), alpha = 0.2) +
  labs(title = "Zero-Inflation Model: Probability of Structural Zeros",
       x = "Grass and Forb % (scaled)",
       y = "Probability of Excess Zeros") +
  theme_classic()
```

Floral abundance

Conditional model:

```{r}
# Define a sequence of values for grass_forb_perc_scaled across its observed range
floral_seq <- seq(
  from = min(surveys$bombus_floral_abun_scaled, na.rm = TRUE),
  to = max(surveys$bombus_floral_abun_scaled, na.rm = TRUE),
  length.out = 50
)

#create new data gri
nsq_data_floral<- expand.grid(
  grass_forb_perc_scaled = mean(surveys$bombus_floral_abun_scaled, na.rm = TRUE),
  plot_type = unique(surveys$plot_type),
  bombus_floral_abun_scaled = floral_seq,
  bee_season_type = "nestsearching"
)

# Get predicted conditional probabilities and standard errors
pred_nsq_floral_cond <- predict(nsq_glm_veg_noRE,
                            newdata = nsq_data_floral,
                            type = "conditional",
                            se.fit = TRUE)

# Combine predictions with data frame
pred_nsq_floral_df <- cbind(nsq_data_floral,
                 cond_fit = pred_nsq_floral_cond$fit,
                 cond_se = pred_nsq_floral_cond$se.fit)

ggplot(pred_nsq_floral_df, aes(x = bombus_floral_abun_scaled, y = cond_fit, color = plot_type)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = cond_fit - 1.96*cond_se, ymax = cond_fit + 1.96*cond_se), alpha = 0.2) +
  labs(title = "Conditional Model Predictions (Log-scale)",
       x = "Floral abundance (scaled)",
       y = "Predicted NSQ count (log-scale)") +
  theme_classic()
```

Zero-inflated model:

```{r}
# Get predicted zero probabilities and standard errors
pred_nsq_floral_zi <- predict(nsq_glm_veg_noRE,
                            newdata = nsq_data_floral,
                            type = "zprob",
                            se.fit = TRUE) 

# Combine predictions with data frame
pred_nsq_floral_df <- cbind(nsq_data_floral,
                 zi_fit = pred_nsq_floral_zi$fit,
                 zi_se = pred_nsq_floral_zi$se.fit)

ggplot(pred_nsq_floral_df, aes(x = bombus_floral_abun_scaled, y = zi_fit, color = plot_type)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = zi_fit - 1.96*zi_se, ymax = zi_fit + 1.96*zi_se), alpha = 0.2) +
  labs(title = "Zero-Inflation Model: Probability of Structural Zeros",
       x = "Floral abundance (scaled)",
       y = "Probability of Excess Zeros") +
  theme_classic()
```

Season

Conditional model:

```{r}
# Create newdata with discrete factor levels for bee_season_type
nsq_data_season <- data.frame(
  grass_forb_perc_scaled = mean(surveys$grass_forb_perc_scaled),
  bombus_floral_abun_scaled = mean(surveys$bombus_floral_abun_scaled),
  bee_season_type = factor(c("nestsearching", "worker"), levels = levels(surveys$bee_season_type)),
  plot_type = "ditch"
)

# Get predicted zero probabilities and standard errors
pred_nsq_season_cond <- predict(nsq_glm_veg_noRE,
                            newdata = nsq_data_season,
                            type = "conditional",
                            se.fit = TRUE) 

# Combine predictions with data frame
pred_nsq_season_df <- cbind(nsq_data_season,
                 cond_fit = pred_nsq_season_cond$fit,
                 cond_se = pred_nsq_season_cond$se.fit)

ggplot(pred_nsq_season_df, aes(x = bee_season_type, y = cond_fit)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = cond_fit - 1.96*cond_se, ymax = cond_fit + 1.96*cond_se), width = 0.2, position = position_dodge(0.6)) +
  labs(title = "Conditional Model Predictions (Log-scale)",
       x = "Bee season",
       y = "Predicted NSQ count (log-scale)") +
  theme_classic()
```

Zero-inflated model:

```{r}
# Get predicted zero probabilities and standard errors
pred_nsq_season_zi <- predict(nsq_glm_veg_noRE,
                            newdata = nsq_data_season,
                            type = "zprob",
                            se.fit = TRUE) 

# Combine predictions with data frame
pred_nsq_season_df <- cbind(nsq_data_season,
                 zi_fit = pred_nsq_season_zi$fit,
                 zi_se = pred_nsq_season_zi$se.fit)

ggplot(pred_nsq_season_df, aes(x = bee_season_type, y = zi_fit)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = zi_fit - 1.96*zi_se, ymax = zi_fit + 1.96*zi_se), width = 0.2, position = position_dodge(0.6)) +
  labs(title = "Zero-Inflation Model: Probability of Structural Zeros",
       x = "Bee season (scaled)",
       y = "Probability of Excess Zeros") +
  theme_classic()
```

### Findings

Conditional model (predicts abundance of nest-searching queens at sites where queens are present). Ditch is reference plot type.

-   Significant positive effect of grass+forb% on number of NSQ in ditch plots

-   Marginal negative effect of plot type on \# NSQ (fewer queens in field)

Zero-inflation model (predicts presence/absence of nest-searching queens).

-   Significant positive effect of floral abundance in **field** plots on NSQ presence (fewer zeros when there are more flowers in the field)

-   Marginal negative effect of floral abundance in ditch plots on NSQ presence (more zeros when there are more flowers along the ditch)

# Workers

## vs. plot type

### Steps

1)  Start with poisson and check all random effect structures.

-   model with plot nested in site (no area) is best

```{r}
workers_plotType_full <- glmmTMB(workers ~ plot_type +
                              bee_season_type +
                              (1|area/site_id/plot_id), 
                            family = poisson, #poisson
                            data = surveys)
workers_plotType_noArea <- glmmTMB(workers ~ plot_type +
                              bee_season_type +
                              (1|site_id/plot_id), 
                            family = poisson, #poisson
                            data = surveys)

workers_plotType_siteOnly <- glmmTMB(workers ~ plot_type +
                              bee_season_type +
                              (1|site_id), 
                            family = poisson, #poisson
                            data = surveys)

workers_plotType_areaSite <- glmmTMB(workers ~ plot_type +
                              bee_season_type +
                              (1|area/site_id), 
                            family = poisson, #poisson
                            data = surveys)

#compare models
anova(workers_plotType_full, workers_plotType_noArea, workers_plotType_siteOnly, workers_plotType_areaSite)
##no area is best
```

2)  Check diagnostics of model.

-   no problems detected

```{r}
#simulate residuals
residuals_workers_plotType_noArea <- simulateResiduals(fittedModel = workers_plotType_noArea, plot = TRUE)
```

3)  Plot results

```{r}
summary(workers_plotType_noArea)
```

### Plots

Plot type

```{r}
# Create newdata with discrete factor levels for plot_type
worker_data_plot <- expand.grid(
  bee_season_type = "worker",
  plot_type = factor(c("field", "ditch"), levels = levels(surveys$plot_type))
)


# Predictions for conditional model (count)
pred_worker_plot <- predict(workers_plotType_noArea, 
                              newdata = worker_data_plot, 
                              se.fit = TRUE,
                              re.form = NA)

# Combine predictions with newdata
pred_worker_plot_df <- cbind(worker_data_plot,
                 fit = pred_worker_plot$fit,
                 se = pred_worker_plot$se.fit)

# Plot conditional predictions
ggplot(pred_worker_plot_df, aes(x = plot_type, y = fit)) +
  geom_point(size=3, position = position_dodge(0.6)) +
  geom_errorbar(aes(ymin = fit - 1.96*se, ymax = fit + 1.96*se), width = 0.2, position = position_dodge(0.6))+
  labs(title = "Model Predictions (Log-scale)", y = "Predicted worker count (log-scale)") +
  theme_classic()
```

Season

```{r}
# Create newdata with discrete factor levels for season
worker_data_season <- expand.grid(
  bee_season_type = factor(c("worker", "nestsearching"), levels = levels(surveys$bee_season_type)),
  plot_type = "field")


# Predictions for conditional model (count)
pred_worker_season <- predict(workers_plotType_noArea, 
                              newdata = worker_data_season, 
                              se.fit = TRUE,
                              re.form = NA)

# Combine predictions with newdata
pred_worker_season_df <- cbind(worker_data_season,
                 fit = pred_worker_season$fit,
                 se = pred_worker_season$se.fit)

# Plot conditional predictions
ggplot(pred_worker_season_df, aes(x = bee_season_type, y = fit)) +
  geom_point(size=3, position = position_dodge(0.6)) +
  geom_errorbar(aes(ymin = fit - 1.96*se, ymax = fit + 1.96*se), width = 0.2, position = position_dodge(0.6))+
  labs(title = "Model Predictions (Log-scale)", y = "Predicted worker count (log-scale)") +
  theme_classic()
```

### Findings

-   significantly more workers in ditch than in field

-   significantly more workers during worker season than during NSQ season

## vs. vegetation predictors

### Steps

1)  Try random effect structures with poisson.

-   model with plot nested in site (no area) is best

```{r}
workers_veg_full <- glmmTMB(workers ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type+
                              (1|area/site_id/plot_id),
                            family = poisson, 
                            data = surveys)

workers_veg_noArea <- glmmTMB(workers ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                bee_season_type +
                              (1|site_id/plot_id),
                            family = poisson, 
                            data = surveys)

workers_veg_siteOnly <- glmmTMB(workers ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                bee_season_type +
                              (1|site_id),
                            family = poisson, 
                            data = surveys)
workers_veg_noPlot <- glmmTMB(workers ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                bee_season_type +
                              (1|area/site_id),
                            family = poisson, 
                            data = surveys)

#compare models
anova(workers_veg_full, workers_veg_noArea, workers_veg_siteOnly, workers_veg_noPlot)
##no area is best
```

2)  Check diagnostics

-   quantile test shows significant deviation

```{r}
residuals_workers_veg_noArea <- simulateResiduals(fittedModel = workers_veg_noArea, plot = TRUE)
```

3)  Try negative binomial and check diagnostics.

-   diagnostic plots look good!

```{r}
workers_veg_noArea_nbin <- glmmTMB(workers ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                bee_season_type +
                              (1|site_id/plot_id),
                            family = nbinom2, 
                            data = surveys)

#check diagnostics
#simulate residuals
residuals_workers_veg_noArea_nbin <- simulateResiduals(fittedModel = workers_veg_noArea_nbin, plot = TRUE)
```

4.  Plot results.

```{r}
summary(workers_veg_noArea_nbin)
```

### Plots

Vegetation

```{r}
#create new data grid
worker_data_veg<- expand.grid(
  grass_forb_perc_scaled = grass_seq,
  plot_type = unique(surveys$plot_type),
  bombus_floral_abun_scaled = mean(surveys$bombus_floral_abun_scaled, na.rm = TRUE),
  bee_season_type = "worker"
)

# Get predicted probabilities and standard errors
pred_worker_veg <- predict(workers_veg_noArea_nbin,
                            newdata = worker_data_veg,
                            se.fit = TRUE,
                            re.form = NA)

# Combine predictions with data frame
pred_worker_veg_df <- cbind(worker_data_veg,
                 fit = pred_worker_veg$fit,
                 se = pred_worker_veg$se.fit)

ggplot(pred_worker_veg_df, aes(x = grass_forb_perc_scaled, y = fit, color = plot_type)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = fit - 1.96*se, ymax = fit + 1.96*se), alpha = 0.2) +
  labs(title = "Model Predictions (Log-scale)",
       x = "Grass and Forb % (scaled)",
       y = "Predicted worker count (log-scale)") +
  theme_classic()
```

Floral abundance

```{r}
#create new data grid
worker_data_floral<- expand.grid(
  grass_forb_perc_scaled = mean(surveys$grass_forb_perc_scaled, na.rm = TRUE),
  plot_type = unique(surveys$plot_type),
  bombus_floral_abun_scaled = floral_seq,
  bee_season_type = "worker"
)

# Get predicted probabilities and standard errors
pred_worker_floral <- predict(workers_veg_noArea_nbin,
                            newdata = worker_data_floral,
                            se.fit = TRUE,
                            re.form = NA)

# Combine predictions with data frame
pred_worker_floral_df <- cbind(worker_data_floral,
                 fit = pred_worker_floral$fit,
                 se = pred_worker_floral$se.fit)

ggplot(pred_worker_floral_df, aes(x = bombus_floral_abun_scaled, y = fit, color = plot_type)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = fit - 1.96*se, ymax = fit + 1.96*se), alpha = 0.2) +
  labs(title = "Model Predictions (Log-scale)",
       x = "Floral abundance (scaled)",
       y = "Predicted worker count (log-scale)") +
  theme_classic()
```

Season

```{r}
#create new data grid
worker_data_season<- expand.grid(
  grass_forb_perc_scaled = mean(surveys$grass_forb_perc_scaled, na.rm = TRUE),
  plot_type = "ditch",
  bombus_floral_abun_scaled = mean(surveys$bombus_floral_abun_scaled, na.rm = TRUE),
  bee_season_type = unique(surveys$bee_season_type)
)

# Get predicted probabilities and standard errors
pred_worker_season <- predict(workers_veg_noArea_nbin,
                            newdata = worker_data_season,
                            se.fit = TRUE,
                            re.form = NA)

# Combine predictions with data frame
pred_worker_season_df <- cbind(worker_data_season,
                 fit = pred_worker_season$fit,
                 se = pred_worker_season$se.fit)

ggplot(pred_worker_season_df, aes(x = bee_season_type, y = fit)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = fit - 1.96*se, ymax = fit + 1.96*se), width = 0.2, position = position_dodge(0.6)) +
  labs(title = "Model Predictions (Log-scale)",
       x = "Bee season",
       y = "Predicted worker count (log-scale)") +
  theme_classic()
```

### Findings

No significant effect of any variables on worker abundance/presence.

# Floral

## vs. plot type

### Steps

1.  Try random effect structures

```{r}
#Load packages needed for CLM
library(ordinal)
library(MASS)

surveys$bombus_floral_abun <- factor(surveys$bombus_floral_abun, ordered = TRUE)

floral_full <- clmm(as.factor(bombus_floral_abun) ~ plot_type +
                         julian.date_scaled + 
                         (1|area/site_id/plot_id),
                       link = "logit", 
                       data = surveys)

floral_noArea <- clmm(as.factor(bombus_floral_abun) ~ plot_type +
                         julian.date_scaled + 
                         (1|site_id/plot_id),
                       link = "logit", 
                       data = surveys)

#floral_plotOnly <- clmm(as.factor(bombus_floral_abun) ~ plot_type +
 #                        julian.date_scaled + 
  #                       (1|plot_id),
   #                    link = "logit", 
    #                   data = surveys)
##can't use since there are only 2 levels

floral_siteOnly <- clmm(as.factor(bombus_floral_abun) ~ plot_type +
                         julian.date_scaled + 
                         (1|site_id),
                       link = "logit", 
                       data = surveys)

floral_AreaPlot <- clmm(as.factor(bombus_floral_abun) ~ plot_type +
                         julian.date_scaled + 
                         (1|area/plot_id),
                       link = "logit", 
                       data = surveys)
floral_noRE <- polr(as.factor(bombus_floral_abun) ~ plot_type +
                         julian.date_scaled,
                       method = "logistic", 
                       data = surveys)

#compare models AIC
AIC(floral_full, floral_noArea, floral_siteOnly, floral_AreaPlot, floral_noRE)
##shows that model without random effect has best fit
```

2.  Compare AIC between model with and without cover_type as a covariate.

-   determined that model without cover type as covariate was better

```{r}
floral_noRE_cover <- polr(as.factor(bombus_floral_abun) ~ plot_type +
                            cover_type +
                         julian.date_scaled,
                       method = "logistic", 
                       data = surveys)

AIC(floral_noRE_cover, floral_noRE)
##not including cover is better
```

3.  Try including julianDate \* plot_type

-   determined that model with interaction term was better

```{r}
floral_noRE_interaction <- polr(as.factor(bombus_floral_abun) ~ plot_type*julian.date_scaled,
                       method = "logistic", 
                       data = surveys)
AIC(floral_noRE_interaction, floral_noRE)
##interaction is better
```

4.  Look at output

```{r}
summary(floral_noRE_interaction)
```

### Plots

```{r}
# Sequence of Julian dates
date_seq <- seq(
  from = min(surveys$julian.date_scaled, na.rm = TRUE),
  to = max(surveys$julian.date_scaled, na.rm = TRUE),
  length.out = 50
)

# Get predicted probabilities per category with CIs
pred_floral_veg <- emmeans(
  floral_noRE_interaction,
  ~ julian.date_scaled * plot_type,
  at = list(julian.date_scaled = date_seq),
  type = "response",  # To get probabilities
)

# Convert to dataframe
pred_floral_veg_df <- as.data.frame(pred_floral_veg)

#plot
ggplot(pred_floral_veg_df, aes(x = julian.date_scaled, y = response, color = plot_type)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL, fill = plot_type), alpha = 0.2, color = NA) +
  labs(
    title = "Predicted Bombus Floral Abundance (scaled)",
    x = "Julian Date (scaled)",
    y = "Floral abundance (scaled)",
    color = "Plot Type",
    fill = "Plot Type"
  ) +
  theme_classic()



```

Extra code – using predict() (doesn't have CI)

```{r}
# Create prediction grid
floraldata <- expand.grid(
  julian.date_scaled = seq(min(surveys$julian.date_scaled), max(surveys$julian.date_scaled), length.out = 50),
  plot_type = unique(surveys$plot_type)
)

# Calculate predicted values 
pred_floral_plot <- predict(floral_noRE_interaction, floraldata, type = "probs")

# Convert matrix to dataframe
pred_floral_plot_df <- as.data.frame(pred_floral_plot)

# Bind predictions to floral data
floraldata_pred <- cbind(floraldata, pred_floral_plot_df)

# Reshape to long format
floraldata_pred_long <- floraldata_pred %>%
  pivot_longer(cols = -c(julian.date_scaled, plot_type),
               names_to = "category",
               values_to = "probability")

# Plot
ggplot(floraldata_pred_long, aes(x = julian.date_scaled, y = probability, color = plot_type)) +
  geom_line(size = 1) +
  facet_wrap(~ category) +
  labs(
    title = "Predicted Probabilities of Bombus Floral Abundance Categories",
    x = "Julian Date (scaled)",
    y = "Probability",
    color = "Bombus Floral Abundance"
  ) +
  theme_classic()
```

Only floral abundance category 4 (1000-10000 blooms)

```{r}
# Filter to only category "4"
floraldata_pred_cat4 <- floraldata_pred %>%
  dplyr::select(julian.date_scaled, plot_type, `4` = `4`) %>%
  dplyr::rename(probability = `4`)

# Plot
ggplot(floraldata_pred_cat4, aes(x = julian.date_scaled, y = probability, color = plot_type)) +
  geom_line(size = 1.2) +
  labs(
    title = "Predicted Probability of Bombus Floral Abundance Category 4(1000-10000 blooms)",
    x = "Julian Date (scaled)",
    y = "Probability",
    color = "Plot Type"
  ) +
  theme_classic()
```

### Findings

-   Julian date is a significant predictor: floral abundance tends to increase as season progresses.
-   The effect of Julian date changes depending on plot type (significant interaction).
    -   In the ditch, floral abundance increases throughout the season.

    -   In the field, floral abundance decreases throughout the season.
-   Plot type alone has no strong main effect but modifies the Julian date effect.
