---
title: "glmCH2_extra"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

Create dataframe where each row represents one survey event

Load dataframes

```{r}
#load packages
library(tidyverse)
library(lubridate) #package used to generate julian date from calendar date


#read in data
bee <- read.csv("06_cleandata/bee_clean.csv")
floral <- read.csv("06_cleandata/floral_clean.csv")
rodent <- read.csv("06_cleandata/rodent_clean.csv")
veg <- read.csv("06_cleandata/veg_clean.csv")
site_info <- read.csv("06_cleandata/site_info_clean.csv")
surveys <- read.csv("06_cleandata/surveys.csv")
```

Fix dataframe structure

```{r}
#bee
str(bee)
bee$date <- ymd(bee$date)

#floral
str(floral)
floral$date <- ymd(floral$date)

#rodent
str(rodent)
rodent$date_day <- ymd(rodent$date_day)

#veg
str(veg)
veg$date <- ymd(veg$date) 

#site_info
str(site_info)

#surveys
str(surveys)
surveys <- surveys %>%
  rename(site_id = site)
surveys <- surveys %>%
  rename(plot_id = plots)
```

# Add date of bee survey to the associated row in the survey data, based on site, plot and round.

```{r}
# Extract relevant bee dates first 
bee_dates <- bee %>% 
  select(site_id, plot_id, round, date) %>% 
  rename(bee_date = date)

#Make bee_dates file only contain one row for each site/plot/round/date combination 
bee_dates_unique <- bee_dates %>% 
  group_by(site_id, round) %>% 
  summarise(bee_date = first(bee_date), .groups = "drop") #choose first date since all dates within a site/plot/round combo are the same

#Join bee_date onto survey data

surveys <- surveys %>% left_join(bee_dates_unique, by = c("site_id", "round"))

#Need to manually add in the dates for 2 surveys where I didn't find any bees (surjit_168 during 2025_r2, bks_neaves during 2025_r1, bks_56 during 2024_r3, bks_56 during 2025_r3, bks_neaves during 2025_r3, surjit_169 during 2025_r3, sandhu_ladnertrunk during 2025_r3)
missing_dates <- tibble::tribble( 
  ~site_id, ~round, ~bee_date, 
  "surjit_168", "2025_r2", as.Date("2025-05-13"),
  "bks_neaves", "2025_r1", as.Date("2025-04-14"), 
  "bks_56", "2024_r3", as.Date("2025-06-21"), 
  "bks_56", "2025_r3", as.Date("2025-06-18"),
  "bks_neaves", "2025_r3", as.Date("2025-06-24"),
  "surjit_168", "2025_r3", as.Date("2025-06-25"), 
  "sandhu_ladnertrunk", "2025_r3", as.Date("2025-06-23") )

surveys <- surveys %>% rows_patch(missing_dates, by = c("site_id", "round"))

#add column for julian date 
surveys$julian.date <- yday(surveys$bee_date)
```

# Add area to survey dataframe based on site_id

```{r}
surveys <- surveys %>%
  left_join(site_info %>% select(site_id, area, cover_type), by = c("site_id"))

```

# Adding all the desired variables summarized at plot level as columns to the surveys dataframe

The random effect structure I first want to try in my models is to have plot nested within area.Therefore, I need to get all my desired variables to be at plot level, not transect or quadrat.

## The desired variables are:

Bee caste

Bee abundance

```{r}
# 1. Summarize your actual bee observations
bee_counts <- bee %>%
  filter(caste %in% c("q", "w")) %>%
  group_by(site_id, round, plot_id, caste) %>%
  summarise(n_bees = n(), .groups = "drop")

# 2. Use your existing 'surveys' dataframe as the base
# This ensures you ONLY have rows for actual survey events
surveys <- surveys %>%
  # Create a slot for both 'q' and 'w' for every real survey row
  crossing(caste = c("q", "w")) %>% 
  # Join the counts
  left_join(bee_counts, by = c("site_id", "round", "plot_id", "caste")) %>%
  # Fill the zeros for the surveys where that caste wasn't seen
  mutate(n_bees = replace_na(n_bees, 0))
```

Run model

```{r}
#Load packages
library(DHARMa) #diagnostic tools for mixed regression models
library(glmmTMB) #fits generalized linear mixed models, including zero-inflated and overdispersed models

#clean surveys dataframe and scale continuous predictors

#change character columns to factors
surveys <- surveys %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     area = as.factor(area),
     cover_type = as.factor(cover_type))

#add plot type column and make it a factor
surveys <- surveys %>% 
  mutate(plot_type = if_else(plot_id == "d", "ditch", "field"))
surveys <- surveys %>%
   mutate(
     plot_type = as.factor(plot_type))

#scale continuous predictors
surveys$julian.date_scaled <- as.numeric(scale(surveys$julian.date))
surveys$n_bees_scaled <- as.numeric(scale(surveys$n_bees))
```

Try different random effects structures. Use zero-inflated model to account for high abundance of zeros in the dataset.

```{r}
{r}
caste_glm_full <- glmmTMB(n_bees ~ caste*julian.date_scaled*plot_type +
                            cover_type +
                              (1|area/site_id/plot_id),
                          ziformula = ~ caste*julian.date_scaled*plot_type +
                            cover_type +
                              (1|area/site_id/plot_id),  # zero-inflation model
                            family = poisson, 
                            data = surveys)

caste_glm_noArea <- glmmTMB(n_bees ~ caste*julian.date_scaled*plot_type +
                            cover_type +
                              (1|site_id/plot_id),
                            ziformula = ~ caste*julian.date_scaled*plot_type +
                            cover_type +
                              (1|site_id/plot_id),
                            family = poisson, 
                            data = surveys)

caste_glm_justSite <- glmmTMB(n_bees ~ caste*julian.date_scaled*plot_type +
                            cover_type +
                              (1|site_id),
                            ziformula = ~ caste*julian.date_scaled*plot_type +
                            cover_type +
                              (1|site_id),
                            family = poisson,
                            data = surveys)

caste_glm_noRE <- glmmTMB(n_bees ~ caste*julian.date_scaled*plot_type +
                            cover_type,
                          ziformula = ~ caste*julian.date_scaled*plot_type +
                            cover_type,
                            family = poisson,
                            data = surveys)

#compare models
anova(caste_glm_noRE, caste_glm_justSite, caste_glm_noArea, caste_glm_full)
##no area model is best
```

2)  Check diagnostics of model

-   look good

```{r}
#simulate residuals
residuals_caste_glm_noArea <- simulateResiduals(fittedModel = caste_glm_noArea, plot = TRUE)
```

3)  Plot results

```{r}
summary(caste_glm_noArea)
```

Use emmeans to look at estimates

```{r}
library(emmeans)

#compare the seasonal slopes for each caste in each plot type
emtrends(caste_glm_noArea, ~ caste | plot_type, var = "julian.date_scaled")

#predicted counts for each group at the mean date
emmeans(caste_glm_noArea, ~ caste * plot_type)
```

Plots

```{r}
#set peak bloom start and end dates
bloom_start <- as.numeric(format(as.Date("2024-04-19"), "%j"))
bloom_end <- as.numeric(format(as.Date("2024-05-24"), "%j"))

ggplot(surveys, aes(x = julian.date, y = n_bees, color = plot_type)) +
  # Add the bloom window as a shaded background
  annotate("rect", xmin = bloom_start, xmax = bloom_end, ymin = -Inf, ymax = Inf, 
           alpha = .2, fill = "steelblue") +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  facet_wrap(~caste) +
  labs(title = "Bee Activity Relative to Blueberry Bloom (Shaded)",
       x = "Julian Date", y = "Average Bees per Survey") +
  theme_classic()
```
