---
title: "glmCH2_clean"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load and clean data

```{r}
#load packages
library(tidyverse) #data manipulation, visualization etc. (dplyr, ggplot2, tidyr)
library(DHARMa) #diagnostic tools for mixed regression models
library(glmmTMB) #fits generalized linear mixed models, including zero-inflated and overdispersed models
library(car) #tools for regression diagnostics and hypothesis tests
library(emmeans)   # for estimated marginal means (predictions)
library(lubridate) #date <-> julian date

#read in data
data <- read.csv("06_cleandata/combined_plot_data_ch2.csv")

#change character columns to factors
surveys <- data %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     bee_season_type = as.factor(bee_season_type),
     area = as.factor(area),
     cover_type = as.factor(cover_type))

#update cover type names
levels(surveys$cover_type) <- c("bare", "herbaceous")

#add plot type column and make it a factor
surveys <- surveys %>% 
  mutate(plot_type = if_else(plot_id == "d", "ditch", "field"))
surveys <- surveys %>%
   mutate(
     plot_type = as.factor(plot_type))

#scale continuous predictors
surveys$julian.date_scaled <- as.numeric(scale(surveys$julian.date))
surveys$bare_perc_scaled <- as.numeric(scale(surveys$bare_perc))
surveys$grass_perc_scaled <- as.numeric(scale(surveys$grass_perc))
surveys$forb_perc_scaled <- as.numeric(scale(surveys$forb_perc))
surveys$grass_forb_perc_scaled <- as.numeric(scale(surveys$grass_forb_perc))
surveys$grass_h_scaled <- as.numeric(scale(surveys$grass_h))
surveys$forb_h_scaled <- as.numeric(scale(surveys$forb_h))
surveys$floral_perc_scaled <- as.numeric(scale(surveys$floral_perc))
surveys$grass_forb_floral_perc_scaled <- as.numeric(scale(surveys$grass_forb_floral_perc))
surveys$bombus_floral_perc_scaled <- as.numeric(scale(surveys$bombus_floral_perc))
surveys$flood_perc_scaled <- as.numeric(scale(surveys$flood_perc))
surveys$dry_root_perc_scaled <- as.numeric(scale(surveys$dry_root_perc))
surveys$bryo_perc_scaled <- as.numeric(scale(surveys$bryo_perc))
surveys$bare_flood_dry_bryo_perc_scaled <- as.numeric(scale(surveys$bare_flood_dry_bryo_perc))
surveys$floral_richness_scaled <- as.numeric(scale(surveys$floral_richness))
surveys$bombus_floral_richness_scaled <- as.numeric(scale(surveys$bombus_floral_richness))
surveys$bombus_floral_abun_scaled <- as.numeric(scale(surveys$bombus_floral_abun))
surveys$nest_searching_queens_scaled <- as.numeric(scale(surveys$nest_searching_queens))
surveys$workers_scaled <- as.numeric(scale(surveys$workers))
surveys$forage_vaco_scaled <- as.numeric(scale(surveys$forage_vaco))
surveys$num_burrows_scaled <- as.numeric(scale(surveys$num_burrows))
```

# Choosing which vegetation variables to include as predictors

1.  Use PCA to determine which vegetation variables to include in my models. I don't want to include all of the predictors, since some will be correlated with each other and therefore dilute their effects of the response variables. I am using PCA so I can pick one variable that drives each of the main 2 components, and use these as the predictors in my model.

```{r}
pca_res <- prcomp(surveys[, c("bare_perc_scaled",
                              "grass_forb_perc_scaled",
                              "grass_h_scaled",
                              "forb_h_scaled",
                              "bombus_floral_perc_scaled",
                              "flood_perc_scaled",
                              "dry_root_perc_scaled",
                              "bombus_floral_richness_scaled",
                              "bombus_floral_abun_scaled")], scale. = TRUE)
summary(pca_res)

#see which variable drives each of the principle components
print(pca_res$rotation)
```

-   PCA Results:

    -   PC1: grass_h_scaled (-0.50),forb_h_scaled (-0.48),grass_forb_perc_scaled (-0.42),bombus_floral_richness_scaled (-0.43),bombus_floral_abun_scaled (-0.22)

    -   PC2: bare_perc_scaled (0.51),bombus_floral_abun_scaled (0.52),bombus_floral_perc_scaled (0.40), dry_root_perc_scaled (-0.40)

-   Based on these PCA results and my research question, I am selecting to include the following predictors in my model:

    -   grass_forb_perc_scaled (loads strongly on PCA1)

    -   bombus_floral_abun_scaled (loads strongly on PCA1 &2)

3.  Check pairwise correlations between this reduced set of variables

-   correlation among all the pairs are all near zero --\> predictors are all sufficiently independent! (R\<\<0.7)

```{r}
cor(surveys[, c("grass_forb_perc_scaled", "bombus_floral_abun_scaled")])
```

# Comparing field and ditch plots at grassy and bare farms.

## Vegetative cover

### Steps

Check raw distribution of grass/forb %

```{r}
hist(surveys$grass_forb_perc)
summary(surveys$grass_forb_perc)
```

Try different random effects structures with gamma model

```{r}
#adjust % grass and forb so that all values are positive
surveys$grass_forb_perc_adj <- surveys$grass_forb_perc + 0.001

#model with gamma distribution
ditchVSfield_veg_full <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled +
                              (1|area/site_id/plot_id),
                            family = Gamma(link = "log"), 
                            data = surveys)

ditchVSfield_veg_noArea <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled +
                              (1|site_id/plot_id),
                            family = Gamma(link = "log"), 
                            data = surveys)

ditchVSfield_veg_plotOnly <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled +
                              (1|plot_id),
                            family = Gamma(link = "log"), 
                            data = surveys)

ditchVSfield_veg_siteOnly <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled +
                              (1|site_id),
                            family = Gamma(link = "log"), 
                            data = surveys)

ditchVSfield_veg_siteArea <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled +
                              (1|area/site_id),
                            family = Gamma(link = "log"), 
                            data = surveys)

ditchVSfield_veg_areaPlot <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled +
                              (1|area/plot_id),
                            family = Gamma(link = "log"), 
                            data = surveys)

ditchVSfield_veg_area <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled +
                              (1|area),
                            family = Gamma(link = "log"), 
                            data = surveys)

ditchVSfield_veg_noRE <- glmmTMB(grass_forb_perc_adj ~ cover_type*plot_type + julian.date_scaled,
                            family = Gamma(link = "log"), 
                            data = surveys)

#compare models
anova(ditchVSfield_veg_full, ditchVSfield_veg_noArea, ditchVSfield_veg_plotOnly, ditchVSfield_veg_siteOnly, ditchVSfield_veg_siteArea,
      ditchVSfield_veg_areaPlot, ditchVSfield_veg_area, ditchVSfield_veg_noRE)
##area is best
```

Check diagnostics

```{r}
#simulate residuals
residuals_ditchVSfield_veg_area <- simulateResiduals(fittedModel = ditchVSfield_veg_area, plot = TRUE)

##results indicate model is not a good fit.
```

Try transforming data log(x+1) and running gaussian model

```{r}
##Log transform and add one to account for zeros.
surveys$log_grass_forb <- log(surveys$grass_forb_perc + 1)

##Model with gaussian distribution

ditchVSfield_veg_Area_gaus <- glmmTMB(log_grass_forb ~ cover_type*plot_type + julian.date_scaled +
                              (1|area),
                            family = gaussian(), 
                            data = surveys)
```

Check diagnostics

```{r}
#simulate residuals
residuals_ditchVSfield_veg_Area_gaus <- simulateResiduals(fittedModel = ditchVSfield_veg_Area_gaus, plot = TRUE)

##model fit is good enough.  
```

Try tweedie model

```{r}
ditchVSfield_veg_tweedie <- glmmTMB(
  grass_forb_perc ~ cover_type*plot_type +
    julian.date_scaled +
    (1|area),
  family = tweedie(link = "log"),
  data = surveys
)
```

Check diagnostics

```{r}
#simulate residuals
residuals_ditchVSfield_veg_tweedie <- simulateResiduals(fittedModel = ditchVSfield_veg_tweedie, plot = TRUE)

##looks good!
```

Print output

```{r}
summary(ditchVSfield_veg_tweedie)
```

Find effect of cover type within each plot type

```{r}
library(emmeans)

pairs(emmeans(ditchVSfield_veg_tweedie, ~ cover_type | plot_type))
```

Find effect of plot type within each cover type

```{r}
pairs(emmeans(ditchVSfield_veg_tweedie, ~ plot_type | cover_type))
```

### 

### Plots

Cover type x plot type

```{r}
#create new data grid
veg_data<- expand.grid(
  cover_type = unique(surveys$cover_type),
  plot_type = unique(surveys$plot_type),
  julian.date_scaled = mean(surveys$julian.date_scaled, na.rm = TRUE)
)

# Get predicted probabilities and standard errors
pred_veg <- predict(ditchVSfield_veg_tweedie,
                            newdata = veg_data,
                            se.fit = TRUE,
                            re.form = NA,
                           type = "link")

# Combine predictions with data frame
pred_veg_df <- cbind(veg_data,
                 fit_log = pred_veg$fit,
                 se_log = pred_veg$se.fit)

#calculate predictions and confidence intervals on response scale (not log scale)
pred_veg_df <- pred_veg_df %>%
  mutate(
    fit = exp(fit_log),
    lower = exp(fit_log - 1.96 * se_log),
    upper = exp(fit_log + 1.96 * se_log)
  )

# Plot conditional predictions
ggplot(pred_veg_df, aes(x = plot_type, y = fit, color = cover_type)) +
  geom_point(size=8, position = position_dodge(0.6)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, position = position_dodge(0.6), linewidth = 1.5)+
  # Use scale_color_manual to define specific colors
  scale_color_manual(values = c("herbaceous" = "darkgreen", "bare" = "tan4")) +
  labs(x = "Plot type",
       y = "Predicted mean\n% grass and forb\n(per plot)",
       color = "Cover type") +
  theme_classic(base_size = 20)
```

Julian date

```{r}
date_seq <- seq(
  from = min(surveys$julian.date_scaled, na.rm = TRUE),
  to = max(surveys$julian.date_scaled, na.rm = TRUE),
  length.out = 50
)

#create new data grid
veg_data_date<- expand.grid(
  cover_type = "herbaceous",
  plot_type = "ditch",
  julian.date_scaled = date_seq
)

# Get predicted probabilities and standard errors
pred_veg_date <- predict(ditchVSfield_veg_tweedie,
                            newdata = veg_data_date,
                            se.fit = TRUE,
                            re.form = NA,
                           type = "link")

# Combine predictions with data frame
pred_veg_date_df <- cbind(veg_data_date,
                 fit_log = pred_veg_date$fit,
                 se_log = pred_veg_date$se.fit)

#calculate predictions and confidence intervals on response scale (not log scale)
pred_veg_date_df <- pred_veg_date_df %>%
  mutate(
    fit = exp(fit_log),
    lower = exp(fit_log - 1.96 * se_log),
    upper = exp(fit_log + 1.96 * se_log)
  )

#convert x-axis to actual dates
## Get scaling parameters from original data
jd_mean <- mean(surveys$julian.date, na.rm = TRUE)
jd_sd <- sd(surveys$julian.date, na.rm = TRUE)

## Convert scaled values back to dates
pred_veg_date_df <- pred_veg_date_df %>%
  mutate(
    # Reverse the (x - mean) / sd formula
    julian_orig = (julian.date_scaled * jd_sd) + jd_mean,
    # Convert Day of Year to an actual Date object (using 2024 as a reference year)
    date_formatted = as.Date(round(julian_orig) - 1, origin = "2024-01-01")
  )

#Plot
ggplot(pred_veg_date_df, aes(x = date_formatted, y = fit)) +
  geom_line(linewidth = 1.5) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color=NA) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") + # Format x-axis to show Month names (e.g., May, Jun, Jul)
  labs(x = "Month",
       y = "Predicted mean\n% grass and forb\n(per plot)") +
  theme_classic(base_size = 20)
```

### Findings

Cover type

-   Significantly more veg at grassy sites than bare both along the ditch (p=0.037) and within the field (p\<0.0001)

-   Significant interaction between plot type and cover type, difference in veg between grassy and bare is much greater in the field than along the ditch (p=0.0004)

-   Significantly less veg within the field than along the ditch (p\<0.0001)

Date

-   \% veg decreases over time, though not significant (p=0.118)

## Floral

### Steps

Try different random effects structures

```{r}
#Load packages needed for CLM
library(ordinal)
library(MASS)

surveys$bombus_floral_abun <- factor(surveys$bombus_floral_abun, ordered = TRUE)

floral_full <- clmm(bombus_floral_abun ~ cover_type*plot_type +
                         julian.date_scaled*plot_type + 
                         (1|area/site_id/plot_id),
                       link = "logit", 
                       data = surveys)

floral_noArea <- clmm(bombus_floral_abun ~ cover_type*plot_type +
                         julian.date_scaled*plot_type + 
                         (1|site_id/plot_id),
                       link = "logit", 
                       data = surveys)

#floral_plotOnly <- clmm(bombus_floral_abun ~ plot_type +
 #                        julian.date_scaled + 
  #                       (1|plot_id),
   #                    link = "logit", 
    #                   data = surveys)
##can't use since there are only 2 levels

floral_siteOnly <- clmm(bombus_floral_abun ~ cover_type*plot_type +
                         julian.date_scaled*plot_type + 
                         (1|site_id),
                       link = "logit", 
                       data = surveys)

floral_AreaPlot <- clmm(bombus_floral_abun ~ cover_type*plot_type +
                         julian.date_scaled*plot_type + 
                         (1|area/plot_id),
                       link = "logit", 
                       data = surveys)

floral_Area <- clmm(bombus_floral_abun ~ cover_type*plot_type +
                         julian.date_scaled*plot_type + 
                         (1|area),
                       link = "logit", 
                       data = surveys)

floral_noRE <- polr(bombus_floral_abun ~ cover_type*plot_type +
                         julian.date_scaled*plot_type,
                       method = "logistic", 
                       data = surveys)

#compare models AIC
AIC(floral_full, floral_noArea, floral_siteOnly, floral_AreaPlot, floral_Area, floral_noRE)
##shows that model without random effect has best fit
```

Print results

```{r}
summary(floral_noRE)
```

Calculate p-values from t-values

```{r}
coefs <- coef(summary(floral_noRE))

# Two-sided p-values
p_values <- 2 * (1 - pnorm(abs(coefs[, "t value"])))

# Combine with coefficient table
coef_table <- cbind(coefs, "p value" = p_values)

coef_table
```

### Plots

Plot type

```{r}
# Get predicted probabilities per category with CIs
pred_floral_plot <- emmeans(
  floral_noRE,
  ~ plot_type,
  mode = "mean.class" ## The expected floral abundance category (mean ordinal score)
)

# Convert to dataframe
pred_floral_plot_df <- as.data.frame(pred_floral_plot)

#plot
ggplot(pred_floral_plot_df, aes(y = mean.class, x = plot_type)) +
  geom_point(size = 8, position = position_dodge(0.6)) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.2, linewidth = 1.5, position = position_dodge(0.6)) +
  labs(
    x = "Plot type",
    y = "Predicted mean floral\nabundance category\n(log-scaled)"
  ) +
  theme_classic(base_size = 20)
```

Cover type

```{r}
# Get predicted probabilities per category with CIs
pred_floral_cover <- emmeans(
  floral_noRE,
  ~ cover_type,
  mode = "mean.class" ## The expected floral abundance category (mean ordinal score)
)

# Convert to dataframe
pred_floral_cover_df <- as.data.frame(pred_floral_cover)

#plot
ggplot(pred_floral_cover_df, aes(x = cover_type, y = mean.class)) +
  geom_point(size = 8, position = position_dodge(0.6)) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.2, linewidth = 1.5, position = position_dodge(0.6)) +
  labs(
    x = "Cover type",
    y = "Predicted mean floral\nabundance category\n(log-scaled)"
  ) +
  theme_classic(base_size = 20)
```

Season x plot type

```{r}
# Sequence of Julian dates
date_seq <- seq(
  from = min(surveys$julian.date_scaled, na.rm = TRUE),
  to = max(surveys$julian.date_scaled, na.rm = TRUE),
  length.out = 200
)

# Get predicted probabilities per category with CIs
pred_floral_date <- emmeans(
  floral_noRE,
  ~ julian.date_scaled * plot_type,
  at = list(julian.date_scaled = date_seq),
  mode = "mean.class" ## The expected floral abundance category (mean ordinal score)
)

# Convert to dataframe
pred_floral_date_df <- as.data.frame(pred_floral_date)

#convert x-axis to actual dates
## Convert scaled values back to dates
pred_floral_date_df <- pred_floral_date_df %>%
  mutate(
    # Reverse the (x - mean) / sd formula
    julian_orig = (julian.date_scaled * jd_sd) + jd_mean,
    # Convert Day of Year to an actual Date object (using 2024 as a reference year)
    date_formatted = as.Date(round(julian_orig) - 1, origin = "2024-01-01")
  )

min(pred_floral_date_df$date_formatted, na.rm = TRUE)
max(pred_floral_date_df$date_formatted, na.rm = TRUE)

#set peak bloom start and end dates
bloom_start <- as.Date("2024-04-19")
bloom_end <- as.Date("2024-05-24")

#plot
ggplot(pred_floral_date_df, aes(x = date_formatted, y = mean.class, color = plot_type)) +
  geom_line(linewidth = 1.5) +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL, fill = plot_type), alpha = 0.2, color = NA) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") + # Format x-axis to show Month names (e.g., May, Jun, Jul) 
  annotate("rect", xmin = bloom_start, xmax = bloom_end, ymin = -Inf, ymax = Inf, 
           alpha = .2, fill = "steelblue")+
  labs(
    x = "Month",
    y = "Predicted mean floral\nabundance category\n(log-scaled)",
    color = "Plot type",
    fill = "Plot type"
  ) +
  theme_classic(base_size = 13)
```

### Findings

Cover type

-   No significant difference in floral abundance between ditch and field at beginning of the study (t=1.01)
-   No difference in floral abundance between grass and bare sites (t=0.531).
-   No significant interaction between plot type and cover type (t=-0.360)

Date

-   Significant effect of julian date (t=3.112)

-   Significant interaction between julian date and plot type (t=-3.444).

    -   Ditches get more flowers over time (+0.64 per unit of time).

    -   Fields get fewer flowers over time (-0.26 per unit of time)

## Rodents

### Steps

1)  Try random effect structures with poisson.

-   model with site only is best

```{r}

rodent_veg_areaOnly <- glmmTMB(num_burrows ~ cover_type*plot_type +
                              (1|area),
                            family = poisson, 
                            data = surveys)

rodent_veg_siteOnly <- glmmTMB(num_burrows ~ cover_type*plot_type +
                              (1|site_id),
                            family = poisson, 
                            data = surveys)
rodent_veg_siteArea <- glmmTMB(num_burrows ~ cover_type*plot_type +
                              (1|area/site_id),
                            family = poisson, 
                            data = surveys)

rodent_veg_noRE <- glmmTMB(num_burrows ~ cover_type*plot_type,
                            family = poisson, 
                            data = surveys)

#compare models
anova(rodent_veg_areaOnly, rodent_veg_siteOnly, rodent_veg_siteArea, rodent_veg_noRE)
##site only is best
```

2)  Check diagnostics

-   KS test shows significant deviation

```{r}
residuals_rodent_veg_siteOnly <- simulateResiduals(fittedModel = rodent_veg_siteOnly, plot = TRUE)
```

3)  Try negative binomial and check diagnostics.

-   diagnostic plots look good!

```{r}
rodent_veg_siteOnly_nbin <- glmmTMB(num_burrows ~ cover_type*plot_type +
                              (1|site_id),
                            family = nbinom2, 
                            data = surveys)

#check diagnostics
#simulate residuals
residuals_rodent_veg_siteOnly_nbin <- simulateResiduals(fittedModel = rodent_veg_siteOnly_nbin, plot = TRUE)
```

4.  Plot results.

```{r}
summary(rodent_veg_siteOnly_nbin)
```

Find effect of cover type in each plot type

```{r}
pairs(emmeans(rodent_veg_siteOnly_nbin, ~ cover_type | plot_type))
```

Find effect of plot type in each cover type

```{r}
pairs(emmeans(rodent_veg_siteOnly_nbin, ~ plot_type | cover_type))
```

### Plots

Plot type

```{r}
#create new data grid
rodent_data_plot<- expand.grid(
  cover_type = "herbaceous",
  plot_type = unique(surveys$plot_type),
  bombus_floral_abun_scaled = mean(surveys$bombus_floral_abun_scaled, na.rm = TRUE)
)

# Get predicted probabilities and standard errors
pred_rodent_plot <- predict(rodent_veg_siteOnly_nbin,
                            newdata = rodent_data_plot,
                            se.fit = TRUE,
                            re.form = NA,
                           type = "link")

# Combine predictions with data frame
pred_rodent_plot_df <- cbind(rodent_data_plot,
                 fit_log = pred_rodent_plot$fit,
                 se_log = pred_rodent_plot$se.fit)

#calculate predictions and confidence intervals on response scale (not log scale)
pred_rodent_plot_df <- pred_rodent_plot_df %>%
  mutate(
    fit = exp(fit_log),
    lower = exp(fit_log - 1.96 * se_log),
    upper = exp(fit_log + 1.96 * se_log)
  )

# Plot conditional predictions
ggplot(pred_rodent_plot_df, aes(x = plot_type, y = fit)) +
  geom_point(size=8, position = position_dodge(0.6)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, position = position_dodge(0.6), linewidth = 1.5)+
  labs(x = "Plot type",
       y = "Predicted mean burrow\ncount (per plot)") +
  theme_classic(base_size = 20)
```

cover type x plot type

```{r}
#create new data grid
rodent_data_veg<- expand.grid(
  cover_type = unique(surveys$cover_type),
  plot_type = unique(surveys$plot_type),
  bombus_floral_abun_scaled = mean(surveys$bombus_floral_abun_scaled, na.rm = TRUE)
)

# Get predicted probabilities and standard errors
pred_rodent_veg <- predict(rodent_veg_siteOnly_nbin,
                            newdata = rodent_data_veg,
                            se.fit = TRUE,
                            re.form = NA,
                           type = "link")

# Combine predictions with data frame
pred_rodent_veg_df <- cbind(rodent_data_veg,
                 fit_log = pred_rodent_veg$fit,
                 se_log = pred_rodent_veg$se.fit)

#calculate predictions and confidence intervals on response scale (not log scale)
pred_rodent_veg_df <- pred_rodent_veg_df %>%
  mutate(
    fit = exp(fit_log),
    lower = exp(fit_log - 1.96 * se_log),
    upper = exp(fit_log + 1.96 * se_log)
  )

# Plot conditional predictions
ggplot(pred_rodent_veg_df, aes(x = plot_type, y = fit, color = cover_type)) +
  geom_point(size=8, position = position_dodge(0.6)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, position = position_dodge(0.6), linewidth = 1.5)+
  labs(x = "Plot type",
       y = "Predicted mean burrow\ncount (per plot)",
       color = "Cover type") +
  scale_color_manual(values = c("herbaceous" = "darkgreen", "bare" = "tan4")) +
  theme_classic(base_size = 20)
```

### Findings

-   Significant interaction between cover type and plot type (p\<0.0001)

    -   Marginal effect of cover type on rodent burrow abundance within the field only (p=0.0735)

    -   No effect of cover type on burrow abundance along the ditch (probably because less difference in vegetation) (p=0.826)

-   Significantly more rodents along the ditch than within the field (p\<0.0001)

## NSQ

Plot model with interaction between vegetation variables and plot type to investigate whether the effects of vegetation cover and floral abundance on queen abundance/presence change depending on plot type (ditch vs field).

### Steps

1)  Check for zero-inflation in the data by comparing expected proportion of zeros in the dataset to the actual proportion of zeros. If observed \> expected, ZI model is justified.

-   use ZI Poisson model to account for high proportion of zeros in NSQ counts (~79% of plots had zero queens), which exceeded expectations under a standard Poisson distribution (~64%).

```{r}
# Observed proportion of zeros
obs_zeros <- mean(surveys$nest_searching_queens == 0)

# Expected proportion of zeros under Poisson
lambda <- mean(surveys$nest_searching_queens)
exp_zeros <- dpois(0, lambda)

obs_zeros ##0.7904762
exp_zeros ##0.6421989
```

2.  Try different random effect structures.

-   Model without random effect fit data the best.

```{r}
nsq_glm_veg_full <- glmmTMB(nest_searching_queens ~ cover_type*plot_type+
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type +
                              (1|area/site_id/plot_id),
                            ziformula = ~cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type +
                              (1|area/site_id/plot_id),  # zero-inflation model
                            family = poisson, 
                            data = surveys)

nsq_glm_veg_noArea <- glmmTMB(nest_searching_queens ~ cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type+
                              bee_season_type+
                              (1|site_id/plot_id),
                            ziformula = ~cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type +
                              (1|site_id/plot_id),  # zero-inflation model
                            family = poisson, 
                            data = surveys)

nsq_glm_veg_justSite <- glmmTMB(nest_searching_queens ~ cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type+
                              (1|site_id),
                            ziformula = ~cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type+
                              bee_season_type +
                              (1|site_id),  # zero-inflation model
                            family = poisson,
                            data = surveys)

nsq_glm_veg_areaPlot <- glmmTMB(nest_searching_queens ~ cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type+
                              (1|area/plot_id),
                            ziformula = ~cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type+
                              bee_season_type +
                              (1|area/plot_id),  # zero-inflation model
                            family = poisson,
                            data = surveys)

nsq_glm_veg_justarea <- glmmTMB(nest_searching_queens ~ cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type+
                              (1|area),
                            ziformula = ~cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type+
                              bee_season_type +
                              (1|area),  # zero-inflation model
                            family = poisson,
                            data = surveys)

nsq_glm_veg_justPlot <- glmmTMB(nest_searching_queens ~ cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type+
                              (1|plot_id),
                            ziformula = ~cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type+
                              bee_season_type +
                              (1|plot_id),  # zero-inflation model
                            family = poisson,
                            data = surveys)

nsq_glm_veg_noRE <- glmmTMB(nest_searching_queens ~ cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type,
                            ziformula = ~cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type,  # zero-inflation model
                            family = poisson,
                            data = surveys)

#compare models
anova(nsq_glm_veg_noRE, nsq_glm_veg_justSite, nsq_glm_veg_noArea, nsq_glm_veg_areaPlot,nsq_glm_veg_justarea, nsq_glm_veg_justPlot,  nsq_glm_veg_full)
##no random effect model is best, followed by just site model
```

3.  Check diagnostics of model

-   look good

```{r}
#simulate residuals
residuals_nsq_glm_veg_noRE <- simulateResiduals(fittedModel = nsq_glm_veg_noRE, plot = TRUE)
```

4.  Plot results

```{r}
summary(nsq_glm_veg_noRE)
```

### Plots

Plot type

Conditional

```{r}
# Create newdata 
nsq_data_plottype <- expand.grid(
  cover_type = "herbaceous",
  plot_type = factor(c("ditch", "field"), levels = levels(surveys$plot_type))
)
nsq_data_plottype$bombus_floral_abun_scaled <- mean(surveys$bombus_floral_abun_scaled)
nsq_data_plottype$bee_season_type <- "nestsearching"


# Get predicted zero probabilities and standard errors
pred_nsq_plottype_cond <- predict(nsq_glm_veg_noRE,
                            newdata = nsq_data_plottype,
                            type = "link",
                            se.fit = TRUE) 

# Combine predictions with data frame
pred_nsq_plottype_df <- cbind(nsq_data_plottype,
                 cond_fit_log = pred_nsq_plottype_cond$fit,
                 cond_se_log = pred_nsq_plottype_cond$se.fit)

#calculate predictions and confidence intervals on response scale (not log scale)
pred_nsq_plottype_df <- pred_nsq_plottype_df %>%
  mutate(
    cond_fit = exp(cond_fit_log),
    cond_lower = exp(cond_fit_log - 1.96 * cond_se_log),
    cond_upper = exp(cond_fit_log + 1.96 * cond_se_log)
  )

#Plot
ggplot(pred_nsq_plottype_df, aes(x = plot_type, y = cond_fit)) +
  geom_point(size = 8, position = position_dodge(0.6)) +
  geom_errorbar(aes(ymin = cond_lower, ymax = cond_upper), width = 0.2, position = position_dodge(0.6), linewidth = 1.5) +
  labs(x = "Plot type",
       y = "Predicted mean\nNSQ count\n(per plot)") +
  theme_classic(base_size = 30)
```

Zero-inflated model:

```{r}
# Predictions for ZI model 
pred_nsq_plottype_zi <- predict(nsq_glm_veg_noRE, 
                                newdata = nsq_data_plottype, 
                                type = "zlink",
                              se.fit = TRUE)

# Combine predictions and transform to response scale (by applying the inverse logit (plogit))
pred_nsq_plottype_zi_df <- cbind(
  nsq_data_plottype,
  zi_fit = plogis(pred_nsq_plottype_zi$fit),
  zi_lower = plogis(pred_nsq_plottype_zi$fit - 1.96 * pred_nsq_plottype_zi$se.fit),
  zi_upper = plogis(pred_nsq_plottype_zi$fit + 1.96 * pred_nsq_plottype_zi$se.fit)
)

# Plot zi predictions
ggplot(pred_nsq_plottype_zi_df, aes(x = plot_type, y = zi_fit)) +
  geom_point(size = 8,  position = position_dodge(0.6)) +
  geom_errorbar(aes(ymin = zi_lower, ymax = zi_upper), width = 0.2, position = position_dodge(0.6), linewidth = 1.5) +
  labs(x = "Plot type",
       y = "Predicted\nprobability of\nstructural zeros") +
  theme_classic(base_size = 30) 
```

Cover type

Conditional

```{r}
# Create newdata 
nsq_data_covertype <- expand.grid(
  cover_type = factor(c("herbaceous", "bare"), levels = levels(surveys$cover_type)),
  plot_type = "ditch"
)
nsq_data_covertype$bombus_floral_abun_scaled <- mean(surveys$bombus_floral_abun_scaled)
nsq_data_covertype$bee_season_type <- "nestsearching"


# Get predicted zero probabilities and standard errors
pred_nsq_covertype_cond <- predict(nsq_glm_veg_noRE,
                            newdata = nsq_data_covertype,
                            type = "link",
                            se.fit = TRUE) 

# Combine predictions with data frame
pred_nsq_covertype_df <- cbind(nsq_data_covertype,
                 cond_fit_log = pred_nsq_covertype_cond$fit,
                 cond_se_log = pred_nsq_covertype_cond$se.fit)

#calculate predictions and confidence intervals on response scale (not log scale)
pred_nsq_covertype_df <- pred_nsq_covertype_df %>%
  mutate(
    cond_fit = exp(cond_fit_log),
    cond_lower = exp(cond_fit_log - 1.96 * cond_se_log),
    cond_upper = exp(cond_fit_log + 1.96 * cond_se_log)
  )

#Plot
ggplot(pred_nsq_covertype_df, aes(x = cover_type, y = cond_fit)) +
  geom_point(size = 8, position = position_dodge(0.6)) +
  geom_errorbar(aes(ymin = cond_lower, ymax = cond_upper), width = 0.2, position = position_dodge(0.6), linewidth = 1.5) +
  labs(x = "Cover type",
       y = "Predicted mean\nNSQ count\n(per plot)") +
  theme_classic(base_size = 30)
```

Zero-inflated model:

```{r}
# Predictions for ZI model 
pred_nsq_covertype_zi <- predict(nsq_glm_veg_noRE, 
                                newdata = nsq_data_covertype, 
                                type = "zlink",
                              se.fit = TRUE)

# Combine predictions and transform to response scale (by applying the inverse logit (plogit))
pred_nsq_covertype_zi_df <- cbind(
  nsq_data_covertype,
  zi_fit = plogis(pred_nsq_covertype_zi$fit),
  zi_lower = plogis(pred_nsq_covertype_zi$fit - 1.96 * pred_nsq_covertype_zi$se.fit),
  zi_upper = plogis(pred_nsq_covertype_zi$fit + 1.96 * pred_nsq_covertype_zi$se.fit)
)

# Plot zi predictions
ggplot(pred_nsq_covertype_zi_df, aes(x = cover_type, y = zi_fit)) +
  geom_point(size = 8,  position = position_dodge(0.6)) +
  geom_errorbar(aes(ymin = zi_lower, ymax = zi_upper), width = 0.2, position = position_dodge(0.6), linewidth = 1.5) +
  labs(x = "Cover type",
       y = "Predicted\nprobability of\nstructural zeros") +
  theme_classic(base_size = 30) 
```

Floral abundance

Conditional model:

```{r}
# Define a sequence of values for grass_forb_perc_scaled across its observed range
floral_seq <- seq(
  from = min(surveys$bombus_floral_abun_scaled, na.rm = TRUE),
  to = max(surveys$bombus_floral_abun_scaled, na.rm = TRUE),
  length.out = 50
)

#create new data grid
nsq_data_floral<- expand.grid(
  cover_type = "herbaceous",
  plot_type = "ditch",
  bombus_floral_abun_scaled = floral_seq,
  bee_season_type = "nestsearching"
)

# Get predicted conditional probabilities and standard errors
pred_nsq_floral_cond <- predict(nsq_glm_veg_noRE,
                            newdata = nsq_data_floral,
                            type = "link",
                            se.fit = TRUE)

# Combine predictions with data frame
pred_nsq_floral_df <- cbind(nsq_data_floral,
                 cond_fit_log = pred_nsq_floral_cond$fit,
                 cond_se_log = pred_nsq_floral_cond$se.fit)

#calculate predictions and confidence intervals on response scale (not log scale)
pred_nsq_floral_df <- pred_nsq_floral_df %>%
  mutate(
    cond_fit = exp(cond_fit_log),
    cond_lower = exp(cond_fit_log - 1.96 * cond_se_log),
    cond_upper = exp(cond_fit_log + 1.96 * cond_se_log)
  )

#plot
ggplot(pred_nsq_floral_df,
       aes(x = bombus_floral_abun_scaled,
           y = cond_fit)) +
  geom_line(linewidth = 1.5) +
  geom_ribbon(aes(ymin = cond_lower, ymax = cond_upper),
              alpha = 0.2,
              color = NA) +
  labs(
    x = "Floral abundance (scaled)",
    y = "Predicted mean\nNSQ count\n(per plot)"
  ) +
  theme_classic(base_size = 30)
```

Zero-inflated model:

```{r}
# Get predicted zero probabilities and standard errors
pred_nsq_floral_zi <- predict(nsq_glm_veg_noRE,
                            newdata = nsq_data_floral,
                            type = "zlink",
                            se.fit = TRUE) 

# Combine predictions and transform to response scale (by applying the inverse logit (plogit))
pred_nsq_floral_df <- cbind(
  nsq_data_floral,
  zi_fit = plogis(pred_nsq_floral_zi$fit),
  zi_lower = plogis(pred_nsq_floral_zi$fit - 1.96 * pred_nsq_floral_zi$se.fit),
  zi_upper = plogis(pred_nsq_floral_zi$fit + 1.96 * pred_nsq_floral_zi$se.fit)
)

#plot
ggplot(pred_nsq_floral_df, aes(x = bombus_floral_abun_scaled, y = zi_fit)) +
  geom_line(linewidth = 1.5) +
  geom_ribbon(aes(ymin = zi_lower, ymax = zi_upper), alpha = 0.2, color=NA) +
  labs(
       x = "Floral abundance (scaled)",
       y = "Predicted\nprobability of\nstructural zeros") +
  theme_classic(base_size = 30)
```

Season

Conditional model:

```{r}
# Create newdata with discrete factor levels for bee_season_type
nsq_data_season <- data.frame(
  cover_type = "herbaceous",
  bombus_floral_abun_scaled = mean(surveys$bombus_floral_abun_scaled),
  bee_season_type = factor(c("nestsearching", "worker"), levels = levels(surveys$bee_season_type)),
  plot_type = "ditch"
)

# Get predicted zero probabilities and standard errors
pred_nsq_season_cond <- predict(nsq_glm_veg_noRE,
                            newdata = nsq_data_season,
                            type = "link",
                            se.fit = TRUE) 

# Combine predictions with data frame
pred_nsq_season_df <- cbind(nsq_data_season,
                 cond_fit_log = pred_nsq_season_cond$fit,
                 cond_se_log = pred_nsq_season_cond$se.fit)

#calculate predictions and confidence intervals on response scale (not log scale)
pred_nsq_season_df <- pred_nsq_season_df %>%
  mutate(
    cond_fit = exp(cond_fit_log),
    cond_lower = exp(cond_fit_log - 1.96 * cond_se_log),
    cond_upper = exp(cond_fit_log + 1.96 * cond_se_log)
  )

ggplot(pred_nsq_season_df, aes(x = bee_season_type, y = cond_fit)) +
  geom_point(size = 8) +
  geom_errorbar(aes(ymin = cond_lower, ymax = cond_upper), width = 0.2, position = position_dodge(0.6), linewidth = 1.5) +
  labs(x = "Bee season",
       y = "Predicted mean NSQ\ncount (per plot)") +
  theme_classic(base_size = 20)
```

Zero-inflated model:

```{r}
# Get predicted zero probabilities and standard errors
pred_nsq_season_zi <- predict(nsq_glm_veg_noRE,
                            newdata = nsq_data_season,
                            type = "zprob",
                            se.fit = TRUE) 

# Combine predictions with data frame
pred_nsq_season_df <- cbind(nsq_data_season,
                 zi_fit = pred_nsq_season_zi$fit,
                 zi_se = pred_nsq_season_zi$se.fit)

#plot
ggplot(pred_nsq_season_df, aes(x = bee_season_type, y = zi_fit)) +
  geom_point(size = 8) +
  geom_errorbar(aes(ymin = zi_fit - 1.96*zi_se, ymax = zi_fit + 1.96*zi_se), width = 0.2, position = position_dodge(0.6), linewidth = 1.5) +
  labs(x = "Bee season",
       y = "Probability of\nExcess Zeros") +
  theme_classic(base_size = 20)
```

### Findings

Conditional model (predicts abundance of nest-searching queens at sites where queens are present). Ditch is reference plot type.

-   No significant effect of cover type, plot type, floral abundance, or season on NSQ abundance

-   No significant interaction between cover type and plot type of floral abundance and plot type

Zero-inflation model (predicts presence/absence of nest-searching queens).

-   NSQ are more likely to be absent from bare sites than grassy sites

-   NSQ are more likely to be absent during worker season than NSQ season

-   No significant effect of plot type or floral abundance on NSQ abundance

-   No significant interaction between cover type and plot type of floral abundance and plot type

## Workers

### Steps

1)  Try random effect structures with poisson.

-   model with plot nested in site (no area) is best

```{r}
workers_veg_full <- glmmTMB(workers ~ cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                              bee_season_type+
                              (1|area/site_id/plot_id),
                            family = poisson, 
                            data = surveys)

workers_veg_noArea <- glmmTMB(workers ~ cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                bee_season_type +
                              (1|site_id/plot_id),
                            family = poisson, 
                            data = surveys)

workers_veg_siteOnly <- glmmTMB(workers ~ cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                bee_season_type +
                              (1|site_id),
                            family = poisson, 
                            data = surveys)
workers_veg_noPlot <- glmmTMB(workers ~ cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                bee_season_type +
                              (1|area/site_id),
                            family = poisson, 
                            data = surveys)

workers_veg_Plot <- glmmTMB(workers ~ cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                bee_season_type +
                              (1|plot_id),
                            family = poisson, 
                            data = surveys)

workers_veg_areaPlot<- glmmTMB(workers ~ cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                bee_season_type +
                              (1|area/plot_id),
                            family = poisson, 
                            data = surveys)

workers_veg_noRE <- glmmTMB(workers ~ cover_type*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                bee_season_type,
                            family = poisson, 
                            data = surveys)

#compare models
anova(workers_veg_full, workers_veg_noArea, workers_veg_siteOnly, workers_veg_noPlot, workers_veg_Plot, workers_veg_areaPlot, workers_veg_noRE)
##no area is best
```

2)  Check diagnostics

-   looks okay

```{r}
residuals_workers_veg_noArea <- simulateResiduals(fittedModel = workers_veg_noArea, plot = TRUE)
```

3.  Plot results.

```{r}
summary(workers_veg_noArea)
```

Effect of floral abundance within each plot type

```{r}
#calculate the slope of floral abundance for each plot type
eff_slopes <- emtrends(workers_veg_noArea, ~ plot_type, var = "bombus_floral_abun_scaled")

#test if those slopes are significantly different from zero
summary(eff_slopes, infer = c(TRUE, TRUE))
```

### Plots

Plot type

```{r}
# Create newdata 
worker_data_plottype <- expand.grid(
  cover_type = "herbaceous",
  plot_type = factor(c("ditch", "field"), levels = levels(surveys$plot_type))
)
worker_data_plottype$bombus_floral_abun_scaled <- mean(surveys$bombus_floral_abun_scaled)
worker_data_plottype$bee_season_type <- "worker"

# Get predicted probabilities and standard errors
pred_worker_plottype <- predict(workers_veg_noArea,
                            newdata = worker_data_plottype,
                            se.fit = TRUE,
                            re.form = NA,
                            type = "link") 

# Combine predictions with data frame
pred_worker_plottype_df <- cbind(worker_data_plottype,
                 fit_log = pred_worker_plottype$fit,
                 se_log = pred_worker_plottype$se.fit)

#calculate predictions and confidence intervals on response scale (not log scale)
pred_worker_plottype_df <- pred_worker_plottype_df %>%
  mutate(
    fit = exp(fit_log),
    lower = exp(fit_log - 1.96 * se_log),
    upper = exp(fit_log + 1.96 * se_log)
  )

#Plot
ggplot(pred_worker_plottype_df, aes(x = plot_type, y = fit)) +
  geom_point(size = 8, position = position_dodge(0.6)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, position = position_dodge(0.6), linewidth = 1.5) +
  labs(x = "Plot type",
       y = "Predicted mean\nworker count\n(per plot)")  +
  theme_classic(base_size = 30)
```

Cover type

```{r}
# Create newdata 
worker_data_covertype <- expand.grid(
  cover_type = factor(c("herbaceous", "bare"), levels = levels(surveys$cover_type)),
  plot_type = "ditch"
)
worker_data_covertype$bombus_floral_abun_scaled <- mean(surveys$bombus_floral_abun_scaled)
worker_data_covertype$bee_season_type <- "worker"

# Get predicted probabilities and standard errors
pred_worker_covertype <- predict(workers_veg_noArea,
                            newdata = worker_data_covertype,
                            se.fit = TRUE,
                            re.form = NA,
                            type = "link") 

# Combine predictions with data frame
pred_worker_covertype_df <- cbind(worker_data_covertype,
                 fit_log = pred_worker_covertype$fit,
                 se_log = pred_worker_covertype$se.fit)

#calculate predictions and confidence intervals on response scale (not log scale)
pred_worker_covertype_df <- pred_worker_covertype_df %>%
  mutate(
    fit = exp(fit_log),
    lower = exp(fit_log - 1.96 * se_log),
    upper = exp(fit_log + 1.96 * se_log)
  )

#Plot
ggplot(pred_worker_covertype_df, aes(x = cover_type, y = fit)) +
  geom_point(size = 8, position = position_dodge(0.6)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, position = position_dodge(0.6), linewidth = 1.5) +
  labs(x = "Cover type",
       y = "Predicted mean\nworker count\n(per plot)") +
  theme_classic(base_size = 30)
```

Floral abundance x plot type

```{r}
#create new data grid
worker_data_floral<- expand.grid(
  cover_type = "herbaceous",
  plot_type = unique(surveys$plot_type),
  bombus_floral_abun_scaled = floral_seq,
  bee_season_type = "worker"
)

# Get predicted probabilities and standard errors
pred_worker_floral <- predict(workers_veg_noArea,
                            newdata = worker_data_floral,
                            se.fit = TRUE,
                            re.form = NA,
                            type = "link")

# Combine predictions with data frame
pred_worker_floral_df <- cbind(worker_data_floral,
                 fit_log = pred_worker_floral$fit,
                 se_log = pred_worker_floral$se.fit)

#calculate predictions and confidence intervals on response scale (not log scale)
pred_worker_floral_df <- pred_worker_floral_df %>%
  mutate(
    fit = exp(fit_log),
    lower = exp(fit_log - 1.96 * se_log),
    upper = exp(fit_log + 1.96 * se_log)
  )

#Plot
ggplot(pred_worker_floral_df, aes(x = bombus_floral_abun_scaled, y = fit, color = plot_type)) +
  geom_line(linewidth = 1.5) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = plot_type), alpha = 0.2, color=NA) +
  labs(x = "Floral abundance (scaled)",
       y = "Predicted mean\nworker count\n(per plot)",
       fill = "Plot type",
       color = "Plot type") +
  theme_classic(base_size = 30) +
  theme(
    legend.title = element_text(size = 25),
    legend.text = element_text(size = 20),
    legend.key.size = unit(0.5, "cm")
  )
```

Season

```{r}
#create new data grid
worker_data_season<- expand.grid(
  cover_type = "herbaceous",
  plot_type = "ditch",
  bombus_floral_abun_scaled = mean(surveys$bombus_floral_abun_scaled, na.rm = TRUE),
  bee_season_type = unique(surveys$bee_season_type)
)

# Get predicted probabilities and standard errors
pred_worker_season <- predict(workers_veg_noArea,
                            newdata = worker_data_season,
                            se.fit = TRUE,
                            re.form = NA,
                            type = "link")

# Combine predictions with data frame
pred_worker_season_df <- cbind(worker_data_season,
                 fit_log = pred_worker_season$fit,
                 se_log = pred_worker_season$se.fit)

#calculate predictions and confidence intervals on response scale (not log scale)
pred_worker_season_df <- pred_worker_season_df %>%
  mutate(
    fit = exp(fit_log),
    lower = exp(fit_log - 1.96 * se_log),
    upper = exp(fit_log + 1.96 * se_log)
  )

#Plot
ggplot(pred_worker_season_df, aes(x = bee_season_type, y = fit)) +
  geom_point(size = 8) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, position = position_dodge(0.6), linewidth = 1.5) +
  labs(x = "Bee season",
       y = "Predicted mean worker\ncount (per plot)",
       fill = "Plot type",
       color = "Plot type") +
  theme_classic(base_size = 20)
```

### Findings

Cover type

-   No effect of cover type on worker abundance (p=0.80)

-   More workers along the ditch than within the field (p=0.0243)

-   No interaction between cover type and plot type

Floral abundance

-   No overall impact of floral abundance on workers (p=0.9984)

-   Significant interaction between plot type and floral abundance (p=0.0169)

-   Floral abundance influences worker abundance within the field but NOT along the ditch

Date

-   Significantly more workers in worker season that NS (p\<0.0001)
