---
title: "glmCH2"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load and clean data

```{r}
#load packages
library(tidyverse)
library(ggplot2)
library(DHARMa)
library(glmmTMB)
library(ggeffects)
library(car)

#read in data
data <- read.csv("06_cleandata/combined_plot_data.csv")

#change character columns to factors
surveys <- data %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     bee_season_type = as.factor(bee_season_type),
     area = as.factor(area))

str(surveys)
```

Add plot type column (ditch vs field)

```{r}
surveys <- surveys %>% 
  mutate(plot_type = if_else(plot_id == "d", "ditch", "field"))

surveys <- surveys %>%
   mutate(
     plot_type = as.factor(plot_type))

```

Scale continuous predictors

```{r}
surveys$julian.date_scaled <- as.numeric(scale(surveys$julian.date))
surveys$bare_perc_scaled <- as.numeric(scale(surveys$bare_perc))
surveys$grass_perc_scaled <- as.numeric(scale(surveys$grass_perc))
surveys$forb_perc_scaled <- as.numeric(scale(surveys$forb_perc))
surveys$grass_forb_perc_scaled <- as.numeric(scale(surveys$grass_forb_perc))
surveys$grass_h_scaled <- as.numeric(scale(surveys$grass_h))
surveys$forb_h_scaled <- as.numeric(scale(surveys$forb_h))
surveys$floral_perc_scaled <- as.numeric(scale(surveys$floral_perc))
surveys$grass_forb_floral_perc_scaled <- as.numeric(scale(surveys$grass_forb_floral_perc))
surveys$bombus_floral_perc_scaled <- as.numeric(scale(surveys$bombus_floral_perc))
surveys$flood_perc_scaled <- as.numeric(scale(surveys$flood_perc))
surveys$dry_root_perc_scaled <- as.numeric(scale(surveys$dry_root_perc))
surveys$bryo_perc_scaled <- as.numeric(scale(surveys$bryo_perc))
surveys$bare_flood_dry_bryo_perc_scaled <- as.numeric(scale(surveys$bare_flood_dry_bryo_perc))
surveys$floral_richness_scaled <- as.numeric(scale(surveys$floral_richness))
surveys$bombus_floral_richness_scaled <- as.numeric(scale(surveys$bombus_floral_richness))
surveys$bombus_floral_abun_scaled <- as.numeric(scale(surveys$bombus_floral_abun))
surveys$nest_searching_queens_scaled <- as.numeric(scale(surveys$nest_searching_queens))
surveys$workers_scaled <- as.numeric(scale(surveys$workers))
surveys$forage_vaco_scaled <- as.numeric(scale(surveys$forage_vaco))
surveys$num_burrows_scaled <- as.numeric(scale(surveys$num_burrows))
surveys$num_inactive_burrows_scaled <- as.numeric(scale(surveys$num_inactive_burrows))
surveys$num_active_dm_scaled <- as.numeric(scale(surveys$num_active_dm))
surveys$num_dm_size_scaled <- as.numeric(scale(surveys$num_dm_size))
surveys$num_inactive_dm_scaled <- as.numeric(scale(surveys$num_inactive_dm))
```

# Choosing which vegetation variables to include as predictors by checking correlation among the variables

```{r}
vif(lm(nest_searching_queens ~ bare_perc_scaled + 
         grass_forb_perc_scaled + 
        grass_h_scaled + 
         forb_h_scaled + 
         bombus_floral_perc_scaled +
         flood_perc_scaled +
         dry_root_perc_scaled +
         bombus_floral_richness_scaled +
         bombus_floral_abun_scaled + bee_season_type, 
        data = surveys))

##all variables have VIF <5 (no multicollinearity). so i can include all these predictors in my model.
```

Run PCA to determine which vegetation variables to include

```{r}
pca_res <- prcomp(surveys[, c("bare_perc_scaled",
                              "grass_forb_perc_scaled",
                              "grass_h_scaled",
                              "forb_h_scaled",
                              "bombus_floral_perc_scaled",
                              "flood_perc_scaled",
                              "dry_root_perc_scaled",
                              "bombus_floral_richness_scaled",
                              "bombus_floral_abun_scaled")], scale. = TRUE)
summary(pca_res)

#see which variable drives each of the principle components
print(pca_res$rotation)
```

PCA Results:

PC1: grass_h_scaled (-0.50),forb_h_scaled (-0.48),grass_forb_perc_scaled (-0.42),bombus_floral_richness_scaled (-0.43),bombus_floral_abun_scaled (-0.22)

PC2: bare_perc_scaled (0.51),bombus_floral_abun_scaled (0.52),bombus_floral_perc_scaled (0.40), dry_root_perc_scaled (-0.40)

PC3: dry_root_perc_scaled (0.63), flood_perc_scaled (-0.64)

Based on these PCA results, I am selecting to include the following predictors in my model:

-   grass_forb_perc_scaled (loads strongly on PCA1)

-   bombus_floral_abun_scaled (loads strongly on PCA1 &2)

-   flood_perc_scaled (loads strongly on PCA3)

Check pairwise correlations between this reduced set of variables

```{r}
cor(surveys[, c("grass_forb_perc_scaled", "flood_perc_scaled", "bombus_floral_abun_scaled")])
##correlation among all the pairs are all near zero --> predictors are all sufficiently independent!
```

# GLM #1: Number of nest-searching queens vs plot type (ditch vs field)

Plotting a zero-inflated poisson model

```{r}
nsq_glm_plotType <- glmmTMB(nest_searching_queens ~ plot_type +
                              bee_season_type +
                              (1|area/site_id/plot_id),
                            ziformula = ~plot_type +
                              bee_season_type +
                              (1|area/site_id/plot_id),  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

#simulate residuals

residuals_nsq_glm_plotType <- simulateResiduals(fittedModel = nsq_glm_plotType, plot = TRUE)

#check results
summary(nsq_glm_plotType)

```

# GLM #2: Number of nest-searching queens vs vegetation predictors * plot type (ditch vs field)

```{r}
nsq_glm_veg <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled +
                              (1|area/site_id/plot_id),
                            ziformula = ~grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled+
                              (1|area/site_id/plot_id),  # zero-inflation model
                            family = poisson, 
                            data = surveys)

#model doesn't converge. try removing area

nsq_glm_veg_noArea <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled +
                              (1|site_id/plot_id),
                            ziformula = ~grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled+
                              (1|site_id/plot_id),  # zero-inflation model
                            family = poisson, 
                            data = surveys)

#Fixes convergence problem. check diagnostic plot

#simulate residuals

residuals_nsq_glm_veg_noArea <- simulateResiduals(fittedModel = nsq_glm_veg_noArea, plot = TRUE)

#try with just plot_id

nsq_glm_veg_justPlot <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled +
                              (1|plot_id),
                            ziformula = ~grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled+
                              (1|plot_id),  # zero-inflation model
                            family = poisson,
                            data = surveys)

#simulate residuals

residuals_nsq_glm_veg_justPlot <- simulateResiduals(fittedModel = nsq_glm_veg_justPlot, plot = TRUE)
#fixes problem

#check summary
summary(nsq_glm_veg_justPlot)
```

