---
title: "glmCH2_clean"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load and clean data

```{r}
#load packages
library(tidyverse)
library(ggplot2)
library(DHARMa)
library(glmmTMB)
library(ggeffects)
library(car)
library(emmeans)   # for estimated marginal means (predictions)

#read in data
data <- read.csv("06_cleandata/combined_plot_data.csv")

#change character columns to factors
surveys <- data %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     bee_season_type = as.factor(bee_season_type),
     area = as.factor(area))

#add plot type column and make it a factor
surveys <- surveys %>% 
  mutate(plot_type = if_else(plot_id == "d", "ditch", "field"))
surveys <- surveys %>%
   mutate(
     plot_type = as.factor(plot_type))

#scale continuous predictors
surveys$julian.date_scaled <- as.numeric(scale(surveys$julian.date))
surveys$bare_perc_scaled <- as.numeric(scale(surveys$bare_perc))
surveys$grass_perc_scaled <- as.numeric(scale(surveys$grass_perc))
surveys$forb_perc_scaled <- as.numeric(scale(surveys$forb_perc))
surveys$grass_forb_perc_scaled <- as.numeric(scale(surveys$grass_forb_perc))
surveys$grass_h_scaled <- as.numeric(scale(surveys$grass_h))
surveys$forb_h_scaled <- as.numeric(scale(surveys$forb_h))
surveys$floral_perc_scaled <- as.numeric(scale(surveys$floral_perc))
surveys$grass_forb_floral_perc_scaled <- as.numeric(scale(surveys$grass_forb_floral_perc))
surveys$bombus_floral_perc_scaled <- as.numeric(scale(surveys$bombus_floral_perc))
surveys$flood_perc_scaled <- as.numeric(scale(surveys$flood_perc))
surveys$dry_root_perc_scaled <- as.numeric(scale(surveys$dry_root_perc))
surveys$bryo_perc_scaled <- as.numeric(scale(surveys$bryo_perc))
surveys$bare_flood_dry_bryo_perc_scaled <- as.numeric(scale(surveys$bare_flood_dry_bryo_perc))
surveys$floral_richness_scaled <- as.numeric(scale(surveys$floral_richness))
surveys$bombus_floral_richness_scaled <- as.numeric(scale(surveys$bombus_floral_richness))
surveys$bombus_floral_abun_scaled <- as.numeric(scale(surveys$bombus_floral_abun))
surveys$nest_searching_queens_scaled <- as.numeric(scale(surveys$nest_searching_queens))
surveys$workers_scaled <- as.numeric(scale(surveys$workers))
surveys$forage_vaco_scaled <- as.numeric(scale(surveys$forage_vaco))
surveys$num_burrows_scaled <- as.numeric(scale(surveys$num_burrows))
surveys$num_inactive_burrows_scaled <- as.numeric(scale(surveys$num_inactive_burrows))
surveys$num_active_dm_scaled <- as.numeric(scale(surveys$num_active_dm))
surveys$num_dm_size_scaled <- as.numeric(scale(surveys$num_dm_size))
surveys$num_inactive_dm_scaled <- as.numeric(scale(surveys$num_inactive_dm))
```

# Choosing which vegetation variables to include as predictors

1)  Check for multicollinearity among vegetation variables.

-   All variables have VIF \<5 (no multicollinearity), so i can include all these predictors in my model

```{r}
vif(lm(nest_searching_queens ~ bare_perc_scaled + 
         grass_forb_perc_scaled + 
        grass_h_scaled + 
         forb_h_scaled + 
         bombus_floral_perc_scaled +
         flood_perc_scaled +
         dry_root_perc_scaled +
         bombus_floral_richness_scaled +
         bombus_floral_abun_scaled + bee_season_type, 
        data = surveys))
```

2)  Use PCA to determine which vegetation variables to include in my models. I don't want to include all of the predictors, since some will be correlated with each other and therefore dilute their effects of the response variables. I am using PCA so I can pick one variable that drives each of the main 3 components, and use these as the predictors in my model.

```{r}
pca_res <- prcomp(surveys[, c("bare_perc_scaled",
                              "grass_forb_perc_scaled",
                              "grass_h_scaled",
                              "forb_h_scaled",
                              "bombus_floral_perc_scaled",
                              "flood_perc_scaled",
                              "dry_root_perc_scaled",
                              "bombus_floral_richness_scaled",
                              "bombus_floral_abun_scaled")], scale. = TRUE)
summary(pca_res)

#see which variable drives each of the principle components
print(pca_res$rotation)
```

-   PCA Results:

    PC1: grass_h_scaled (-0.50),forb_h_scaled (-0.48),grass_forb_perc_scaled (-0.42),bombus_floral_richness_scaled (-0.43),bombus_floral_abun_scaled (-0.22)

    PC2: bare_perc_scaled (0.51),bombus_floral_abun_scaled (0.52),bombus_floral_perc_scaled (0.40), dry_root_perc_scaled (-0.40)

    PC3: dry_root_perc_scaled (0.63), flood_perc_scaled (-0.64)

    Based on these PCA results, I am selecting to include the following predictors in my model:

    -   grass_forb_perc_scaled (loads strongly on PCA1)

    -   bombus_floral_abun_scaled (loads strongly on PCA1 &2)

    -   flood_perc_scaled (loads strongly on PCA3)

3)  Check pairwise correlations between this reduced set of variables

-   correlation among all the pairs are all near zero --\> predictors are all sufficiently independent!

```{r}
cor(surveys[, c("grass_forb_perc_scaled", "flood_perc_scaled", "bombus_floral_abun_scaled")])
```

# Nest-searching queens

## [vs plot type (ditch vs field)]{.underline}

### Steps

1)  Plotting a zero-inflated poisson model with multiple different random effect structures to see which is best. I used a zero-inflated poisson model since #NSQ is count data with a lot of zeros.

-   site-only was best according to maximum likelihood tests

```{r}
nsq_glm_plotType <- glmmTMB(nest_searching_queens ~ plot_type +
                              bee_season_type +
                              (1|area/site_id/plot_id),
                            ziformula = ~plot_type +
                              bee_season_type +
                              (1|area/site_id/plot_id),  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

nsq_glm_plotType_noArea <- glmmTMB(nest_searching_queens ~ plot_type +
                              bee_season_type +
                              (1|site_id/plot_id),
                            ziformula = ~plot_type +
                              bee_season_type +
                              (1|site_id/plot_id),  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

nsq_glm_plotType_plotOnly <- glmmTMB(nest_searching_queens ~ plot_type +
                              bee_season_type +
                              (1|plot_id),
                            ziformula = ~plot_type +
                              bee_season_type +
                              (1|plot_id),  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

nsq_glm_plotType_siteOnly <- glmmTMB(nest_searching_queens ~ plot_type +
                              bee_season_type +
                              (1|site_id),
                            ziformula = ~plot_type +
                              bee_season_type +
                              (1|site_id),  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

nsq_glm_plotType_areaSite <- glmmTMB(nest_searching_queens ~ plot_type +
                              bee_season_type +
                              (1|area/site_id),
                            ziformula = ~plot_type +
                              bee_season_type +
                              (1|area/site_id),  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

#Compare models
anova(nsq_glm_plotType,nsq_glm_plotType_noArea,nsq_glm_plotType_plotOnly, nsq_glm_plotType_siteOnly,nsq_glm_plotType_areaSite)
##siteOnly is best
```

2)  Check diagnostics of site-only model

-   look good

```{r}
#simulate residuals
residuals_nsq_glm_plotType_siteOnly <- simulateResiduals(fittedModel = nsq_glm_plotType_siteOnly, plot = TRUE)
```

3)  Plot results

```{r}
summary(nsq_glm_plotType_siteOnly)
```

### Plots

Conditional model:

```{r}
#Use emmeans to get predicted values on response scale
pred_nsq_plot <- emmeans(nsq_glm_plotType_siteOnly, 
               ~ plot_type * bee_season_type, 
               type = "response")

#Convert emmeans results to a dataframe for plotting
pred_nsq_plot_df <- as.data.frame(pred_nsq_plot)

#Plot predicted counts with confidence intervals
ggplot(pred_nsq_plot_df, aes(x = bee_season_type, y = rate, fill = plot_type)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL),
                position = position_dodge(width = 0.7), width = 0.2) +
  labs(y = "Predicted Count of Nest Searching Queens",
       x = "Bee Season Type",
       fill = "Plot Type",
       title = "Predicted Nest Searching Queens by Plot Type and Bee Season") +
  theme_classic()
```

Zero-inflation model:

```{r}

```

### Findings

Conditional model (plots NSQ abundance at sites where there at NSQ)

-   Significantly higher abundance of NSQ in ditch plots than in field plots.

Zero-inflation model (predicts presence/absence of NSQ at sites)

-   Significantly more zeros in worker season than in nest-searching season.

## [vs vegetation predictors]{.underline}

Plot model with interaction between vegetation variables and plot type to investigate whether the effects of vegetation cover and floral abundance on queen abundance/presence change depending on plot type (ditch vs field).

### Steps

1)  Try different random effect structures.

-   Model without random effect fit data the best.

```{r}
nsq_glm_veg_full <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled +
                              (1|area/site_id/plot_id),
                            ziformula = ~grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled+
                              (1|area/site_id/plot_id),  # zero-inflation model
                            family = poisson, 
                            data = surveys)

nsq_glm_veg_noArea <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled +
                              (1|site_id/plot_id),
                            ziformula = ~grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled+
                              (1|site_id/plot_id),  # zero-inflation model
                            family = poisson, 
                            data = surveys)

nsq_glm_veg_justSite <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled +
                              (1|site_id),
                            ziformula = ~grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled+
                              (1|site_id),  # zero-inflation model
                            family = poisson,
                            data = surveys)

nsq_glm_veg_noRE <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled,
                            ziformula = ~grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled,  # zero-inflation model
                            family = poisson,
                            data = surveys)

#compare models
anova(nsq_glm_veg_noRE, nsq_glm_veg_justSite, nsq_glm_veg_noArea, nsq_glm_veg_full)
##no random effect model is best, followed by just site model
```

2)  Check diagnostics of model

-   look good

```{r}
#simulate residuals
residuals_nsq_glm_veg_noRE <- simulateResiduals(fittedModel = nsq_glm_veg_noRE, plot = TRUE)
```

3)  Plot results

```{r}
summary(nsq_glm_veg_noRE)
```

### Plots

Grass and forb % cover

```{r}
# Define a sequence of values for grass_forb_perc_scaled across its observed range
grass_seq <- seq(
  from = min(surveys$grass_forb_perc_scaled, na.rm = TRUE),
  to = max(surveys$grass_forb_perc_scaled, na.rm = TRUE),
  length.out = 50
)

# Get predicted response for grass_forb_perc_scaled * plot_type, holding other variables at their means
pred_nsq_veg <- emmeans(nsq_glm_veg_noRE,
               ~ grass_forb_perc_scaled * plot_type,
               at = list(grass_forb_perc_scaled = grass_seq,
                         bombus_floral_abun_scaled = mean(surveys$bombus_floral_abun_scaled, na.rm=TRUE),
                         flood_perc_scaled = mean(surveys$flood_perc_scaled, na.rm=TRUE)),
               type = "response")

# Convert to dataframe for plotting
pred_nsq_veg_df <- as.data.frame(pred_nsq_veg)

ggplot(pred_nsq_veg_df, aes(x = grass_forb_perc_scaled, y = rate, color = plot_type)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL, fill = plot_type), alpha = 0.2, color = NA) +
  labs(title = "Predicted Nest Searching Queens vs Grass/Forb %",
       x = "Grass and Forb Percentage (scaled)",
       y = "Predicted Count") +
  theme_classic()
```

Floral abundance

```{r}
# Define a sequence of values for grass_forb_perc_scaled across its observed range
floral_seq <- seq(
  from = min(surveys$bombus_floral_abun_scaled, na.rm = TRUE),
  to = max(surveys$bombus_floral_abun_scaled, na.rm = TRUE),
  length.out = 50
)

# Get predicted response for grass_forb_perc_scaled * plot_type, holding other variables at their means
pred_nsq_floral <- emmeans(nsq_glm_veg_noRE,
               ~ bombus_floral_abun_scaled * plot_type,
               at = list(bombus_floral_abun_scaled = floral_seq,
                         grass_forb_perc_scaled = mean(surveys$grass_forb_perc_scaled, na.rm=TRUE),
                         flood_perc_scaled = mean(surveys$flood_perc_scaled, na.rm=TRUE)),
               type = "response")

# Convert to dataframe for plotting
pred_nsq_floral_df <- as.data.frame(pred_nsq_floral)

ggplot(pred_nsq_floral_df, aes(x = bombus_floral_abun_scaled, y = rate, color = plot_type)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL, fill = plot_type), alpha = 0.2, color = NA) +
  labs(title = "Predicted Nest Searching Queens vs Bombus Floral Abundance",
       x = "Bombus floral abundance (scaled)",
       y = "Predicted Count") +
  theme_classic()
```

### Findings

Conditional model (predicts abundance of nest-searching queens at sites where queens are present). Ditch is reference plot type.

-   Significant positive effect of grass+forb% on number of NSQ in ditch plots

-   Marginal negative effect of plot type on \# NSQ (fewer queens in field)

Zero-inflation model (predicts presence/absence of nest-searching queens).

-   Significant positive effect of floral abundance in **field** plots on NSQ presence (fewer zeros when there are more flowers in the field)

-   Marginal negative effect of floral abundance in ditch plots on NSQ presence (more zeros when there are more flowers along the ditch)

# Workers

## vs. plot type

### Steps

1)  Start with poisson and check all random effect structures.

-   model with plot nested in site (no area) is best

```{r}
workers_plotType_full <- glmmTMB(workers ~ plot_type +
                              bee_season_type +
                              (1|area/site_id/plot_id), 
                            family = poisson, #poisson
                            data = surveys)
workers_plotType_noArea <- glmmTMB(workers ~ plot_type +
                              bee_season_type +
                              (1|site_id/plot_id), 
                            family = poisson, #poisson
                            data = surveys)

workers_plotType_siteOnly <- glmmTMB(workers ~ plot_type +
                              bee_season_type +
                              (1|site_id), 
                            family = poisson, #poisson
                            data = surveys)

workers_plotType_areaSite <- glmmTMB(workers ~ plot_type +
                              bee_season_type +
                              (1|area/site_id), 
                            family = poisson, #poisson
                            data = surveys)

#compare models
anova(workers_plotType_full, workers_plotType_noArea, workers_plotType_siteOnly, workers_plotType_areaSite)
##no area is best
```

2)  Check diagnostics of model.

-   no problems detected

```{r}
#simulate residuals
residuals_workers_plotType_noArea <- simulateResiduals(fittedModel = workers_plotType_noArea, plot = TRUE)
```

3)  Plot results

```{r}
summary(workers_plotType_noArea)
```

### Plots

```{r}
# Use emmeans to get predicted values on response scale
pred_worker_plot <- emmeans(workers_plotType_noArea, 
               ~ plot_type * bee_season_type, 
               type = "response")

# Convert emmeans results to a dataframe for plotting
pred_worker_plot_df <- as.data.frame(pred_worker_plot)

# Plot predicted counts with confidence intervals
ggplot(pred_worker_plot_df, aes(x = bee_season_type, y = rate, fill = plot_type)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL),
                position = position_dodge(width = 0.7), width = 0.2) +
  labs(y = "Predicted Count of Workers",
       x = "Bee Season Type",
       fill = "Plot Type",
       title = "Predicted Worker Number by Plot Type and Bee Season") +
  theme_classic()
```

### Findings

-   significantly more workers in ditch than in field

-   significantly more workers during worker season than during NSQ season

## vs. vegetation predictors

### Steps

1)  Try random effect structures with poisson.

-   model with plot nested in site (no area) is best

```{r}
workers_veg_full <- glmmTMB(workers ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled +
                              (1|area/site_id/plot_id),
                            family = poisson, 
                            data = surveys)

workers_veg_noArea <- glmmTMB(workers ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled +
                              (1|site_id/plot_id),
                            family = poisson, 
                            data = surveys)

workers_veg_siteOnly <- glmmTMB(workers ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled +
                              (1|site_id),
                            family = poisson, 
                            data = surveys)
workers_veg_noPlot <- glmmTMB(workers ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled +
                              (1|area/site_id),
                            family = poisson, 
                            data = surveys)

#compare models
anova(workers_veg_full, workers_veg_noArea, workers_veg_siteOnly, workers_veg_noPlot)
##no area is best
```

2)  Check diagnostics

-   quantile test and KS test show significant deviations

```{r}
residuals_workers_veg_noArea <- simulateResiduals(fittedModel = workers_veg_noArea, plot = TRUE)
```

3)  Try negative binomial and check diagnostics.

-   fixes QQ plot but still significant quantile test

```{r}
workers_veg_noArea_nbin <- glmmTMB(workers ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled +
                              (1|site_id/plot_id),
                            family = nbinom2, 
                            data = surveys)

#check diagnostics
#simulate residuals
residuals_workers_veg_noArea_nbin <- simulateResiduals(fittedModel = workers_veg_noArea_nbin, plot = TRUE)
```

4)  Try zero-inflated poisson and check diagnostics.

-   diagnostic plots look good!

```{r}
workers_veg_noArea_zipois <- glmmTMB(workers ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled +
                              (1|site_id/plot_id),
                              ziformula = ~ grass_forb_perc_scaled*plot_type +
                                bombus_floral_abun_scaled*plot_type +
                                flood_perc_scaled +
                              (1|site_id/plot_id),
                            family = poisson, 
                            data = surveys)

#check diagnostics
#simulate residuals
residuals_workers_veg_noArea_zipois <- simulateResiduals(fittedModel = workers_veg_noArea_zipois, plot = TRUE)
```

5)  Plot results.

```{r}
summary(workers_veg_noArea_zipois)
```

### Plots

```{r}
# Get predicted response for grass_forb_perc_scaled * plot_type, holding other variables at their means
pred_worker_veg <- emmeans(workers_veg_noArea_zipois,
               ~ grass_forb_perc_scaled * plot_type,
               at = list(grass_forb_perc_scaled = grass_seq,
                         bombus_floral_abun_scaled = mean(surveys$bombus_floral_abun_scaled, na.rm=TRUE),
                         flood_perc_scaled = mean(surveys$flood_perc_scaled, na.rm=TRUE)),
               type = "response")

# Convert to dataframe for plotting
pred_worker_veg_df <- as.data.frame(pred_worker_veg)

ggplot(pred_worker_veg_df, aes(x = grass_forb_perc_scaled, y = rate, color = plot_type)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL, fill = plot_type), alpha = 0.2, color = NA) +
  labs(title = "Predicted Workers vs Grass/Forb %",
       x = "Grass/Forb Percentage (scaled)",
       y = "Predicted Count") +
  theme_classic()
```

Floral abundance

```{r}
# Get predicted response for grass_forb_perc_scaled * plot_type, holding other variables at their means
pred_worker_floral <- emmeans(workers_veg_noArea_zipois,
               ~ bombus_floral_abun_scaled * plot_type,
               at = list(bombus_floral_abun_scaled = floral_seq,
                         grass_forb_perc_scaled = mean(surveys$grass_forb_perc_scaled, na.rm=TRUE),
                         flood_perc_scaled = mean(surveys$flood_perc_scaled, na.rm=TRUE)),
               type = "response")

# Convert to dataframe for plotting
pred_worker_floral_df <- as.data.frame(pred_worker_floral)

ggplot(pred_worker_floral_df, aes(x = bombus_floral_abun_scaled, y = rate, color = plot_type)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL, fill = plot_type), alpha = 0.2, color = NA) +
  labs(title = "Predicted Workers vs Bombus Floral Abundance",
       x = "Bombus floral abundance (scaled)",
       y = "Predicted Count") +
  theme_minimal()
```

### Findings

No significant effect of any variables on worker abundance/presence.

# Rodents

## vs. plot type

### Steps

### Findings

## vs. vegetation predictors

### Steps

### Findings

# Floral

## vs. plot type

### Steps

1.  Try random effect structures

```{r}
#Load packages needed for CLM
library(ordinal)
library(MASS)

floral_full <- clmm(as.factor(bombus_floral_abun) ~ plot_type +
                         julian.date_scaled + 
                         (1|area/site_id/plot_id),
                       link = "logit", 
                       data = surveys)

floral_noArea <- clmm(as.factor(bombus_floral_abun) ~ plot_type +
                         julian.date_scaled + 
                         (1|site_id/plot_id),
                       link = "logit", 
                       data = surveys)

#floral_plotOnly <- clmm(as.factor(bombus_floral_abun) ~ plot_type +
 #                        julian.date_scaled + 
  #                       (1|plot_id),
   #                    link = "logit", 
    #                   data = surveys)
##can't use since there are only 2 levels

floral_siteOnly <- clmm(as.factor(bombus_floral_abun) ~ plot_type +
                         julian.date_scaled + 
                         (1|site_id),
                       link = "logit", 
                       data = surveys)

floral_AreaPlot <- clmm(as.factor(bombus_floral_abun) ~ plot_type +
                         julian.date_scaled + 
                         (1|area/plot_id),
                       link = "logit", 
                       data = surveys)
floral_noRE <- polr(as.factor(bombus_floral_abun) ~ plot_type +
                         julian.date_scaled,
                       method = "logistic", 
                       data = surveys)

#compare models AIC
AIC(floral_full, floral_noArea, floral_siteOnly, floral_AreaPlot, floral_noRE)
##shows that model without random effect has best fit
```

2.  Compare AIC between model with and without cover_type as a covariate.

-   determined that model without cover type as covariate was better

```{r}
floral_noRE_cover <- polr(as.factor(bombus_floral_abun) ~ plot_type +
                            cover_type +
                         julian.date_scaled,
                       method = "logistic", 
                       data = surveys)

AIC(floral_noRE_cover, floral_noRE)
##not including cover is better
```

3.  Try including julianDate \* plot_type

-   determined that model with interaction term was better

```{r}
floral_noRE_interaction <- polr(as.factor(bombus_floral_abun) ~ plot_type*julian.date_scaled,
                       method = "logistic", 
                       data = surveys)
AIC(floral_noRE_interaction, floral_noRE)
##interaction is better
```

4.  Look at output

```{r}
summary(floral_noRE_interaction)
```

### Plots

```{r}
# Create prediction grid
floraldata <- expand.grid(
  julian.date_scaled = seq(min(surveys$julian.date_scaled), max(surveys$julian.date_scaled), length.out = 100),
  plot_type = unique(surveys$plot_type)
)

# Calculate predicted values 
pred_floral_plot <- predict(floral_noRE_interaction, floraldata, type = "probs")

# Convert matrix to dataframe
pred_floral_plot_df <- as.data.frame(pred_floral_plot)

# Bind predictions to floral data
floraldata_pred <- cbind(floraldata, pred_floral_plot_df)

# Reshape to long format
floraldata_pred_long <- floraldata_pred %>%
  pivot_longer(cols = -c(julian.date_scaled, plot_type),
               names_to = "category",
               values_to = "probability")

# Plot
ggplot(floraldata_pred_long, aes(x = julian.date_scaled, y = probability, color = plot_type)) +
  geom_line(size = 1) +
  facet_wrap(~ category) +
  labs(
    title = "Predicted Probabilities of Bombus Floral Abundance Categories",
    x = "Julian Date (scaled)",
    y = "Probability",
    color = "Bombus Floral Abundance"
  ) +
  theme_classic()
```

Only floral abundance category 4 (1000-10000 blooms)

```{r}
# Filter to only category "4"
floraldata_pred_cat4 <- floraldata_pred %>%
  dplyr::select(julian.date_scaled, plot_type, `4` = `4`) %>%
  dplyr::rename(probability = `4`)

# Plot
ggplot(floraldata_pred_cat4, aes(x = julian.date_scaled, y = probability, color = plot_type)) +
  geom_line(size = 1.2) +
  labs(
    title = "Predicted Probability of Bombus Floral Abundance Category 4(1000-10000 blooms)",
    x = "Julian Date (scaled)",
    y = "Probability",
    color = "Plot Type"
  ) +
  theme_classic()
```

### Findings

-   Julian date is a significant predictor: floral abundance tends to increase as season progresses.

<!-- -->

-   The effect of Julian date changes depending on plot type (significant interaction).

    -   In the ditch, floral abundance increases throughout the season.

    -   In the field, floral abundance decreases throughout the season.

-   Plot type alone has no strong main effect but modifies the Julian date effect.
