---
title: "glmCH2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load and clean data

```{r}
#load packages
library(tidyverse)
library(ggplot2)
library(DHARMa)
library(glmmTMB)
library(ggeffects)
library(car)

#read in data
data <- read.csv("06_cleandata/combined_plot_data.csv")

#change character columns to factors
surveys <- data %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     bee_season_type = as.factor(bee_season_type),
     area = as.factor(area))

str(surveys)
```

Add plot type column (ditch vs field)

```{r}
surveys <- surveys %>% 
  mutate(plot_type = if_else(plot_id == "d", "ditch", "field"))

surveys <- surveys %>%
   mutate(
     plot_type = as.factor(plot_type))

```

Scale continuous predictors

```{r}
surveys$julian.date_scaled <- as.numeric(scale(surveys$julian.date))
surveys$bare_perc_scaled <- as.numeric(scale(surveys$bare_perc))
surveys$grass_perc_scaled <- as.numeric(scale(surveys$grass_perc))
surveys$forb_perc_scaled <- as.numeric(scale(surveys$forb_perc))
surveys$grass_forb_perc_scaled <- as.numeric(scale(surveys$grass_forb_perc))
surveys$grass_h_scaled <- as.numeric(scale(surveys$grass_h))
surveys$forb_h_scaled <- as.numeric(scale(surveys$forb_h))
surveys$floral_perc_scaled <- as.numeric(scale(surveys$floral_perc))
surveys$grass_forb_floral_perc_scaled <- as.numeric(scale(surveys$grass_forb_floral_perc))
surveys$bombus_floral_perc_scaled <- as.numeric(scale(surveys$bombus_floral_perc))
surveys$flood_perc_scaled <- as.numeric(scale(surveys$flood_perc))
surveys$dry_root_perc_scaled <- as.numeric(scale(surveys$dry_root_perc))
surveys$bryo_perc_scaled <- as.numeric(scale(surveys$bryo_perc))
surveys$bare_flood_dry_bryo_perc_scaled <- as.numeric(scale(surveys$bare_flood_dry_bryo_perc))
surveys$floral_richness_scaled <- as.numeric(scale(surveys$floral_richness))
surveys$bombus_floral_richness_scaled <- as.numeric(scale(surveys$bombus_floral_richness))
surveys$floral_abun_scaled <- as.numeric(scale(surveys$floral_abun))
surveys$bombus_floral_abun_scaled <- as.numeric(scale(surveys$bombus_floral_abun))
surveys$nest_searching_queens_scaled <- as.numeric(scale(surveys$nest_searching_queens))
surveys$workers_scaled <- as.numeric(scale(surveys$workers))
surveys$forage_vaco_scaled <- as.numeric(scale(surveys$forage_vaco))
surveys$num_burrows_scaled <- as.numeric(scale(surveys$num_burrows))
surveys$num_inactive_burrows_scaled <- as.numeric(scale(surveys$num_inactive_burrows))
surveys$num_active_dm_scaled <- as.numeric(scale(surveys$num_active_dm))
surveys$num_dm_size_scaled <- as.numeric(scale(surveys$num_dm_size))
surveys$num_inactive_dm_scaled <- as.numeric(scale(surveys$num_inactive_dm))
```

# GLMs #1: Number of nest-searching queens vs plot type

Plotting a zero-inflated poisson model

```{r}
nsq_glm_zeroPois <- glmmTMB(nest_searching_queens ~ plot_type +
                              bee_season_type + 
                              (1|area/site_id/plot_id),
                            ziformula = ~1,  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

summary(nsq_glm_zeroPois)

##area random effect has zero variance. remove area from model and check summary

nsq_glm_zeroPois_noArea <- glmmTMB(nest_searching_queens ~ plot_type +
                              bee_season_type + 
                              (1|site_id/plot_id),
                            ziformula = ~1,  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

summary(nsq_glm_zeroPois_noArea)
##AIC improved slightly with this model.

#compare 2 models using likelihood ratio test
anova(nsq_glm_zeroPois, nsq_glm_zeroPois_noArea)
##p=1. This tells us that adding area model does not improve model. I should go with model that drops area.

#try model without area and site
nsq_glm_zeroPois_noAreaSite <- glmmTMB(nest_searching_queens ~ plot_type +
                              bee_season_type + 
                              (1|plot_id),
                            ziformula = ~1,  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

#compare 2 models using likelihood ratio test
anova(nsq_glm_zeroPois_noAreaSite, nsq_glm_zeroPois_noArea)
##p=0.1283. This tells us that adding site does not significantly improve model. I should go with model that drops area and site since it is the simplest.

nsq_glm_zeroPois_noRE <- glmmTMB(nest_searching_queens ~ plot_type +
                              bee_season_type,
                            ziformula = ~1,  # zero-inflation model
                            family = poisson, #poisson
                            data = surveys)

#compare 2 models using likelihood ratio test
anova(nsq_glm_zeroPois_noArea, nsq_glm_zeroPois_noRE)
##p=0.e146. This tells us that adding random effects does not improve model. I should go with model without random effects since it is the simplest.

```

Diagnostics

```{r}
#simulate residuals

residuals_nsq_glm_zeroPois_noRE <- simulateResiduals(fittedModel = nsq_glm_zeroPois_noRE, plot = TRUE)

##no evidence of overdispersion or significant outliers
```

Plot results of model

```{r}
summary(nsq_glm_zeroPois_noRE)

##number of nest searching queens vs plot type
# Get predicted values
pred1 <- ggpredict(nsq_glm_zeroPois_noRE, terms = "plot_type")

# Plot
plot(pred1) + ggtitle("Predicted NSQ by plot type")

##number of nest searching queens vs season
# Get predicted values
pred2 <- ggpredict(nsq_glm_zeroPois_noRE, terms = "bee_season_type")

# Plot
plot(pred2) + ggtitle("Predicted NSQ by bee_season_type")
```

# GLM #2: NSQ vs veg variables

Choosing which vegetation variables to include as predictors by checking correlation among the variables

```{r}
vif(lm(nest_searching_queens ~ bare_perc_scaled + 
         grass_forb_perc_scaled + 
        grass_h_scaled + 
         forb_h_scaled + 
         bombus_floral_perc_scaled +
         flood_perc_scaled +
         dry_root_perc_scaled +
         bombus_floral_richness_scaled +
         bombus_floral_abun_scaled + bee_season_type, 
        data = surveys))

##all variables have VIF <5 (no multicollinearity). so i can include all these predictors in my model.
```

Run PCA to determine which vegetation variables to include

```{r}
pca_res <- prcomp(surveys[, c("bare_perc_scaled",
                              "grass_forb_perc_scaled",
                              "grass_h_scaled",
                              "forb_h_scaled",
                              "bombus_floral_perc_scaled",
                              "flood_perc_scaled",
                              "dry_root_perc_scaled",
                              "bombus_floral_richness_scaled",
                              "bombus_floral_abun_scaled")], scale. = TRUE)
summary(pca_res)

#see which variable drives each of the principle components
print(pca_res$rotation)
```

PCA Results:

PC1: grass_h_scaled (-0.50),forb_h_scaled (-0.48),grass_forb_perc_scaled (-0.42),bombus_floral_richness_scaled (-0.43),bombus_floral_abun_scaled (-0.22)

PC2: bare_perc_scaled (0.51),bombus_floral_abun_scaled (0.52),bombus_floral_perc_scaled (0.40), dry_root_perc_scaled (-0.40)

PC3: dry_root_perc_scaled (0.63), flood_perc_scaled (-0.64)

Based on these PCA results, I am selecting to include the following predictors in my model:

-   grass_forb_perc_scaled (loads strongly on PCA1)

-   bombus_floral_abun_scaled (loads strongly on PCA1 &2)

-   flood_perc_scaled (loads strongly on PCA3)

Check pairwise correlations between this reduced set of variables

```{r}
cor(surveys[, c("grass_forb_perc_scaled", "flood_perc_scaled", "bombus_floral_abun_scaled")])
##correlation among all the pairs are all near zero --> predictors are all sufficiently independent!
```

## DITCH

Filter to only include ditch data

```{r}
ditch_data <- surveys %>%
  filter(plot_id == "d")
```

Add random effects and check which is best

```{r}
nsq_ditch_zeroPois <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled +
                                bombus_floral_abun_scaled +
                                flood_perc_scaled +
                                bee_season_type +
                                (1|area/site_id/plot_id),
                            ziformula = ~1,  # zero-inflation model
                            family = poisson, #poisson
                            data = ditch_data)
#model convergence problem

diagnose(nsq_ditch_zeroPois)
##tells us that zero-inflation term may not be necessary.

nsq_ditch_Pois <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled +
                                bombus_floral_abun_scaled +
                                flood_perc_scaled +
                                bee_season_type +
                                (1|area/site_id/plot_id),
                            family = poisson, #poisson
                            data = ditch_data)

diagnose(nsq_ditch_Pois)
##try dropping flood perc

nsq_ditch_Pois <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled +
                                bombus_floral_abun_scaled +
                                bee_season_type +
                                (1|area/site_id/plot_id),
                            family = poisson, #poisson
                            data = ditch_data)

#test different random effect structures

nsq_ditch_Pois_noArea <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled +
                                bombus_floral_abun_scaled +
                                bee_season_type +
                                (1|site_id/plot_id),
                            family = poisson, #poisson
                            data = ditch_data)

anova(nsq_ditch_Pois, nsq_ditch_Pois_noArea)
##more complex model doesn't have better fit


nsq_ditch_Pois_noAreaSite <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled +
                                bombus_floral_abun_scaled +
                                bee_season_type +
                                (1|plot_id),
                            family = poisson, #poisson
                            data = ditch_data)
anova(nsq_ditch_Pois_noArea, nsq_ditch_Pois_noAreaSite)
##more complex model has better fit. go with plot nested in site
```

Diagnostics

```{r}
#simulate residuals

residuals_nsq_ditch_Pois_noArea <- simulateResiduals(fittedModel = nsq_ditch_Pois_noArea, plot = TRUE)

##there is no evidence of overdispersion or significant outliers
```

Plot results of model

```{r}
summary(nsq_ditch_Pois_noArea)

##only effect of season is significant. marginal negative effect of floral abundance on nsq
```

P0-lotting a binary model to predict if #NSQ \> 0

```{r}

#create binary variable for nsq presence
ditch_data$nsq_binary <- ifelse(ditch_data$nest_searching_queens > 0, 1, 0)

nsq_ditch_binary_allRE <- glmmTMB(nsq_binary ~ grass_forb_perc_scaled +
                                bombus_floral_abun_scaled +
                                bee_season_type + 
                              (1|area/site_id/plot_id),
                              family = binomial, 
                              data = ditch_data)

nsq_ditch_binary_noArea <- glmmTMB(nsq_binary ~ grass_forb_perc_scaled +
                                bombus_floral_abun_scaled +
                                bee_season_type + 
                              (1|site_id/plot_id),
                              family = binomial, 
                              data = ditch_data)
#no convergence

nsq_ditch_binary_noAreaSite <- glmmTMB(nsq_binary ~ grass_forb_perc_scaled +
                                bombus_floral_abun_scaled +
                                bee_season_type + 
                              (1|plot_id),
                              family = binomial, 
                              data = ditch_data)

nsq_ditch_binary_noRE <- glmmTMB(nsq_binary ~ grass_forb_perc_scaled +
                                bombus_floral_abun_scaled +
                                bee_season_type,
                                family = binomial, 
                                data = ditch_data)

AIC(nsq_ditch_binary_allRE,
    nsq_ditch_binary_noAreaSite,
    nsq_ditch_binary_noRE
    )
##model with no random effects is best

summary(nsq_ditch_binary_noRE)
##no significant predictors other than season
```

## Field veg variables

Filter to only include ditch data

```{r}
field_data <- surveys %>%
  filter(plot_id != "d")
```

Plotting a zero-inflated poisson model **(Q:** **should I check random effects for this model too?)**

```{r}
nsq_field_zeroPois <- glmmTMB(nest_searching_queens ~  grass_forb_perc_scaled +
                                bombus_floral_abun_scaled +
                                flood_perc_scaled +
                                bee_season_type +
                                (1|area/site_id/plot_id),
                            ziformula = ~1,  # zero-inflation model
                            family = poisson, #poisson
                            data = field_data)

#test different random effect structures
nsq_field_zeroPois_noArea <- glmmTMB(nest_searching_queens ~  grass_forb_perc_scaled +
                                bombus_floral_abun_scaled +
                                flood_perc_scaled +
                                bee_season_type +
                                (1|site_id/plot_id),
                            ziformula = ~1,  # zero-inflation model
                            family = poisson, #poisson
                            data = field_data)
anova(nsq_field_zeroPois, nsq_field_zeroPois_noArea)
##more complex model doesn't have better fit

nsq_field_zeroPois_noAreaSite <- glmmTMB(nest_searching_queens ~  grass_forb_perc_scaled +
                                bombus_floral_abun_scaled +
                                flood_perc_scaled +
                                bee_season_type +
                                (1|plot_id),
                            ziformula = ~1,  # zero-inflation model
                            family = poisson, #poisson
                            data = field_data)
anova(nsq_field_zeroPois_noArea, nsq_field_zeroPois_noAreaSite)
##more complex model doesn't have better fit

nsq_field_zeroPois_noRE <- glmmTMB(nest_searching_queens ~  grass_forb_perc_scaled +
                                bombus_floral_abun_scaled +
                                flood_perc_scaled +
                                bee_season_type ,
                            ziformula = ~1,  # zero-inflation model
                            family = poisson, #poisson
                            data = field_data)
anova(nsq_field_zeroPois_noAreaSite, nsq_field_zeroPois_noRE)
##more complex model doesn't have better fit.


```

Diagnostics

```{r}
#simulate residuals

residuals_nsq_field_zeroPois_noRE <- simulateResiduals(fittedModel = nsq_field_zeroPois_noRE, plot = TRUE)

##no evidence of overdispersion or significant outliers
```

Plot results of model

```{r}
summary(nsq_field_zeroPois_noRE)

##marginal positive effect of grass+forb
```

Plotting a binary model to predict if #NSQ \> 0

```{r}

#create binary variable for nsq presence
field_data$nsq_binary <- ifelse(field_data$nest_searching_queens > 0, 1, 0)

nsq_field_binary <- glmmTMB(nsq_binary ~ grass_forb_perc_scaled +
                                bombus_floral_abun_scaled +
                                flood_perc_scaled +
                                bee_season_type, 
                  family = binomial, 
                  data = field_data)

summary(nsq_field_binary)
##significant positive effect of grass+forb cover on nesting abundance
```

# 
