---
title: "GLMM Practice"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load data

```{r}
#load packages
library(tidyverse)
library(ggplot2)
library(lme4)
library(visreg)

#read in data
data <- read.csv("06_cleandata/combined_plot_data.csv")

#check that variables are in appropriate format
str(data)

#change character columns to factors
data <- data %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     area = as.factor(area))

#filter data to only include within field data
field_data <- data %>%
  filter(plot_id != "d")

#read in data
site_info <- read.csv("00_rawdata/exploratory/site-info.csv")
site_info <- site_info %>%
  rename(site_id = name)

bee <- read.csv("06_cleandata/bees_CLEAN.csv")
bee <- bee %>%
  mutate(plot_id = str_replace(plot_id, "^d $", "d"))
bee <- bee %>%
  left_join(site_info %>% select(site_id, area), by = c("site_id"))

floral <- read.csv("06_cleandata/flowers_CLEAN.csv")
floral <- floral %>%
  left_join(site_info %>% select(site_id, area), by = c("site_id"))

rodent <- read.csv("06_cleandata/rodents_CLEAN.csv")
rodent <- rodent %>%
  mutate(plot_id = str_replace(plot_id, "^d $", "d"))
rodent <- rodent %>%
  left_join(site_info %>% select(site_id, area), by = c("site_id"))

veg <- read.csv("06_cleandata/vegetation_CLEAN.csv")
veg <- veg %>%
  mutate(bare = str_replace(bare, "unknown", "")) %>%
  mutate(bare = str_replace(forb_height, " ", "")) %>%
  mutate(bare = as.numeric(bare))
veg <- veg %>%
  left_join(site_info %>% select(site_id, area), by = c("site_id"))

veg <- veg %>%
  mutate(bare = str_replace(bare, "unknown", "")) %>%
  mutate(across(contains("perc"), ~as.numeric(.)))
```

# Step 1: Check for missing data

```{r}

```

# Step 2: Look at variance among area, site, plot, and transect for each response variable to determine random effect structure that is appropriate for each response variable.

## Floral percent

```{r}
#calculate metric from veg dataset
flower_summary <- veg %>%
  rowwise() %>% #operate across rows
  mutate(row_perc_sum = sum(c_across(contains("perc")), na.rm = TRUE)) %>% #sum across the row for each column that contains "perc" (ie. all floral perc)
  ungroup() %>% # drop groups
  group_by(area,site_id, round, plot_id, transect_id) %>% 
  summarise(floral_perc = mean(row_perc_sum, na.rm = TRUE), .groups = "drop") #average across all rows in the each group


# Null model with nested random effects
null_model <- lmer(floral_perc ~ 1 + 
                (1 | area) + 
                (1 | site_id) + 
                (1 | plot_id) +
                  (1 | transect_id),
              data = flower_summary)

#check variance of each random effect
summary(null_model)
```

Based on the output of this model, I should **only keep site_id as the random effect** since it explains non-trivial variance. Drop area from model since there is zero variance among areas. Very little variance among plots, likely due to there only being 2 levels. Also almost 0 variance explained by transect.

## Burrow number

```{r}
#calculate metric from rodent dataset
rodent_summary <- rodent %>%
  group_by(area, site_id, plot_id, transect_id, transect_section) %>%
  summarise(num_burrows = n(), .groups = "drop") ##drop the groupings

# Null model with nested random effects
null_model_rod1 <- lmer(num_burrows ~ 1 + 
                (1 | area) + 
                (1 | site_id) + 
                (1 | plot_id) +
                  (1 | transect_id),
              data = rodent_summary)

#check variance of each random effect
summary(null_model_rod1)
```

This model shows that variation in the data is meaningfully described by site_id, area, and plot_id. Therefore, we should drop transect_id and summarize across plots.

## Inactive burrow number

```{r}
rodent_summary2 <- rodent %>%
  filter(powder_active_FINAL == "n") %>%
  group_by(area, site_id, plot_id, transect_id) %>%
  summarise(num_inactive_burrows = n(), .groups = "drop") ##drop the groupings

# Null model with nested random effects
null_model_rod2 <- lmer(num_inactive_burrows ~ 1 + 
                (1 | area) + 
                (1 | site_id) + 
                (1 | plot_id) +
                  (1 | transect_id),
              data = rodent_summary2)

#check variance of each random effect
summary(null_model_rod2)
```

This model shows that variation in the data is only meaningfully described by site_id.

## Num nest-searching queens

```{r}
bee_summary <- bee %>%
  filter(caste == "q" & behaviour == "nest_searching") %>%
  group_by(area, site_id, plot_id) %>%
  summarise(nest_searching_queens = n(), .groups = "drop") ##drop the groupings

null_model_bee <- lmer(nest_searching_queens ~ 1 + 
                (1 | area) + 
                (1 | site_id) + 
                (1 | plot_id),
              data = bee_summary)

#check variance of each random effect
summary(null_model_bee)
```

This model shows that variation in the data is only meaningfully described by area and plot_id.

## Num workers

```{r}
bee_summary2 <- bee %>%
  filter(caste == "w") %>%
  group_by(area, site_id, plot_id) %>%
  summarise(workers = n(), .groups = "drop") ##drop the groupings

null_model_bee2 <- lmer(workers ~ 1 + 
                (1 | area) + 
                (1 | site_id) + 
                (1 | plot_id),
              data = bee_summary2)

#check variance of each random effect
summary(null_model_bee2)
```

This model shows that variation in the data is only meaningfully described by site_id and plot_id.

## Workers foraging on vaco

```{r}
bee_summary3 <- bee %>%
  filter(flower_code == "vaco") %>%
  group_by(area, site_id, plot_id) %>%
  summarise(forage_vaco = n(), .groups = "drop")##drop the groupings

null_model_bee3 <- lmer(forage_vaco ~ 1 + 
                (1 | area) + 
                (1 | site_id) + 
                (1 | plot_id),
              data = bee_summary3)

#check variance of each random effect
summary(null_model_bee3)
```

This model shows that variation in the data is only meaningfully described by site_id and plot_id.

## Overall finding:

Variation among response variables was not explained by transect_id in any of the null models. Therefore, it is appropriate to summarize the data at the **plot level** for all survey types, rather than retaining the finer transect_id or transect_section level.

# Step 3: Scale continuous predictors

```{r}
data$julian.date_scaled <- scale(data$julian.date)
data$bare_perc_scaled <- scale(data$bare_perc)
data$grass_perc_scaled <- scale(data$grass_perc)
data$forb_perc_scaled <- scale(data$forb_perc)
data$grass_forb_perc_scaled <- scale(data$grass_forb_perc)
data$grass_h_scaled <- scale(data$grass_h)
data$forb_h_scaled <- scale(data$forb_h)
data$floral_perc_scaled <- scale(data$floral_perc)
data$grass_forb_floral_perc_scaled <- scale(data$grass_forb_floral_perc)
data$flood_perc_scaled <- scale(data$flood_perc)
data$dry_root_perc_scaled <- scale(data$dry_root_perc)
data$bryo_perc_scaled <- scale(data$bryo_perc)
data$bare_flood_dry_bryo_perc_scaled <- scale(data$bare_flood_dry_bryo_perc)
data$floral_richness_scaled <- scale(data$floral_richness)
data$nest_searching_queens_scaled <- scale(data$nest_searching_queens)
data$workers_scaled <- scale(data$workers)
data$forage_vaco_scaled <- scale(data$forage_vaco)
data$num_burrows_scaled <- scale(data$num_burrows)
data$num_inactive_burrows_scaled <- scale(data$num_inactive_burrows)
data$num_active_dm_scaled <- scale(data$num_active_dm)
data$num_dm_size_scaled <- scale(data$num_dm_size)
data$num_inactive_dm_scaled <- scale(data$num_inactive_dm)
```

# Step 4: Visualize response variables

## Floral percent

```{r}
ggplot(field_data, aes(x = floral_perc)) + 
  geom_histogram(fill = "firebrick", col = "black", closed = "left") + 
  labs(x = "% floral cover", y = "Number observations") + 
  theme_classic() +
  theme(aspect.ratio = 0.80)
```

Left skew. Zero inflation. Continuous

## Burrow number

```{r}
ggplot(field_data, aes(x = num_burrows)) + 
  geom_histogram(fill = "firebrick", col = "black", closed = "left") + 
  labs(x = "Number of burrows", y = "Number observations") + 
  theme_classic() +
  theme(aspect.ratio = 0.80)
```

Zero inflated. Count data.

## Inactive burrows

```{r}
ggplot(field_data, aes(x = num_inactive_burrows)) + 
  geom_histogram(fill = "firebrick", col = "black", closed = "left") + 
  labs(x = "Number of inactive burrows", y = "Number observations") + 
  theme_classic() +
  theme(aspect.ratio = 0.80)
```

Zero-inflated. Count data.

## Number of nest-searching queens

```{r}
ggplot(field_data, aes(x = nest_searching_queens)) + 
  geom_histogram(fill = "firebrick", col = "black", closed = "left") + 
  labs(x = "# NSQ", y = "Number observations") + 
  theme_classic() +
  theme(aspect.ratio = 0.80)
```

Zero-inflated. Count data.

## Number of workers

```{r}
ggplot(field_data, aes(x = workers)) + 
  geom_histogram(fill = "firebrick", col = "black", closed = "left") + 
  labs(x = "# workers", y = "Number observations") + 
  theme_classic() +
  theme(aspect.ratio = 0.80)
```

Zero-inflated. Count data.

## Bees foraging on vaco

```{r}
ggplot(field_data, aes(x = forage_vaco)) + 
  geom_histogram(fill = "firebrick", col = "black", closed = "left") + 
  labs(x = "# bees foraging on vaco", y = "Number observations") + 
  theme_classic() +
  theme(aspect.ratio = 0.80)
```

Zero-inflated. Count data.

# Step 5: Visualize predictors

## % cover: grass+forb

```{r}
ggplot(field_data, aes(x = site_id, y = grass_forb_perc)) +
  geom_boxplot(fill = "goldenrod1", varwidth = FALSE) +
  labs(x = "Site", y = "% cover") +
  theme_classic() +
  theme(text = element_text(size = 14))
```

## % floral

```{r}
ggplot(field_data, aes(x = site_id, y = floral_perc)) + geom_boxplot(fill = "goldenrod1", varwidth = FALSE) + labs(x = "Site", y = "% floral cover") + theme_classic() + theme(text = element_text(size = 14))
```

## burrow number

```{r}
ggplot(field_data, aes(x = site_id, y = num_burrows)) + geom_boxplot(fill = "goldenrod1", varwidth = FALSE) + labs(x = "Site", y = "# burrows") + theme_classic() + theme(text = element_text(size = 14))
```

## NSQ Abun

```{r}
ggplot(field_data, aes(x = site_id, y = nest_searching_queens)) + geom_boxplot(fill = "goldenrod1", varwidth = FALSE) + labs(x = "Site", y = "# NSQ") + theme_classic() + theme(text = element_text(size = 14))
```

## Sampling date

```{r}
ggplot(field_data, aes(x = site_id, y = julian.date)) + geom_boxplot(fill = "goldenrod1", varwidth = FALSE) + labs(x = "Site", y = "Julian date") + theme_classic() + theme(text = element_text(size = 14))
```

# Step 6: Check for correlation among predictors

```{r}
#select only columns that are numeric or integers
numeric_vars <- field_data %>% 
  select(where(~ is.numeric(.x) || is.integer(.x)))

#find correlation coefficients 
cor_matrix <- cor(numeric_vars, use = "complete.obs")
#plot
corrplot(cor_matrix, method = "circle")
```

# Step 7: Check for zero-inflation or overdispersion

```{r}

```

# Step 8: Plot bivariate relationships between each predictor and response variable

```{r}

```
