---
title: "GLMM Practice"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load data

```{r}
#load packages
library(tidyverse)
library(ggplot2)
library(lme4)
library(visreg)

#read in data
data <- read.csv("06_cleandata/combined_plot_data.csv")

#check that variables are in appropriate format
str(data)

#change character columns to factors
data <- data %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     area = as.factor(area))

#filter data to only include within field data
field_data <- data %>%
  filter(plot_id != "d")

#read in data
site_info <- read.csv("00_rawdata/exploratory/site-info.csv")
site_info <- site_info %>%
  rename(site_id = name)

bee <- read.csv("06_cleandata/bee_CLEAN.csv")
bee <- bee %>%
  mutate(plot_id = str_replace(plot_id, "^d $", "d"))
bee <- bee %>%
  left_join(site_info %>% select(site_id, area), by = c("site_id"))

floral <- read.csv("06_cleandata/floral_CLEAN.csv")
floral <- floral %>%
  left_join(site_info %>% select(site_id, area), by = c("site_id"))

rodent <- read.csv("06_cleandata/rodent_CLEAN.csv")
rodent <- rodent %>%
  mutate(plot_id = str_replace(plot_id, "^d $", "d"))
rodent <- rodent %>%
  left_join(site_info %>% select(site_id, area), by = c("site_id"))

veg <- read.csv("06_cleandata/veg_CLEAN.csv")
veg <- veg %>%
  mutate(bare = str_replace(bare, "unknown", "")) %>%
  mutate(bare = str_replace(forb_height, " ", "")) %>%
  mutate(bare = as.numeric(bare))
veg <- veg %>%
  left_join(site_info %>% select(site_id, area), by = c("site_id"))

veg <- veg %>%
  mutate(bare = str_replace(bare, "unknown", "")) %>%
  mutate(across(contains("perc"), ~as.numeric(.)))
```

# Step 1: Check for missing data

```{r}

```

# Step 2: Look at variance among area, site, plot, and transect for each response variable to determine random effect structure that is appropriate for each response variable.

## Floral abundance

```{r}
#convert floral abundance estimates so they are properly on log scale (subtract 1 from each)
floral$abundance = 10^(floral$abundance - 1)

#exponentiate the remaining values (undo the log) and add the actual abundance estimates across all species together for each plot
floral_summary <- floral %>%
  group_by(area, site_id, plot_id) %>%
  summarise(floral_abun = sum(abundance, na.rm = TRUE), .groups = "drop") ##drop the groupings

#take log scale of the summed values
floral_summary$floral_abun = log10(floral_summary$floral_abun + 1)


# Null model with nested random effects
null_model <- lmer(floral_abun ~ 1 + 
                (1 | area) + 
                (1 | site_id) + 
                (1 | plot_id),
              data = floral_summary)

#check variance of each random effect
summary(null_model)
```

Based on the output of this model, I should **only keep area and plot_id as the random effects** since they explain non-trivial variance. Drop site_id from model since there is zero variance among sites.

## Burrow number

```{r}
#calculate metric from rodent dataset
rodent_summary <- rodent %>%
  group_by(area, site_id, plot_id, transect_id, transect_section) %>%
  summarise(num_burrows = n(), .groups = "drop") ##drop the groupings

# Null model with nested random effects
null_model_rod1 <- lmer(num_burrows ~ 1 + 
                (1 | area) + 
                (1 | site_id) + 
                (1 | plot_id) +
                  (1 | transect_id),
              data = rodent_summary)

#check variance of each random effect
summary(null_model_rod1)
```

This model shows that variation in the data is meaningfully described by site_id, area, and plot_id. Therefore, we should drop transect_id and summarize across plots.

## Inactive burrow number

```{r}
rodent_summary2 <- rodent %>%
  filter(powder_active_FINAL == "n") %>%
  group_by(area, site_id, plot_id, transect_id) %>%
  summarise(num_inactive_burrows = n(), .groups = "drop") ##drop the groupings

# Null model with nested random effects
null_model_rod2 <- lmer(num_inactive_burrows ~ 1 + 
                (1 | area) + 
                (1 | site_id) + 
                (1 | plot_id) +
                  (1 | transect_id),
              data = rodent_summary2)

#check variance of each random effect
summary(null_model_rod2)
```

This model shows that variation in the data is only meaningfully described by site_id.

## Num nest-searching queens

```{r}
bee_summary <- bee %>%
  filter(caste == "q" & behaviour == "nest_searching") %>%
  group_by(area, site_id, plot_id) %>%
  summarise(nest_searching_queens = n(), .groups = "drop") ##drop the groupings

null_model_bee <- lmer(nest_searching_queens ~ 1 + 
                (1 | area) + 
                (1 | site_id) + 
                (1 | plot_id),
              data = bee_summary)

#check variance of each random effect
summary(null_model_bee)
```

This model shows that variation in the data is only meaningfully described by area and plot_id.

## Num workers

```{r}
bee_summary2 <- bee %>%
  filter(caste == "w") %>%
  group_by(area, site_id, plot_id) %>%
  summarise(workers = n(), .groups = "drop") ##drop the groupings

null_model_bee2 <- lmer(workers ~ 1 + 
                (1 | area) + 
                (1 | site_id) + 
                (1 | plot_id),
              data = bee_summary2)

#check variance of each random effect
summary(null_model_bee2)
```

This model shows that variation in the data is only meaningfully described by site_id and plot_id.

## Workers foraging on vaco

```{r}
bee_summary3 <- bee %>%
  filter(flower_code == "vaco") %>%
  group_by(area, site_id, plot_id) %>%
  summarise(forage_vaco = n(), .groups = "drop")##drop the groupings

null_model_bee3 <- lmer(forage_vaco ~ 1 + 
                (1 | area) + 
                (1 | site_id) + 
                (1 | plot_id),
              data = bee_summary3)

#check variance of each random effect
summary(null_model_bee3)
```

This model shows that variation in the data is only meaningfully described by site_id and plot_id.

## Overall finding:

Variation among response variables was not explained by transect_id in any of the null models. Therefore, it is appropriate to summarize the data at the **plot level** for all survey types, rather than retaining the finer transect_id or transect_section level.

# Step 3: Visualize response variables

## Floral abun

```{r}
ggplot(field_data, aes(x = bombus_floral_abun)) + 
  geom_histogram(fill = "firebrick", col = "black", closed = "left") + 
  labs(x = "Bombus-relevant Floral Abundance", y = "Number observations") + 
  theme_classic() +
  theme(aspect.ratio = 0.80)
```

Left skew. Zero inflation. Continuous

## Burrow number

```{r}
ggplot(field_data, aes(x = num_burrows)) + 
  geom_histogram(fill = "firebrick", col = "black", closed = "left") + 
  labs(x = "Number of burrows", y = "Number observations") + 
  theme_classic() +
  theme(aspect.ratio = 0.80)
```

Zero inflated. Count data.

## Inactive burrows

```{r}
ggplot(field_data, aes(x = num_inactive_burrows)) + 
  geom_histogram(fill = "firebrick", col = "black", closed = "left") + 
  labs(x = "Number of inactive burrows", y = "Number observations") + 
  theme_classic() +
  theme(aspect.ratio = 0.80)
```

Zero-inflated. Count data.

## Number of nest-searching queens

```{r}
ggplot(field_data, aes(x = nest_searching_queens)) + 
  geom_histogram(fill = "firebrick", col = "black", closed = "left") + 
  labs(x = "# NSQ", y = "Number observations") + 
  theme_classic() +
  theme(aspect.ratio = 0.80)
```

Zero-inflated. Count data.

## Number of workers

```{r}
ggplot(field_data, aes(x = workers)) + 
  geom_histogram(fill = "firebrick", col = "black", closed = "left") + 
  labs(x = "# workers", y = "Number observations") + 
  theme_classic() +
  theme(aspect.ratio = 0.80)
```

Zero-inflated. Count data.

## Bees foraging on vaco

```{r}
ggplot(field_data, aes(x = forage_vaco)) + 
  geom_histogram(fill = "firebrick", col = "black", closed = "left") + 
  labs(x = "# bees foraging on vaco", y = "Number observations") + 
  theme_classic() +
  theme(aspect.ratio = 0.80)
```

Zero-inflated. Count data.

# Step 4: Visualize predictors

## % cover: grass+forb

```{r}
ggplot(field_data, aes(x = site_id, y = grass_forb_perc)) +
  geom_boxplot(fill = "goldenrod1", varwidth = FALSE) +
  labs(x = "Site", y = "% cover") +
  theme_classic() +
  theme(text = element_text(size = 14))
```

## Bombus-relevant floral abun

```{r}
ggplot(field_data, aes(x = site_id, y = bombus_floral_abun)) + geom_boxplot(fill = "goldenrod1", varwidth = FALSE) + labs(x = "Site", y = "Bombus-relevant floral abundance") + theme_classic() + theme(text = element_text(size = 14))
```

## burrow number

```{r}
ggplot(field_data, aes(x = site_id, y = num_burrows)) + geom_boxplot(fill = "goldenrod1", varwidth = FALSE) + labs(x = "Site", y = "# burrows") + theme_classic() + theme(text = element_text(size = 14))
```

## NSQ Abun

```{r}
ggplot(field_data, aes(x = site_id, y = nest_searching_queens)) + geom_boxplot(fill = "goldenrod1", varwidth = FALSE) + labs(x = "Site", y = "# NSQ") + theme_classic() + theme(text = element_text(size = 14))
```

## Sampling date

```{r}
ggplot(field_data, aes(x = site_id, y = julian.date)) + geom_boxplot(fill = "goldenrod1", varwidth = FALSE) + labs(x = "Site", y = "Julian date") + theme_classic() + theme(text = element_text(size = 14))
```

# Step 5: Check for correlation among predictors

```{r}
library(corrplot)
#select only columns that are numeric or integers, except scaled variables
numeric_vars <- field_data %>% 
  select(where(~ is.numeric(.x) || is.integer(.x))) %>%
  select(-contains("_scaled"))

#find correlation coefficients 
cor_matrix <- cor(numeric_vars, use = "complete.obs")
#plot
corrplot(cor_matrix, method = "circle")
```

# Step 6: Plot bivariate relationships between each predictor and response variable

Floral abun vs. % cover

```{r}
ggplot(field_data, aes(x = grass_forb_perc, y = bombus_floral_abun)) +
  geom_point(size=2, col="red") +
  labs(y = "Bombus-relevant floral abundance (log)", x = "% cover") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```

Num burrows vs % cover

```{r}
ggplot(field_data, aes(x = grass_forb_perc, y = num_burrows)) +
  geom_point(size=2, col="red") +
  labs(y = "Burrow num", x = "% cover") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```

Num inactive burrows vs % covers

```{r}
ggplot(field_data, aes(x = grass_forb_perc, y = num_inactive_burrows)) +
  geom_point(size=2, col="red") +
  labs(y = "Inactive burrow num", x = "% cover") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```

Num NSQ vs % cover

```{r}
ggplot(field_data, aes(x = grass_forb_perc, y = nest_searching_queens)) +
  geom_point(size=2, col="red") +
  labs(y = "# NSQ", x = "% cover") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```

Num NSQ vs floral abun

```{r}
ggplot(field_data, aes(x = bombus_floral_abun, y = nest_searching_queens)) +
  geom_point(size=2, col="red") +
  labs(y = "# NSQ", x = "bombus-relevant floral abun") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```

Num workers vs floral abun

```{r}
ggplot(field_data, aes(x = bombus_floral_abun, y = workers)) +
  geom_point(size=2, col="red") +
  labs(y = "# workers", x = "bombus-relevant floral abun") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```

Num foraging on vaco vs floral abun

```{r}
ggplot(field_data, aes(x = bombus_floral_abun, y = forage_vaco)) +
  geom_point(size=2, col="red") +
  labs(y = "# foraging on vaco", x = "bombus-relevant floral abun") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```

Num NSQ vs num burrows

```{r}
ggplot(field_data, aes(x = num_burrows, y = nest_searching_queens)) +
  geom_point(size=2, col="red") +
  labs(y = "# NSQ", x = "Burrow number") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```

Num workers vs num burrows

```{r}
ggplot(field_data, aes(x = num_burrows, y = workers)) +
  geom_point(size=2, col="red") +
  labs(y = "# workers", x = "Burrow number") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```

Num forage on vaco vs num burrows

```{r}
ggplot(field_data, aes(x = num_burrows, y = forage_vaco)) +
  geom_point(size=2, col="red") +
  labs(y = "# forage on vaco", x = "Burrow number") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```

Num NSQ vs num inactive burrows

```{r}
ggplot(field_data, aes(x = num_inactive_burrows, y = nest_searching_queens)) +
  geom_point(size=2, col="red") +
  labs(y = "# NSQ", x = "Inactive burrow number") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```

Num workers vs num inactive burrows

```{r}
ggplot(field_data, aes(x = num_inactive_burrows, y = workers)) +
  geom_point(size=2, col="red") +
  labs(y = "# workers", x = "Inactive burrow number") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```

Num bees foraging on vaco vs num inactive burrows

```{r}
ggplot(field_data, aes(x = num_inactive_burrows, y = forage_vaco)) +
  geom_point(size=2, col="red") +
  labs(y = "# foraging on vaco", x = "Inactive burrow number") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```

Num workers vs num NSQ

```{r}
ggplot(field_data, aes(x = nest_searching_queens, y = workers)) +
  geom_point(size=2, col="red") +
  labs(y = "# workers", x = "# NSQ") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```

Num forage on vaco vs num NSQ

```{r}
ggplot(field_data, aes(x = nest_searching_queens, y = forage_vaco)) +
  geom_point(size=2, col="red") +
  labs(y = "# foraging on vaco", x = "# NSQ") +
  theme_classic() +
  theme(text = element_text(size = 14)) +
  geom_smooth(method = "lm", se = TRUE)
```
