---
title: "glm"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load and clean data

```{r}
#load packages
library(tidyverse)
library(ggplot2)
library(DHARMa)
library(glmmTMB)


#read in data
data <- read.csv("06_cleandata/combined_plot_data.csv")

#check that variables are in appropriate format
str(data)

#change character columns to factors
surveys <- data %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     bee_season_type = as.factor(bee_season_type),
     area = as.factor(area))

```

Scale continuous predictors

```{r}
surveys$julian.date_scaled <- as.numeric(scale(surveys$julian.date))
surveys$bare_perc_scaled <- as.numeric(scale(surveys$bare_perc))
surveys$grass_perc_scaled <- as.numeric(scale(surveys$grass_perc))
surveys$forb_perc_scaled <- as.numeric(scale(surveys$forb_perc))
surveys$grass_forb_perc_scaled <- as.numeric(scale(surveys$grass_forb_perc))
surveys$grass_h_scaled <- as.numeric(scale(surveys$grass_h))
surveys$forb_h_scaled <- as.numeric(scale(surveys$forb_h))
surveys$floral_perc_scaled <- as.numeric(scale(surveys$floral_perc))
surveys$grass_forb_floral_perc_scaled <- as.numeric(scale(surveys$grass_forb_floral_perc))
surveys$bombus_floral_perc_scaled <- as.numeric(scale(surveys$bombus_floral_perc))
surveys$flood_perc_scaled <- as.numeric(scale(surveys$flood_perc))
surveys$dry_root_perc_scaled <- as.numeric(scale(surveys$dry_root_perc))
surveys$bryo_perc_scaled <- as.numeric(scale(surveys$bryo_perc))
surveys$bare_flood_dry_bryo_perc_scaled <- as.numeric(scale(surveys$bare_flood_dry_bryo_perc))
surveys$floral_richness_scaled <- as.numeric(scale(surveys$floral_richness))
surveys$bombus_floral_richness_scaled <- as.numeric(scale(surveys$bombus_floral_richness))
surveys$floral_abun_scaled <- as.numeric(scale(surveys$floral_abun))
surveys$bombus_floral_abun_scaled <- as.numeric(scale(surveys$bombus_floral_abun))
surveys$nest_searching_queens_scaled <- as.numeric(scale(surveys$nest_searching_queens))
surveys$workers_scaled <- as.numeric(scale(surveys$workers))
surveys$forage_vaco_scaled <- as.numeric(scale(surveys$forage_vaco))
surveys$num_burrows_scaled <- as.numeric(scale(surveys$num_burrows))
surveys$num_inactive_burrows_scaled <- as.numeric(scale(surveys$num_inactive_burrows))
surveys$num_active_dm_scaled <- as.numeric(scale(surveys$num_active_dm))
surveys$num_dm_size_scaled <- as.numeric(scale(surveys$num_dm_size))
surveys$num_inactive_dm_scaled <- as.numeric(scale(surveys$num_inactive_dm))
```

Filter to only include field data

```{r}
field_data <- surveys %>%
  filter(plot_id != "d")
```

# Path diagram

![](images/IMG_8271.heic)

# GLM #1: Number of nest-searching queens

Plotting a zero-inflated poisson model

```{r}
nsq_lm_zeroPois <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled + 
                    bombus_floral_abun_scaled + 
                    num_burrows_scaled + 
                     num_inactive_burrows_scaled + 
                    bee_season_type, 
                  ziformula = ~1,  # zero-inflation model
                  family = poisson, #poisson
                  data = field_data)

summary(nsq_lm_zeroPois)
```

Diagnostics

```{r}
#simulate residuals

residuals_nsq_zeroPois <- simulateResiduals(fittedModel = nsq_lm_zeroPois, plot = TRUE)

##no evidence of overdispersion or significant outliers
```

Plotting a binary model to predict if #NSQ \> 0

```{r}

#create binary variable for nsq presence
field_data$nsq_binary <- ifelse(field_data$nest_searching_queens > 0, 1, 0)

nsq_lm_binary <- glmmTMB(nsq_binary ~ grass_forb_perc_scaled + 
                    bombus_floral_abun_scaled + 
                    num_burrows_scaled + 
                    num_inactive_burrows_scaled + 
                    bee_season_type, 
                  family = binomial, 
                  data = field_data)

summary(nsq_lm_binary)
```

# GLM #2: Number of workers

Model selection between poisson, negative binomial and zero-inflated models.

First, plot it as a poisson model.

```{r}
worker_lm_pois <- glmmTMB(workers ~ nest_searching_queens_scaled +
                       bombus_floral_abun_scaled + 
                       num_burrows_scaled + 
                       num_inactive_burrows_scaled +
                       bee_season_type, 
                     family = poisson,
                     data = field_data
                       )

summary(worker_lm_pois)
```

Run diagnostic tests

```{r}
#simulate residuals

residuals_workers_pois <- simulateResiduals(fittedModel = worker_lm_pois, plot = TRUE)

##residual plots show that there is significant overdispersion --> i should plot the negative binomial to account for this
```

Plot negative binomial

```{r}
worker_lm_negbin <- glmmTMB(workers ~ nest_searching_queens_scaled +
                       bombus_floral_abun_scaled + 
                       num_burrows_scaled + 
                       num_inactive_burrows_scaled +
                       bee_season_type, 
                     family = nbinom1,
                     data = field_data
                       )
```

Diagnostics

```{r}
#simulate residuals

residuals_workers_negbin <- simulateResiduals(fittedModel = worker_lm_negbin, plot = TRUE)

##residual plots show that there is no overdispersion and no excessive outliers. this shows that the negative binomial is a much better fit
```

Test zero-inflation

```{r}
testZeroInflation(residuals_workers_negbin)

##p-value = 1. No evidence of zero-inflation. Do not fit zero-inflated model, leave as negative binomial.
```

Look at results of final model

```{r}
summary(worker_lm_negbin)
```

# GLM #3: Number of burrows

First, plot poisson

```{r}
burrows_lm_pois <- glmmTMB(num_burrows ~ grass_forb_perc_scaled + 
                    julian.date_scaled, 
                  family = poisson, #poisson
                  data = field_data)
```

Diagnostics

```{r}
#simulate residuals

residuals_burrows_pois <- simulateResiduals(fittedModel = burrows_lm_pois, plot = TRUE)

##residual plots show that there is overdispersion and  excessive outliers. Plot negative binomial
```

Plot negative binomial to see if that fixes the overdispersion.

```{r}
burrows_lm_negbin <- glmmTMB(num_burrows ~ grass_forb_perc_scaled + 
                    julian.date_scaled, 
                  family = nbinom1, 
                  data = field_data)
```

Diagnostics

```{r}
#simulate residuals

residuals_burrows_negbin <- simulateResiduals(fittedModel = burrows_lm_negbin, plot = TRUE)

##residual plots show that there is no overdispersion or excessive outliers. Now check zero-inflation.
```

Test zero-inflation

```{r}
testZeroInflation(residuals_burrows_negbin)

##p-value = 0.256. No strong evidence of zero-inflation. I'll try the zero-inflated model to see if it helps. 
```

Plot zero-inflated negative binomial and recheck diagnostics.

```{r}
burrows_lm_zero_negbin <- glmmTMB(num_burrows ~ grass_forb_perc_scaled + 
                    julian.date_scaled, 
                    ziformula = ~1,  # zero-inflation model
                  family = nbinom1, 
                  data = field_data)
```

Diagnostics

```{r}
#simulate residuals

residuals_burrows_zero_negbin <- simulateResiduals(fittedModel = burrows_lm_zero_negbin, plot = TRUE)

##residual plots does not look significantly better than the non-zero-inflated model.


```

Compare AIC values between the 2 models

```{r}
summary(burrows_lm_negbin)
summary(burrows_lm_zero_negbin)

##AIC and BIC values are slightly lover for negative binomial and there was no strong evidence for zero-inflation. Therefore, I think the negative binomal model is the best. 
```

Results for the final model

```{r}
summary(burrows_lm_negbin)
```

# GLM #4: Number of inactive rodent burrows

First, plot poisson

```{r}
inactiveburrows_lm_pois <- glmmTMB(num_inactive_burrows ~ grass_forb_perc_scaled + 
                    julian.date_scaled, 
                  family = poisson, #poisson
                  data = field_data)
```

Diagnostics

```{r}
#simulate residuals

residuals_inactiveburrows_pois <- simulateResiduals(fittedModel = inactiveburrows_lm_pois, plot = TRUE)

##residual plots show that there is overdispersion and  excessive outliers. Plot negative binomial
```

Plot negative binomial to see if that fixes the overdispersion.

```{r}
inactiveburrows_lm_negbin <- glmmTMB(num_inactive_burrows ~ grass_forb_perc_scaled + 
                    julian.date_scaled, 
                  family = nbinom1, 
                  data = field_data)
```

Diagnostics

```{r}
#simulate residuals

residuals_inactiveburrows_negbin <- simulateResiduals(fittedModel = inactiveburrows_lm_negbin, plot = TRUE)

##residual plots show that there is no overdispersion or excessive outliers. Now check zero-inflation.
```

Test zero-inflation

```{r}
testZeroInflation(residuals_inactiveburrows_negbin)

##p-value = 0.664. No strong evidence of zero-inflation. I'll try the zero-inflated model to see if it helps. 
```

Plot zero-inflated negative binomial and recheck diagnostics.

```{r}
inactiveburrows_lm_zero_negbin <- glmmTMB(num_inactive_burrows ~ grass_forb_perc_scaled + 
                    julian.date_scaled, 
                    ziformula = ~1,  # zero-inflation model
                  family = nbinom1, 
                  data = field_data)
```

Compare AIC values between the 2 models

```{r}
summary(inactiveburrows_lm_negbin)
summary(inactiveburrows_lm_zero_negbin)

##AIC and BIC values are slightly lover for negative binomial and there was no strong evidence for zero-inflation. Therefore, I think the negative binomal model is the best. 
```

Results for the final model

```{r}
summary(inactiveburrows_lm_negbin)
```

# GLM #5: Floral abundance \*\*\*still need to work on this one

Plot gamma regression since distribution of floral abundance across sites is positive and right-skewed.

```{r}
floral_lm_tweedie <- glmmTMB(bombus_floral_abun ~ grass_forb_perc_scaled + 
                    julian.date_scaled, 
                  family = tweedie(link = "log"), ##use tweedie family instead of Gamma since it can handle zeros and positive continuous data
                  data = field_data)
```

Diagnostics

```{r}
#simulate residuals

residuals_floral_tweedie <- simulateResiduals(fittedModel = floral_lm_tweedie, plot = TRUE)

##residual plots show that there is overdispersion and  excessive outliers.
```

Try plotting the log(floral abundance) as a linear regression

```{r}
field_data$log_bombus_floral_abun <- log(field_data$bombus_floral_abun + 1e-4)

floral_lm_log <- lm(log_bombus_floral_abun ~ grass_forb_perc_scaled + julian.date_scaled, data = field_data)
```

Diagnostics

```{r}
residuals_floral_log <- simulateResiduals(fittedModel = floral_lm_log, plot = TRUE)
##fixed overdispersion and outliers. KS test suggests non-normality but I will check this in the Q-Q plot

library(car)
qqPlot(floral_lm_log, main = "Q-Q Plot")

##Q-Q plot shows that residuals are still not normal. Need to try a different model.
```

Try running Tweedie model again but adjusting for overdispersion by adding a random effect for data structure

```{r}
floral_glmm_tweedie <- glmmTMB(bombus_floral_abun ~ grass_forb_perc_scaled + 
                    julian.date_scaled + (1|area/site_id/plot_id), 
                  family = tweedie(link = "log"), 
                  data = field_data)
```

Diagnostics

```{r}
#simulate residuals

residuals_floral_tweedie_mixed <- simulateResiduals(fittedModel = floral_glmm_tweedie, plot = TRUE)

##residual plots show that there is still overdispersion. try different random effect.
```

Try simpler random effect structure (ignore site)

```{r}
floral_glmm_tweedie2 <- glmmTMB(bombus_floral_abun ~ grass_forb_perc_scaled + 
                    julian.date_scaled + (1|area/plot_id), 
                  family = tweedie(link = "log"), 
                  data = field_data)
```

Diagnostics –check for overdispersion

```{r}
sim_res <- simulateResiduals(floral_glmm_tweedie2)
plot(sim_res)
testDispersion(sim_res)

##still overdispersion. Keep original random effect structure for now but try modelling zero-inflation in model
```

Zero-inflated model

```{r}
floral_glmm_tweediezero <- glmmTMB(bombus_floral_abun ~ grass_forb_perc_scaled + 
                    julian.date_scaled + (1|area/site_id/plot_id), 
                    ziformula = ~1, #zero-inflated model
                  family = tweedie(link = "log"), 
                  data = field_data)
```

Diagnostics –check for overdispersion

```{r}
sim_res <- simulateResiduals(floral_glmm_tweediezero)
plot(sim_res)
testDispersion(sim_res)

##no more overdispersion!
```

Check other random effect structures to determine which is best

```{r}
#plot nested in area (drop site)

floral_glmm_tweediezero_plot <- glmmTMB(bombus_floral_abun ~ grass_forb_perc_scaled + 
                    julian.date_scaled + (1|area/plot_id), 
                    ziformula = ~1, #zero-inflated model
                  family = tweedie(link = "log"), 
                  data = field_data)

#site nested in area (drop plot)

floral_glmm_tweediezero_site <- glmmTMB(bombus_floral_abun ~ grass_forb_perc_scaled + 
                    julian.date_scaled + (1|area/site_id), 
                    ziformula = ~1, #zero-inflated model
                  family = tweedie(link = "log"), 
                  data = field_data)

#plot only

floral_glmm_tweediezero_plotonly <- glmmTMB(bombus_floral_abun ~ grass_forb_perc_scaled + 
                    julian.date_scaled + (1|plot_id), 
                    ziformula = ~1, #zero-inflated model
                  family = tweedie(link = "log"), 
                  data = field_data)

#site only

floral_glmm_tweediezero_siteonly <- glmmTMB(bombus_floral_abun ~ grass_forb_perc_scaled + 
                    julian.date_scaled + (1|site_id), 
                    ziformula = ~1, #zero-inflated model
                  family = tweedie(link = "log"), 
                  data = field_data)

#area only

floral_glmm_tweediezero_areaonly <- glmmTMB(bombus_floral_abun ~ grass_forb_perc_scaled + 
                    julian.date_scaled + (1|area), 
                    ziformula = ~1, #zero-inflated model
                  family = tweedie(link = "log"), 
                  data = field_data)

#check AIC
AIC(
  floral_glmm_tweediezero, 
  floral_glmm_tweediezero_plot,
  floral_glmm_tweediezero_site,
  floral_glmm_tweediezero_plotonly, 
  floral_glmm_tweediezero_siteonly, 
  floral_glmm_tweediezero_areaonly
)


##plot only model has lowest AIC--> go with that one!
```

Check overdispersion in plot only model

```{r}
sim_res <- simulateResiduals(floral_glmm_tweediezero_plotonly)
plot(sim_res)
testDispersion(sim_res)

##still no overdispersion
```

Look at summary for final model

```{r}
summary(floral_glmm_tweediezero_plotonly)
```
