---
title: "glmCH1"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load and clean data

```{r}
#load packages
library(tidyverse)
library(ggplot2)
library(DHARMa)
library(glmmTMB)

#read in data
data <- read.csv("06_cleandata/combined_plot_data.csv")

#check that variables are in appropriate format
str(data)

#change character columns to factors
surveys <- data %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     bee_season_type = as.factor(bee_season_type),
     area = as.factor(area))

```

Scale continuous predictors

```{r}
surveys$julian.date_scaled <- as.numeric(scale(surveys$julian.date))
surveys$bare_perc_scaled <- as.numeric(scale(surveys$bare_perc))
surveys$grass_perc_scaled <- as.numeric(scale(surveys$grass_perc))
surveys$forb_perc_scaled <- as.numeric(scale(surveys$forb_perc))
surveys$grass_forb_perc_scaled <- as.numeric(scale(surveys$grass_forb_perc))
surveys$grass_h_scaled <- as.numeric(scale(surveys$grass_h))
surveys$forb_h_scaled <- as.numeric(scale(surveys$forb_h))
surveys$floral_perc_scaled <- as.numeric(scale(surveys$floral_perc))
surveys$grass_forb_floral_perc_scaled <- as.numeric(scale(surveys$grass_forb_floral_perc))
surveys$bombus_floral_perc_scaled <- as.numeric(scale(surveys$bombus_floral_perc))
surveys$flood_perc_scaled <- as.numeric(scale(surveys$flood_perc))
surveys$dry_root_perc_scaled <- as.numeric(scale(surveys$dry_root_perc))
surveys$bryo_perc_scaled <- as.numeric(scale(surveys$bryo_perc))
surveys$bare_flood_dry_bryo_perc_scaled <- as.numeric(scale(surveys$bare_flood_dry_bryo_perc))
surveys$floral_richness_scaled <- as.numeric(scale(surveys$floral_richness))
surveys$bombus_floral_richness_scaled <- as.numeric(scale(surveys$bombus_floral_richness))
surveys$bombus_floral_abun_scaled <- as.numeric(scale(surveys$bombus_floral_abun))
surveys$bombus_floral_abun_nv_scaled <- as.numeric(scale(surveys$bombus_floral_abun_nv))
surveys$nest_searching_queens_scaled <- as.numeric(scale(surveys$nest_searching_queens))
surveys$workers_scaled <- as.numeric(scale(surveys$workers))
surveys$forage_vaco_scaled <- as.numeric(scale(surveys$forage_vaco))
surveys$num_burrows_scaled <- as.numeric(scale(surveys$num_burrows))
surveys$num_active_burrows_scaled <- as.numeric(scale(surveys$num_active_burrows))
surveys$num_inactive_burrows_scaled <- as.numeric(scale(surveys$num_inactive_burrows))
surveys$num_active_dm_scaled <- as.numeric(scale(surveys$num_active_dm))
surveys$num_dm_size_scaled <- as.numeric(scale(surveys$num_dm_size))
surveys$num_inactive_dm_scaled <- as.numeric(scale(surveys$num_inactive_dm))
surveys$prop_inactive_burrows_scaled <- as.numeric(scale(surveys$prop_inactive_burrows))
surveys$num_runway_burrows_scaled <- as.numeric(scale(surveys$num_runway_burrows))
```

Filter to only include field data

```{r}
field_data <- surveys %>%
  filter(plot_id != "d")

```

# GLM #1: Number of nest-searching queens

Plotting a zero-inflated poisson model

```{r}
nsq_zeroPois <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled + 
                    bombus_floral_abun_scaled + 
                    num_burrows_scaled + 
                    bee_season_type, 
                  ziformula = ~ grass_forb_perc_scaled + 
                    bombus_floral_abun_scaled + 
                    num_burrows_scaled + 
                    bee_season_type,  # zero-inflation model
                  family = poisson, #poisson
                  data = field_data)
```

Diagnostics

```{r}
#simulate residuals

residuals_nsq_zeroPois <- simulateResiduals(fittedModel = nsq_zeroPois, plot = TRUE)

##No significant problems with residual distribution (QQ plot). No signs of overdispersion or outliers. Model fits well according to DHARMa goodness-of-fit.
```

Plot model

```{r}
summary(nsq_zeroPois)
```

Interestingly, the number of rodent burrows is associated with more bombus nest-searching in the conditional model, but the number of rodent burrows is associated with more structural zeros in the zero-inflation model. This means that if a site is suitable for nesting, more rodent burrows is associated with more bombus nest-searching. However, in some cases, rodent burrows may signal inhospitable habitat conditions for bumble bees.

## Theory: habitat suitability may be related to number of inactive burrows.

First, check correlation between the number of burrows and proportion of inactive burrows.

```{r}
cor.test(surveys$num_burrows, surveys$prop_inactive_burrows, use = "complete.obs")
#significant negative correlation --> plots with more burrows tend to have a lower proportion of inactive burrows
```

Run model with proportion of inactive burrows.

```{r}
nsq_zeroPois_propInactive <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled +
                                          bombus_floral_abun_scaled +
                                       prop_inactive_burrows_scaled +
                    bee_season_type, 
                  ziformula = ~ grass_forb_perc_scaled + 
                    bombus_floral_abun_scaled + prop_inactive_burrows_scaled +
                    bee_season_type,  # zero-inflation model
                  family = poisson, #poisson
                  data = field_data)

#simulate residuals
residuals_nsq_zeroPois_propInactive <- simulateResiduals(fittedModel = nsq_zeroPois_propInactive, plot = TRUE)
##diagnostic plots show that this model fits the data well

summary(nsq_zeroPois_propInactive)
##no significant effect of proportion inactive burrows on # NSQ
```

Look at num_inactive_burrows in the model instead of overall number of burrows.

```{r}
nsq_zeroPois_numInactive <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled + 
                    bombus_floral_abun_scaled + 
                    num_inactive_burrows_scaled +
                    bee_season_type, 
                  ziformula = ~ grass_forb_perc_scaled + 
                    bombus_floral_abun_scaled + 
                    num_inactive_burrows_scaled +
                    bee_season_type,  # zero-inflation model
                  family = poisson, #poisson
                  data = field_data)

#simulate residuals
residuals_nsq_zeroPois_numInactive <- simulateResiduals(fittedModel = nsq_zeroPois_numInactive, plot = TRUE)
##diagnostic plots show that this model fits the data well

summary(nsq_zeroPois_numInactive)
```

Check model with number of active burrows

```{r}
nsq_zeroPois_numActive <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled + 
                    bombus_floral_abun_scaled + 
                    num_active_burrows_scaled +
                    bee_season_type, 
                  ziformula = ~ grass_forb_perc_scaled + 
                    bombus_floral_abun_scaled + 
                    num_active_burrows_scaled +
                    bee_season_type,  # zero-inflation model
                  family = poisson, #poisson
                  data = field_data)

#simulate residuals
residuals_nsq_zeroPois_numActive <- simulateResiduals(fittedModel = nsq_zeroPois_numActive, plot = TRUE)
##diagnostic plots show that this model fits the data well

summary(nsq_zeroPois_numActive)
```

Add runway abundance instead of num burrows to model

```{r}
nsq_runways_zeroPois <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled + 
                    bombus_floral_abun_scaled + 
                    num_runway_burrows_scaled + 
                    bee_season_type, 
                  ziformula = ~ grass_forb_perc_scaled + 
                    bombus_floral_abun_scaled + 
                    num_runway_burrows_scaled + 
                    bee_season_type,  # zero-inflation model
                  family = poisson, #poisson
                  data = field_data)

#simulate residuals
residuals_nsq_runways_zeroPois <- simulateResiduals(fittedModel = nsq_runways_zeroPois, plot = TRUE)
##looks good

#check results
summary(nsq_runways_zeroPois)
```

# GLM #2: Number of workers

Model selection between poisson, negative binomial and zero-inflated models.

First, plot it as a poisson model.

```{r}
worker_lm_pois <- glmmTMB(workers ~ nest_searching_queens_scaled +
                       bombus_floral_abun_scaled + 
                       num_burrows_scaled + 
                       bee_season_type, 
                     family = poisson,
                     data = field_data
                       )

summary(worker_lm_pois)
```

Run diagnostic tests

```{r}
#simulate residuals

residuals_workers_pois <- simulateResiduals(fittedModel = worker_lm_pois, plot = TRUE)

##residual plots show that there is significant overdispersion --> i should plot the negative binomial to account for this
```

Plot negative binomial

```{r}
worker_lm_negbin <- glmmTMB(workers ~ nest_searching_queens_scaled +
                       bombus_floral_abun_scaled + 
                       num_burrows_scaled + 
                       bee_season_type, 
                     family = nbinom2,
                     data = field_data
                       )
```

Diagnostics

```{r}
#simulate residuals

residuals_workers_negbin <- simulateResiduals(fittedModel = worker_lm_negbin, plot = TRUE)

##residual plots show that there is no overdispersion and no excessive outliers. this shows that the negative binomial is a much better fit.
```

Look at results of final model

```{r}
summary(worker_lm_negbin)
```

# GLM #3: Number of burrows

First plot poisson with different random effect structures

```{r}
burrows_full <- glmmTMB(num_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|area/site_id/plot_id),
                           family = poisson, 
                           data = field_data)


burrows_noArea <- glmmTMB(num_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|site_id/plot_id),
                           family = poisson, 
                           data = field_data)

burrows_plotOnly <- glmmTMB(num_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|plot_id),
                           family = poisson, 
                           data = field_data)

burrows_siteOnly <- glmmTMB(num_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|site_id),
                           family = poisson, 
                           data = field_data)

burrows_AreaPlot <- glmmTMB(num_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|area/plot_id),
                           family = poisson, 
                           data = field_data)


#compare models using maximum likelihood test
anova(burrows_full, burrows_noArea, burrows_plotOnly, burrows_siteOnly, burrows_AreaPlot)

##results tells us the burrows_noArea is best model (lowest AIC)
```

Check summary

```{r}
summary(burrows_noArea)
```

Diagnostics

```{r}
#simulate residuals
residuals_burrows_noArea <- simulateResiduals(fittedModel = burrows_noArea, plot = TRUE)
```

Try negative binomial

```{r}
burrows_noArea_negBin <- glmmTMB(num_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|site_id/plot_id),
                           family = nbinom1, #use nbinom1 since nbinom2 doesn't converge
                           data = field_data)
```

Diagnostics

```{r}
#simulate residuals
residuals_burrows_noArea_negBin <- simulateResiduals(fittedModel = burrows_noArea_negBin, plot = TRUE)
##looks the same as poisson, stick with poisson model
```

Try zero-inflated poisson

```{r}
burrows_zi_pois <- glmmTMB(num_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|site_id/plot_id),
                           ziformula = ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|site_id/plot_id), # zero-inflation model
                           family = poisson, 
                           data = field_data)

```

Check diagnostics

```{r}
#simulate residuals
residuals_burrows_zi <- simulateResiduals(fittedModel = burrows_zi_pois, plot = TRUE)

#worse. stick with non-zero-inflated poisson
```

Try zero-inflated negbinom

```{r}
burrows_zi_nbin <- glmmTMB(num_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|site_id/plot_id),
                           ziformula = ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|site_id/plot_id), # zero-inflation model
                           family = nbinom2, 
                           data = field_data)

#simulate residuals
residuals_burrows_zi_nbin <- simulateResiduals(fittedModel = burrows_zi_nbin, plot = TRUE)
#signficant KS test and combined adjusted quantile test
```

Add additional variable for cover type (grassy vs bare) to address quantile test issue.

```{r}
burrows_pois_cover <- glmmTMB(num_burrows ~ grass_forb_perc_scaled + 
                                         julian.date_scaled +
                                         cover_type + 
                             (1|site_id/plot_id),
                           family = poisson, 
                           data = field_data)

#simulate residuals
residuals_burrows_pois_cover <- simulateResiduals(fittedModel = burrows_pois_cover, plot = TRUE)
##looks better! 

summary(burrows_pois_cover)
##cover type improved model diagnostics but was not statistically significant (p = 0.149), suggesting habitat differences may exist but were not detectable with the current sample size.
```

Try number of burrows with runways instead.

```{r}
runwayburrows_full <- glmmTMB(num_runway_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|area/site_id/plot_id),
                             ziformula = ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|area/site_id/plot_id),
                           family = poisson, 
                           data = field_data)


runwayburrows_noArea <- glmmTMB(num_runway_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|site_id/plot_id),
                             ziformula = ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|site_id/plot_id),
                           family = poisson, 
                           data = field_data)

runwayburrows_plotOnly <- glmmTMB(num_runway_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|plot_id),
                             ziformula = ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|plot_id),
                           family = poisson, 
                           data = field_data)

runwayburrows_siteOnly <- glmmTMB(num_runway_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|site_id),
                             ziformula = ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|site_id),
                           family = poisson, 
                           data = field_data)

runwayburrows_AreaPlot <- glmmTMB(num_runway_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|area/plot_id),
                             ziformula = ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|area/plot_id),
                           family = poisson, 
                           data = field_data)
##doesnt converge

#compare models using maximum likelihood test
anova(runwayburrows_full, runwayburrows_noArea, runwayburrows_plotOnly, runwayburrows_siteOnly)
##results tells us the rodentburrows_noArea is best model (lowest AIC)

#check residuals
residuals_runwayburrows_noArea <- simulateResiduals(fittedModel = runwayburrows_noArea, plot = TRUE)
##quantile test is signficant, try adding cover covariate

runwayburrows_noArea_cover <- glmmTMB(num_runway_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                               cover_type +
                             (1|site_id/plot_id),
                             ziformula = ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                               cover_type +
                             (1|site_id/plot_id),
                           family = poisson, 
                           data = field_data)
##didn't converge

#try negative binomial
runwayburrows_noArea_nbinzi <- glmmTMB(num_runway_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|site_id/plot_id),
                             ziformula = ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|site_id/plot_id),
                           family = nbinom2, 
                           data = field_data)
##doesn't converge, try without zi 
runwayburrows_noArea_nbin <- glmmTMB(num_runway_burrows ~ grass_forb_perc_scaled + 
                             julian.date_scaled + 
                             (1|site_id/plot_id),
                           family = nbinom2, 
                           data = field_data)

#check residuals
residuals_runwayburrows_noArea_nbin <- simulateResiduals(fittedModel = runwayburrows_noArea_nbin, plot = TRUE)
##still signficant

#add cover type as covariate and remove julian date
runwayburrows_noArea_nbin_cover <- glmmTMB(num_runway_burrows ~ grass_forb_perc_scaled + 
                               cover_type +
                             (1|site_id/plot_id),
                           family = nbinom2, 
                           data = field_data)

#check residuals
residuals_runwayburrows_noArea_nbin_cover <- simulateResiduals(fittedModel = runwayburrows_noArea_nbin_cover, plot = TRUE)
##still signficant

#add zi
runwayburrows_noArea_nbinzi_cover <- glmmTMB(num_runway_burrows ~ grass_forb_perc_scaled + 
                               cover_type +
                             (1|site_id/plot_id),
                             ziformula = ~ grass_forb_perc_scaled + 
                               cover_type +
                             (1|site_id/plot_id),
                           family = nbinom2, 
                           data = field_data)
##doesn't converge

#try zi poisson without random effect
runwayburrows_noArea_poiszi_cover <- glmmTMB(num_runway_burrows ~ grass_forb_perc_scaled +
                                               cover_type,
                             ziformula = ~ grass_forb_perc_scaled +
                               cover_type,
                           family = poisson, 
                           data = field_data)

#check residuals
residuals_runwayburrows_noArea_poiszi_cover <- simulateResiduals(fittedModel = runwayburrows_noArea_poiszi_cover, plot = TRUE)
##fixed
```

# GLM #4: Floral abundance

Use cumulative link model aka ordinal logistic regression because bombus floral abundance has:

-   discrete levels (0-5 scale) representing binned counts of blooms

-   ordered categories (higher numbers represent more blooms)

-   unequal intervals between categories

Try model with different random effect structures.

```{r}
library(ordinal)
library(MASS)

floral_full <- clmm(as.factor(bombus_floral_abun) ~ grass_forb_perc_scaled +
                         julian.date_scaled + 
                         (1|area/site_id/plot_id),
                       link = "logit", 
                       data = field_data)

floral_noArea <- clmm(as.factor(bombus_floral_abun) ~ grass_forb_perc_scaled +
                         julian.date_scaled + 
                         (1|site_id/plot_id),
                       link = "logit", 
                       data = field_data)

floral_plotOnly <- clmm(as.factor(bombus_floral_abun) ~ grass_forb_perc_scaled +
                         julian.date_scaled + 
                         (1|plot_id),
                       link = "logit", 
                       data = field_data)
##can't use since there are only 2 levels

floral_siteOnly <- clmm(as.factor(bombus_floral_abun) ~ grass_forb_perc_scaled +
                         julian.date_scaled + 
                         (1|site_id),
                       link = "logit", 
                       data = field_data)

floral_AreaPlot <- clmm(as.factor(bombus_floral_abun) ~ grass_forb_perc_scaled +
                         julian.date_scaled + 
                         (1|area/plot_id),
                       link = "logit", 
                       data = field_data)
floral_noRE <- polr(as.factor(bombus_floral_abun) ~ grass_forb_perc_scaled +
                         julian.date_scaled,
                       method = "logistic", 
                       data = field_data)

#compare models AIC
AIC(floral_full, floral_noArea, floral_siteOnly, floral_AreaPlot, floral_noRE)
##shows that model without random effect has best fit

```

Try adding coverType as covariate and compare AIC score

```{r}
floral_noRE_cover <- polr(as.factor(bombus_floral_abun) ~ grass_forb_perc_scaled +
                            cover_type +
                         julian.date_scaled,
                       method = "logistic", 
                       data = field_data)

AIC(floral_noRE_cover, floral_noRE)
##not including cover is better
```

Check summary for final model

```{r}
summary(floral_noRE)
```

# GLM #5: Number of bees foraging on vaco vs vegetation variables

```{r}
forageVaco_pois <- glmmTMB(forage_vaco ~ nest_searching_queens_scaled +
                       bombus_floral_abun_scaled + 
                       num_burrows_scaled + 
                       bee_season_type, 
                     family = poisson,
                     data = field_data
                       )

#simulate residuals

residuals_forageVaco_pois <- simulateResiduals(fittedModel = forageVaco_pois, plot = TRUE)

##residual plots show that there is significant overdispersion --> i should plot the negative binomial to account for this
```

Negative binomial

```{r}
forageVaco_nbin <- glmmTMB(forage_vaco ~ nest_searching_queens_scaled +
                       bombus_floral_abun_scaled + 
                       num_burrows_scaled + 
                       bee_season_type, 
                     family = nbinom2,
                     data = field_data
                       )

#simulate residuals

residuals_forageVaco_nbin <- simulateResiduals(fittedModel = forageVaco_nbin, plot = TRUE)
#quantile test significant --> try zero-inflated poisson instead

```

Zero-inflated poisson

```{r}
forageVaco_zipois <- glmmTMB(forage_vaco ~ nest_searching_queens_scaled +
                       bombus_floral_abun_scaled + 
                       num_burrows_scaled + 
                       bee_season_type, 
                       ziformula = ~ nest_searching_queens_scaled +
                       bombus_floral_abun_scaled + 
                       num_burrows_scaled + 
                       bee_season_type,
                     family = nbinom2,
                     data = field_data
                       )

#simulate residuals

residuals_forageVaco_zipois <- simulateResiduals(fittedModel = forageVaco_zipois, plot = TRUE)
#looks good!

#check summary
summary(forageVaco_zipois)
```
