---
title: "glmCH1"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load and clean data

```{r}
#load packages
library(tidyverse)
library(ggplot2)
library(DHARMa)
library(glmmTMB)

#read in data
data <- read.csv("06_cleandata/combined_plot_data.csv")

#check that variables are in appropriate format
str(data)

#change character columns to factors
surveys <- data %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     bee_season_type = as.factor(bee_season_type),
     area = as.factor(area))

```

Scale continuous predictors

```{r}
surveys$julian.date_scaled <- as.numeric(scale(surveys$julian.date))
surveys$bare_perc_scaled <- as.numeric(scale(surveys$bare_perc))
surveys$grass_perc_scaled <- as.numeric(scale(surveys$grass_perc))
surveys$forb_perc_scaled <- as.numeric(scale(surveys$forb_perc))
surveys$grass_forb_perc_scaled <- as.numeric(scale(surveys$grass_forb_perc))
surveys$grass_h_scaled <- as.numeric(scale(surveys$grass_h))
surveys$forb_h_scaled <- as.numeric(scale(surveys$forb_h))
surveys$floral_perc_scaled <- as.numeric(scale(surveys$floral_perc))
surveys$grass_forb_floral_perc_scaled <- as.numeric(scale(surveys$grass_forb_floral_perc))
surveys$bombus_floral_perc_scaled <- as.numeric(scale(surveys$bombus_floral_perc))
surveys$flood_perc_scaled <- as.numeric(scale(surveys$flood_perc))
surveys$dry_root_perc_scaled <- as.numeric(scale(surveys$dry_root_perc))
surveys$bryo_perc_scaled <- as.numeric(scale(surveys$bryo_perc))
surveys$bare_flood_dry_bryo_perc_scaled <- as.numeric(scale(surveys$bare_flood_dry_bryo_perc))
surveys$floral_richness_scaled <- as.numeric(scale(surveys$floral_richness))
surveys$bombus_floral_richness_scaled <- as.numeric(scale(surveys$bombus_floral_richness))
surveys$bombus_floral_abun_scaled <- as.numeric(scale(surveys$bombus_floral_abun))
surveys$nest_searching_queens_scaled <- as.numeric(scale(surveys$nest_searching_queens))
surveys$workers_scaled <- as.numeric(scale(surveys$workers))
surveys$forage_vaco_scaled <- as.numeric(scale(surveys$forage_vaco))
surveys$num_burrows_scaled <- as.numeric(scale(surveys$num_burrows))
surveys$num_inactive_burrows_scaled <- as.numeric(scale(surveys$num_inactive_burrows))
surveys$num_active_dm_scaled <- as.numeric(scale(surveys$num_active_dm))
surveys$num_dm_size_scaled <- as.numeric(scale(surveys$num_dm_size))
surveys$num_inactive_dm_scaled <- as.numeric(scale(surveys$num_inactive_dm))
```

Filter to only include field data

```{r}
field_data <- surveys %>%
  filter(plot_id != "d")

```

# GLM #1: Number of nest-searching queens

Plotting a zero-inflated poisson model

```{r}
nsq_lm_zeroPois <- glmmTMB(nest_searching_queens ~ grass_forb_perc_scaled + 
                    bombus_floral_abun_scaled + 
                    num_burrows_scaled + 
                    bee_season_type, 
                  ziformula = ~ grass_forb_perc_scaled + 
                    bombus_floral_abun_scaled + 
                    num_burrows_scaled + 
                    bee_season_type,  # zero-inflation model
                  family = poisson, #poisson
                  data = field_data)

summary(nsq_lm_zeroPois)
```

Diagnostics

```{r}
#simulate residuals

residuals_nsq_zeroPois <- simulateResiduals(fittedModel = nsq_lm_zeroPois, plot = TRUE)

##no evidence of overdispersion or significant outliers
```

# GLM #2: Number of workers

Model selection between poisson, negative binomial and zero-inflated models.

First, plot it as a poisson model.

```{r}
worker_lm_pois <- glmmTMB(workers ~ nest_searching_queens_scaled +
                       bombus_floral_abun_scaled + 
                       num_burrows_scaled + 
                       bee_season_type, 
                     family = poisson,
                     data = field_data
                       )

summary(worker_lm_pois)
```

Run diagnostic tests

```{r}
#simulate residuals

residuals_workers_pois <- simulateResiduals(fittedModel = worker_lm_pois, plot = TRUE)

##residual plots show that there is significant overdispersion --> i should plot the negative binomial to account for this
```

Plot negative binomial

```{r}
worker_lm_negbin <- glmmTMB(workers ~ nest_searching_queens_scaled +
                       bombus_floral_abun_scaled + 
                       num_burrows_scaled + 
                       bee_season_type, 
                     family = nbinom1,
                     data = field_data
                       )
```

Diagnostics

```{r}
#simulate residuals

residuals_workers_negbin <- simulateResiduals(fittedModel = worker_lm_negbin, plot = TRUE)

##residual plots show that there is no overdispersion and no excessive outliers. this shows that the negative binomial is a much better fit
```

Test zero-inflation

```{r}
testZeroInflation(residuals_workers_negbin)

##p-value = 1. No evidence of zero-inflation. Do not fit zero-inflated model, leave as negative binomial.
```

Look at results of final model

```{r}
summary(worker_lm_negbin)
```

# GLM #3: Number of burrows

# GLM #4: Number of inactive rodent burrows

# GLM #5: Floral abundance

