---
title: "dataframe_setup2_doc"
output: pdf_document
date: "2025-07-07"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")

```

# Loading the dataframes

```{r}
#load packages
library(tidyverse)
library(lubridate) #package used to generate julian date from calendar date


#read in data
bee <- read.csv("06_cleandata/bees_CLEAN.csv")

bee <- bee %>%
  mutate(plot_id = str_replace(plot_id, "^d $", "d"))

floral <- read.csv("06_cleandata/flowers_CLEAN.csv")

rodent <- read.csv("06_cleandata/rodents_CLEAN.csv")
rodent <- rodent %>%
  mutate(plot_id = str_replace(plot_id, "^d $", "d"))

veg <- read.csv("06_cleandata/vegetation_CLEAN.csv")
veg <- veg %>%
  mutate(bare = str_replace(bare, "unknown", "")) %>%
  mutate(bare = str_replace(forb_height, " ", "")) %>%
  mutate(bare = as.numeric(bare))

site_info <- read.csv("00_rawdata/exploratory/site-info.csv")
site_info <- site_info %>%
  rename(site_id = name)

veg <- veg %>%
  mutate(bare = str_replace(bare, "unknown", ""))

surveys <- read.csv("06_cleandata/surveys.csv")
surveys <- surveys %>%
  rename(site_id = site)
surveys <- surveys %>%
  rename(plot_id = plots)
```

# Assign round numbers for each row in each dataset

First, I need to convert dates to dmy format that lubridate can
understand

```{r}
#Bee
bee$date <- ymd(bee$date)

#Floral
floral <- floral %>%
  mutate(date = ymd(date))

#Rodent
rodent$date_day <- dmy(rodent$date_day)

#Veg
veg$date <- dmy(veg$date) 
```

Then, I need to assign round numbers to each survey depending on the
survey date.

```{r}
rodent_r <- rodent %>%
  mutate(
    round = case_when(
      date_day < ymd("2024-06-08") ~ "2024_r1",
      date_day >= ymd("2024-06-09") & date_day <= ymd("2024-10-01") ~ "2024_r4"))

#Bee
bee_r <- bee %>%
  mutate(
    round = case_when(
      date <= ymd("2024-05-01") ~ "2024_r1",
      date >= ymd("2024-05-14") & date <= ymd("2024-06-06") ~ "2024_r2",
      date >= ymd("2024-06-18") & date <= ymd("2024-07-03") ~ "2024_r3",
      date >= ymd("2025-04-03") & date <= ymd("2025-04-19") ~ "2025_r1",
      date >= ymd("2025-04-27") & date <= ymd("2025-05-13") ~ "2025_r2",
      date > ymd("2025-05-13") ~ "2025_r3"
      ))
#Veg
veg_r <- veg %>%
  mutate(
    round = case_when(
      date < ymd("2024-05-01") ~ "2024_r1",
      date >= ymd("2024-05-14") & date <= ymd("2024-06-06") ~ "2024_r2",
      date >= ymd("2024-06-12") & date <= ymd("2024-07-03") ~ "2024_r3",
      date >= ymd("2024-07-10") & date <= ymd("2024-08-14") ~ "2024_r4",
      date >= ymd("2025-04-03") & date <= ymd("2025-04-24") ~ "2025_r1",
      date >= ymd("2025-04-27") & date <= ymd("2025-05-30") ~ "2025_r2",
      date > ymd("2025-05-30") ~ "2025_r3"
      ))

#Flowers
floral_r <- floral %>%
  mutate(
    round = case_when(
      date < ymd("2024-05-01") ~ "2024_r1",
      date >= ymd("2024-05-14") & date <= ymd("2024-06-06") ~ "2024_r2",
      date >= ymd("2024-06-12") & date <= ymd("2024-07-03") ~ "2024_r3",
      date >= ymd("2024-07-10") & date <= ymd("2024-08-14") ~ "2024_r4",
      date >= ymd("2025-04-03") & date <= ymd("2025-04-24") ~ "2025_r1",
      date >= ymd("2025-04-27") & date <= ymd("2025-05-30") ~ "2025_r2",
      date > ymd("2025-05-30") ~ "2025_r3"
      ))

```

# Add date of bee survey to the associated row survey data, based on site, plot and round.

```{r}
# Extract relevant bee dates first
bee_dates <- bee_r %>%
  select(site_id, plot_id, round, date) %>%
  rename(bee_date = date)

#Make bee_dates file only contain one row for each site/plot/round/date combination
bee_dates_unique <- bee_dates %>%
  group_by(site_id, round) %>%
  summarise(bee_date = first(bee_date), .groups = "drop") #choose first date since all dates within a site/plot/round combo are the same

## Join bee_date onto survey data
surveys <- surveys %>%
  left_join(bee_dates_unique, by = c("site_id", "round"))

#Need to manually add in the dates for 2 surveys where I didn't find any bees (surjit_168 during 2025_r2, bks_neaves during 2025_r1, bks_56 during 2024_r3)
missing_dates <- tibble::tribble(
  ~site_id,       ~round,    ~bee_date,
  "surjit_168",   "2025_r2", as.Date("2025-05-13"),  
  "bks_neaves",   "2025_r1", as.Date("2025-04-14"),
  "bks_56",   "2024_r3", as.Date("2025-06-21")
)

surveys <- surveys %>%
  rows_patch(missing_dates, by = c("site_id", "round"))

#add column for julian date
surveys$julian.date <- yday(surveys$bee_date)

```

# Add area to survey dataframe based on site_id

```{r}
surveys <- surveys %>%
  left_join(site_info, by = c("site_id"))

```

# Adding all the desired variables summarized at plot level as columns to the surveys dataframe

The random effect structure I first want to try in my models is to have
plot nested within area.Therefore, I need to get all my desired
variables to be at plot level, not transect or quadrat.

## The desired variables are:

### [Vegetation]{.underline}

-   quadrat level — conversion: Add together across all quadrats within
    a plot and divide by total area of quadrats (1m2 x 12 quadrats =
    12m\^2).

\% floral cover

```{r}

```

\% bare ground

```{r}
#create all combinations of sites, round, and plots for veg data
all_combinations_veg <- expand_grid(
  site_id = unique(veg_r$site_id),
  round = unique(veg_r$round),
  plot_id = unique(floral_r$plot_id)
)

#create summary dataframe for veg data
veg_summary <- veg_r %>%
  group_by(site_id, round, plot_id) %>%
  summarise(bare_avg = sum(as.numeric(bare), na.rm = TRUE) / 12, .groups = "drop") ##drop the groupings

#left join counts onto full combos and replace NAs with 0
veg_summary_full <- all_combinations_veg %>%
  left_join(veg_summary, by = c("site_id", "round", "plot_id")) %>%
  mutate(bare_avg = replace_na(bare_avg, 0))

#Add average % bare to surveys dataframe
surveys <- surveys %>%
  left_join(veg_summary_full, by = c("site_id", "round", "plot_id"))
```

grass height

```{r}
#create summary dataframe for floral data
veg_summary2 <- veg_r %>%
  group_by(site_id, round, plot_id) %>%
  summarise(grass_h_avg = sum(as.numeric(grass_height), na.rm = TRUE) / 12, .groups = "drop") ##drop the groupings

#left join counts onto full combos and replace NAs with 0
veg_summary2_full <- all_combinations_veg %>%
  left_join(veg_summary2, by = c("site_id", "round", "plot_id")) %>%
  mutate(grass_h_avg = replace_na(grass_h_avg, 0))

#Add average % bare to surveys dataframe
surveys <- surveys %>%
  left_join(veg_summary2_full, by = c("site_id", "round", "plot_id"))

```

forb height

```{r}
#create summary dataframe for floral data
veg_summary3 <- veg_r %>%
  group_by(site_id, round, plot_id) %>%
  summarise(forb_h_avg = sum(as.numeric(forb_height), na.rm = TRUE) / 12, .groups = "drop") ##drop the groupings

#left join counts onto full combos and replace NAs with 0
veg_summary3_full <- all_combinations_veg %>%
  left_join(veg_summary3, by = c("site_id", "round", "plot_id")) %>%
  mutate(forb_h_avg = replace_na(forb_h_avg, 0))

#Add average % bare to surveys dataframe
surveys <- surveys %>%
  left_join(veg_summary3_full, by = c("site_id", "round", "plot_id"))
```

### [Flower]{.underline}

-   plot level – no conversion needed

Num Floral species

```{r}
#create all combinations of sites, round, and plots for bee data
all_combinations <- expand_grid(
  site_id = unique(floral_r$site_id),
  round = unique(floral_r$round),
  plot_id = unique(floral_r$plot_id)
)

#create summary dataframe for floral data
floral_summary <- floral_r %>%
  group_by(site_id, round, plot_id) %>%
  summarise(floral_num = n(), .groups = "drop") ##drop the groupings

#left join counts onto full combos and replace NAs with 0
floral_summary_full <- all_combinations %>%
  left_join(floral_summary, by = c("site_id", "round", "plot_id")) %>%
  mutate(floral_num = replace_na(floral_num, 0))

#Add number of bees to surveys dataframe
surveys <- surveys %>%
  left_join(floral_summary_full, by = c("site_id", "round", "plot_id"))

```

Floral diversity

```{r}

```

### [Bee]{.underline}

-   plot level – no conversion needed

Num NS queens

```{r}
#create all combinations of sites, round, and plots for bee data
all_combinations_bee <- expand_grid(
  site_id = unique(bee_r$site_id),
  round = unique(bee_r$round),
  plot_id = unique(bee_r$plot_id)
)

#create summary dataframe for bee data
bee_summary <- bee_r %>%
  filter(caste == "q" & behaviour == "nest_searching") %>%
  group_by(site_id, round, plot_id) %>%
  summarise(nest_searching_queens = n(), .groups = "drop") ##drop the groupings

#left join counts onto full combos and replace NAs with 0
bee_summary_full <- all_combinations_bee %>%
  left_join(bee_summary, by = c("site_id", "round", "plot_id")) %>%
  mutate(nest_searching_queens = replace_na(nest_searching_queens, 0))


#Add number of bees to surveys dataframe
surveys <- surveys %>%
  left_join(bee_summary_full, by = c("site_id", "round", "plot_id"))

```

Num Workers

```{r}
bee_summary2 <- bee_r %>%
  filter(caste == "w") %>%
  group_by(site_id, round, plot_id) %>%
  summarise(workers = n(), .groups = "drop") ##drop the groupings

#left join counts onto full combos and replace NAs with 0
bee_summary2_full <- all_combinations_bee %>%
  left_join(bee_summary2, by = c("site_id", "round", "plot_id")) %>%
  mutate(workers = replace_na(workers, 0))


#Add number of bees to surveys dataframe
surveys <- surveys %>%
  left_join(bee_summary2_full, by = c("site_id", "round", "plot_id"))
```

Num foraging on VACO

```{r}
bee_summary3 <- bee_r %>%
  filter(flower_code == "vaco") %>%
  group_by(site_id, round, plot_id) %>%
  summarise(forage_vaco = n(), .groups = "drop")##drop the groupings

#left join counts onto full combos and replace NAs with 0
bee_summary3_full <- all_combinations_bee %>%
  left_join(bee_summary3, by = c("site_id", "round", "plot_id")) %>%
  mutate(forage_vaco = replace_na(forage_vaco, 0))

#Add number of bees to surveys dataframe
surveys <- surveys %>%
  left_join(bee_summary3_full, by = c("site_id", "round", "plot_id"))

```

### [Rodent]{.underline}

-   transect section level — conversion: add together \# of burrows
    across all transect sections within the same plot
-   rodent burrows were only surveyed once throughout the survey period.
    Ditch burrows were surveyed during 2024 R1. P1 and P2 burrows were
    surveyed during 2024 R4. Therefore, I want to apply the rodent data
    (eg. num_burrows, num_inactive) for each site and plot across all
    sampling rounds.

Num burrows

```{r}
#create all combinations of sites, round, and plots for bee data
all_combinations_rodent <- expand_grid(
  site_id = unique(rodent_r$site_id),
  plot_id = unique(rodent_r$plot_id)
)

#create summary dataframe for rodent data
rodent_summary <- rodent_r %>%
  group_by(site_id, plot_id) %>%
  summarise(num_burrows = n(), .groups = "drop") ##drop the groupings

#left join counts onto full combos and replace NAs with 0
rodent_summary_full <- all_combinations_rodent %>%
  left_join(rodent_summary, by = c("site_id", "plot_id")) %>%
  mutate(num_burrows = replace_na(num_burrows, 0))

#Add number of burrows to surveys dataframe
surveys <- surveys %>%
  left_join(rodent_summary_full, by = c("site_id", "plot_id"))
```

Num inactive burrows

```{r}
#create summary dataframe for rodent data
rodent_summary2 <- rodent_r %>%
  filter(powder_active_FI..L == "n") %>%
  group_by(site_id, plot_id) %>%
  summarise(num_inactive_burrows = n(), .groups = "drop") ##drop the groupings

#left join counts onto full combos and replace NAs with 0
rodent_summary2_full <- all_combinations_rodent %>%
  left_join(rodent_summary2, by = c("site_id", "plot_id")) %>%
  mutate(num_inactive_burrows = replace_na(num_inactive_burrows, 0))

#Add number of bees to surveys dataframe
surveys <- surveys %>%
  left_join(rodent_summary2_full, by = c("site_id", "plot_id"))
```

# Save final combined surveys dataframe as new CSV

```{r}
write_csv(surveys, file = file.path("06_cleandata","combined_plot_data.csv"))
```

# 
