---
title: "rodent_burrow_over_time"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load data

```{r}
#load package
library(tidyverse)   #data manipulation, visualization etc. (dplyr, ggplot2, tidyr)

#read in data
rodent <- read.csv("06_cleandata/rodent_cleanv2.csv")

#add plot type
rodent <- rodent %>%
  mutate(plot_type = if_else(plot_id == "d", "ditch", "field")) %>%
  mutate(plot_type = as.factor(plot_type)) 

#correct round names
rodent <- rodent %>%
  filter(round != "")

#check
unique(rodent$round)

#Add area names to dataset
site_info <- read.csv("06_cleandata/site_info_clean.csv")

rodent <- rodent %>%
  left_join(site_info %>% dplyr::select(site_id, area, cover_type), by = c("site_id"))

```

# Plot difference in rodent burrow abundance between R1 and R4

## 1. Create dataframe for r1 and r4 rodent burrow surveys at all sites

```{r}
#create vector for site names
site_names <- c("bates_57b", "bks_56", "chong_68", "bks_68", "chong_96", "bks_96", "gwb_neaves", "bks_neaves", "avtar_168", "surjit_168", "sandhu_ladnertrunk", "lbf_ladnertrunk")

#create vector for round numbers
round <-c("2024_r1", "2024_r4")


#combine vectors into dataframe 
samples_plot <- tibble(site = site_names) %>%
  expand_grid(round)

#remove rows from 2024 NS bee round and 2024 R1 veg for Sandhu farms and LBF farms (not sampled during that round)
surveys <- samples_plot %>%
  filter(!(site == "sandhu_ladnertrunk" & round == "2024_r1"))
surveys <- surveys %>%
  filter(!(site == "lbf_ladnertrunk" & round == "2024_r1"))

str(surveys)
surveys <- surveys %>%
  rename(site_id = site)

surveys <- surveys %>%
  left_join(site_info %>% select(site_id, area, cover_type), by = c("site_id"))

surveys <- surveys %>%
   mutate(
     cover_type = as.factor(cover_type),
     cover_type = fct_recode(cover_type, "bare" = "bare", "herbaceous" = "grass")
   )

```

## 2. Generate burrow counts for each site within each round

```{r}
#filter rodent data to only within field
rodent_field <- rodent %>%
  filter(plot_type == "field") 

#create summary dataframe for rodent data
rodent_summary <- rodent_field %>%
  group_by(site_id, round) %>%
  summarise(num_burrows = n(), .groups = "drop") ##drop the groupings

#Add number of burrows to surveys dataframe
surveys <- surveys %>%
  left_join(rodent_summary, by = c("site_id", "round")) %>%
   mutate(num_burrows = replace_na(num_burrows, 0))

#remove sandhu ladnertrunk and lbf ladnertrunk since they were only recorded in r4
surveys <- surveys %>%
  filter(!(site_id == "sandhu_ladnertrunk"))
surveys <- surveys %>%
  filter(!(site_id == "lbf_ladnertrunk"))

#create bare vs grass surveys
bare_surveys <- surveys %>%
  filter(cover_type == "bare")
  
grass_surveys <- surveys %>%
  filter(cover_type == "herbaceous")



```

## 3. Plot

```{r}
# Plotting the trajectories
ggplot(surveys, aes(x = round, y = num_burrows, group = site_id)) +
  # 1. Draw the connecting lines first so they sit behind the points
  geom_line(color = "grey50", alpha = 0.6, size = 0.8) +
  
  # 2. Draw the points on top (using shape = 21 so they can take a 'fill' color)
  geom_point(shape = 21, size = 5, fill = "black", stroke = 1) +
  labs(
    x = "Sampling round",
    y = "Within-field burrow abundance",
    fill = "Cover type"
  ) +
  # 3. Split into panels by cover type
  facet_wrap(~cover_type) +
  theme_classic(base_size = 20)

```

## 4. Stats test

```{r}
#All sites
#check significance using paired wilcoxon signed-rank test
#Reshape data to wide format (one row per site, columns for R1 and R4)
rodent_wide <- surveys %>%
  pivot_wider(
    id_cols = c(site_id, cover_type, area), 
    names_from = round, 
    values_from = num_burrows,
    values_fill = 0 # If a site had no burrows in a round, it's a 0
  )

# 2. Run the paired test
wilcox.test(rodent_wide$`2024_r1`, rodent_wide$`2024_r4`, paired = TRUE)
##p=0.6236. We fail to reject the null hypothesisi that true location shift is equal to zero. 
##Therefore, we did not find a significant change in rodent burrow abundance over the course of the season.

#bare sites
rodent_wide_bare <- rodent_wide %>%
  filter(cover_type == "bare")
# 2. Run the paired test
wilcox.test(rodent_wide_bare$`2024_r1`, rodent_wide_bare$`2024_r4`, paired = TRUE)
##p=0.5839. We fail to reject the null hypothesisi that true location shift is equal to zero. 
##Therefore, we did not find a significant change in rodent burrow abundance over the course of the season.

#herbaceous sites
rodent_wide_herb <- rodent_wide %>%
  filter(cover_type == "herbaceous")
# 2. Run the paired test
wilcox.test(rodent_wide_herb$`2024_r1`, rodent_wide_herb$`2024_r4`, paired = TRUE)
##p=0.8551. We fail to reject the null hypothesisi that true location shift is equal to zero. 
##Therefore, we did not find a significant change in rodent burrow abundance over the course of the season.

```

# Plot the difference between ditch and field over time.

## 1. Create dataframe for all rodent surveys and plot ids in r1

```{r}
#filter to round 1 data (since that has ditch and within field measurements)
rodent_r1 <- rodent %>%
  filter(round == "2024_r1")

#calculate number of burrows in ditch and average number of burrows in field plots at each site 
#create vector for site names not "sandhu_ladnertrunk" and "lbf_ladnertrunk" since they were not sampled during that round
site_names2 <- c("bates_57b", "bks_56", "chong_68", "bks_68", "chong_96", "bks_96", "gwb_neaves", "bks_neaves", "avtar_168", "surjit_168")

#create vector for round numbers
plot_type <-c("ditch", "field")


#combine vectors into dataframe 
samples_plot2 <- tibble(site_id = site_names2) %>%
  expand_grid(plot_type)

samples_plot2 <- samples_plot2 %>%
  left_join(site_info %>% select(site_id, area, cover_type), by = c("site_id"))

surveys2 <- samples_plot2 %>%
   mutate(
     cover_type = as.factor(cover_type),
     cover_type = fct_recode(cover_type, "bare" = "bare", "herbaceous" = "grass")
   )

#add sampling date
site_dates <- rodent_r1 %>%
  group_by(site_id, plot_type) %>%
  summarise(
    date = first(date_day),
    .groups = "drop"
  )
surveys2 <- surveys2 %>%
  # Pull the dates in by matching site_id
  left_join(site_dates, by = c("site_id", "plot_type"))

library(lubridate) #date <-> julian date

surveys2 <- surveys2 %>%
   mutate(date = as.Date(date),
          date = case_when( 
            site_id == "bks_68" & plot_type == "ditch"   ~ as.Date("2024-05-04"),
            site_id == "bks_96" & plot_type == "field" ~ as.Date("2024-05-06"),
            TRUE ~ date # Keeps the original joined date for everything else
            ))
```

## 2. Calculate avg rodent abundance by plot type

```{r}
#create summary dataframe for rodent data
rodent_summary2 <- rodent_r1 %>%
  group_by(site_id, plot_id, plot_type) %>%
  summarise(num_burrows = n(), .groups = "drop") %>%
  group_by(site_id, plot_type) %>%
  summarise(avg_burrows = mean(num_burrows), .groups = "drop")

#Add number of burrows to surveys dataframe
surveys2 <- surveys2 %>%
  left_join(rodent_summary2, by = c("site_id", "plot_type")) %>%
   mutate(avg_burrows = replace_na(avg_burrows, 0))


```

## 3. Calculate difference in burrow abundance between ditch and field for each site

```{r}
burrow_differences <- surveys2 %>%
  # 1. Keep only the columns needed for this calculation
  select(site_id, cover_type, plot_type, avg_burrows) %>%
  
  # 2. Pivot plot_type values (ditch, field) into their own columns
  pivot_wider(
    names_from = plot_type, 
    values_from = avg_burrows
  ) %>%
  
  # 3. Calculate the difference (e.g., Ditch minus Field)
  mutate(
    ditch_field_diff = ditch - field
  )

#add date to dataframe
# 1. Pivot your dates wide so each site has exactly one row
site_dates_wide <- site_dates %>%
  mutate(plot_type = paste0("date_", plot_type)) %>% # turns into "date_ditch" and "date_field"
  pivot_wider(names_from = plot_type, values_from = date)

# 2. Join it seamlessly to your differences dataframe
burrow_differences <- burrow_differences %>%
  left_join(site_dates_wide, by = "site_id") %>%
  mutate(date_ditch = as.Date(date_ditch),
         date_field = as.Date(date_field),
         date_ditch = case_when(site_id == "bks_68" & cover_type == "bare"   ~ as.Date("2024-05-04"),
            TRUE ~ date_ditch), # Keeps the original joined date for everything else
         date_field = case_when(site_id == "bks_96" & cover_type == "bare"   ~ as.Date("2024-05-06"),
            TRUE ~ date_field # Keeps the original joined date for everything else
            )
         )


range(burrow_differences$date_ditch) ##april 17-may 6
range(burrow_differences$date_field) ##april 1 - may 23
```

## 4. Plot

```{r}
# Plotting the trajectories
ggplot(burrow_differences, aes(x = date_field, y = ditch_field_diff)) +
  geom_line(color = "grey50", alpha = 0.6, size = 0.8) +
  geom_point(shape = 21, size = 5, fill = "black", stroke = 1) +
  labs(
    x = "Survey date",
    y = "Ditch burrow - field burrow",
    fill = "Cover type"
  ) +
  geom_smooth(method = "lm", color = "blue", fill = "blue", alpha = 0.15, se = TRUE) +
  # 3. Split into panels by cover type
  facet_wrap(~cover_type) +
  theme_classic(base_size = 20)
```

## 5. Stats

```{r}
#filter for each cover type
bare_diffs <- burrow_differences %>% filter(cover_type == "bare")
grass_diffs <- burrow_differences %>% filter(cover_type == "herbaceous")

# Non-parametric trend test
cor.test(~ ditch_field_diff + as.numeric(date_field), data = bare_diffs, method = "kendall")
##
cor.test(~ ditch_field_diff + as.numeric(date_field), data = grass_diffs, method = "kendall")
```

# Plot the difference between grass and bare over time (within-field plots only).

## 1. Create dataframe for rodent surveys for all sites and rounds

```{r}
#filter to round 1 data (since that has ditch and within field measurements)
rodent_field <- rodent %>%
  filter(plot_type == "field")

#create vector for site names not "sandhu_ladnertrunk" and "lbf_ladnertrunk" since they were not sampled during round 1
site_names2 <- c("bates_57b", "bks_56", "chong_68", "bks_68", "chong_96", "bks_96", "gwb_neaves", "bks_neaves", "avtar_168", "surjit_168")

#create vector for round
round <-c("2024_r1", "2024_r4")


#combine vectors into dataframe 
samples_plot3 <- tibble(site_id = site_names2) %>%
  expand_grid(round)

samples_plot3 <- samples_plot3 %>%
  left_join(site_info %>% select(site_id, area, cover_type), by = c("site_id"))

surveys3 <- samples_plot3 %>%
   mutate(
     cover_type = as.factor(cover_type),
     cover_type = fct_recode(cover_type, "bare" = "bare", "herbaceous" = "grass")
   )
```

## 2. Calculate burrow abundance in each area/site/covertype/round

```{r}
#create summary dataframe for rodent data
rodent_summary3 <- rodent_field %>%
  group_by(site_id, cover_type, area, round) %>%
  summarise(num_burrows = n(), .groups = "drop")

rodent_summary3 <- rodent_summary3 %>%
   mutate(
     cover_type = as.factor(cover_type),
     cover_type = fct_recode(cover_type, "bare" = "bare", "herbaceous" = "grass")
   )

#Add number of burrows to surveys dataframe
surveys3 <- surveys3 %>%
  left_join(rodent_summary3, by = c("site_id", "area", "cover_type", "round")) %>%
   mutate(num_burrows = replace_na(num_burrows, 0))
```

## 3. Calculate difference between bare and grassy sites within each area

```{r}
burrow_differences_covertype <- surveys3 %>%
  # 1. Keep only the columns needed for this calculation
  select(area, cover_type, round, num_burrows) %>%
  
  # 2. Pivot plot_type values (ditch, field) into their own columns
  pivot_wider(
    names_from = cover_type, 
    values_from = num_burrows
  ) %>%
  
  # 3. Calculate the difference (e.g., Ditch minus Field)
  mutate(
    herb_bare_diff = herbaceous - bare
  )
```

## 4. Plot

```{r}
ggplot(burrow_differences_covertype, aes(x = round, y = herb_bare_diff)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.5) +
  # 2. Add jittered points so you can see the individual 'area' blocks
  geom_jitter(width = 0.1, size = 3, alpha = 0.8, color = "black", shape = 21, fill = "white") +
  # 3. Add a reference line at 0 (where herbaceous count == bare count)
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.8) +
  labs(
    x = "Sampling round",
    y = "Burrow abundance difference\nbetween cover types\n(herbaceous - bare)"
  ) +
  # Using a clean theme and removing the redundant legend
  theme_classic(base_size = 18) 
```

# Plot difference in burrow abundance between grassy and bare sites over date (only in R2) 

## 1. Add date to burrow dataframe

```{r}
# 1. Create a reference dataframe containing just the herbaceous dates for each area
herb_dates_r2 <- rodent %>%
  filter(round == "2024_r4" & plot_type == "field" & cover_type == "grass") %>%
  select(area, herb_date = date_day) %>%
  distinct(area, .keep_all = TRUE)

burrow_differences_covertype_r2 <- surveys3 %>%
  # 1. Filter explicitly to only Round 1 data
  filter(round == "2024_r4") %>%
  left_join(herb_dates_r2,by = "area") %>%
  # 3. Clean up columns before the pivot 
  select(area, cover_type, herb_date, num_burrows) %>%
  # 4. Pivot cover type wide to calculate the pairing difference
  pivot_wider(
    names_from = cover_type, 
    values_from = num_burrows
  ) %>%
  # 5. Calculate your treatment difference
  mutate(
    herb_bare_diff = herbaceous - bare
  )
  
```

## 2. Plot

```{r}
ggplot(burrow_differences_covertype_r2, aes(x = herb_date, y = herb_bare_diff)) +
  geom_line(color = "grey50", alpha = 0.6, size = 0.8) +
  geom_point(shape = 21, size = 5, fill = "black", stroke = 1) +
  labs(
    x = "Survey date",
    y = "Burrow count difference\nbetween cover types"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.8) +
  theme_classic(base_size = 20) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```
