---
title: "rodent_burrow_over_time"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load data

```{r}
#load package
library(tidyverse)   #data manipulation, visualization etc. (dplyr, ggplot2, tidyr)
library(lubridate) #date <-> julian date


#read in data
rodent <- read.csv("06_cleandata/rodent_cleanv2.csv")

#add plot type
rodent <- rodent %>%
  mutate(plot_type = if_else(plot_id == "d", "ditch", "field")) %>%
  mutate(plot_type = as.factor(plot_type)) 

#correct round names
rodent <- rodent %>%
  filter(round != "") %>%
  # change r4 to r2
  mutate(round = ifelse(round == "2024_r4", "2024_r2", round))

#check
unique(rodent$round)

#Add area names to dataset
site_info <- read.csv("06_cleandata/site_info_clean.csv")

rodent <- rodent %>%
  left_join(site_info %>% dplyr::select(site_id, area, cover_type), by = c("site_id")) %>%
  mutate(area = ifelse(area == "pittmeadows_1", "pittmeadows", area)) %>%
  mutate(area = ifelse(area == "surrey_1", "surrey", area))


#load date data
rodent_dates_dVf <- read.csv("06_cleandata/rodent_dates_dVf.csv")

rodent_dates_fVb <- read.csv("06_cleandata/rodent_dates_fVb.csv")
```

# Plot difference in rodent burrow abundance between R1 and R4

## 1. Create dataframe for r1 and r4 rodent burrow surveys at all sites

```{r}
#create vector for site names
site_names <- c("bates_57b", "bks_56", "chong_68", "bks_68", "chong_96", "bks_96", "gwb_neaves", "bks_neaves", "avtar_168", "surjit_168", "sandhu_ladnertrunk", "lbf_ladnertrunk")

#create vector for round numbers
round <-c("2024_r1", "2024_r2")


#combine vectors into dataframe 
samples_plot <- tibble(site = site_names) %>%
  expand_grid(round)

#remove rows from 2024 NS bee round and 2024 R1 veg for Sandhu farms and LBF farms (not sampled during that round)
surveys <- samples_plot %>%
  filter(!(site == "sandhu_ladnertrunk" & round == "2024_r1"))
surveys <- surveys %>%
  filter(!(site == "lbf_ladnertrunk" & round == "2024_r1"))

#rename site column
surveys <- surveys %>%
  rename(site_id = site)

#add area and cover type to df
surveys <- surveys %>%
  left_join(site_info %>% select(site_id, area, cover_type), by = c("site_id"))

#update cover type categories
surveys <- surveys %>%
   mutate(
     cover_type = as.factor(cover_type),
     cover_type = fct_recode(cover_type, "bare" = "bare", "herbaceous" = "grass")
   )

```

## 2. Generate burrow counts for each site within each round

Only include within-field data since ditch data was only collected in R1

```{r}
#filter rodent data to only within field, 
rodent_field <- rodent %>%
  filter(plot_type == "field") 

#create summary dataframe for rodent data
rodent_summary <- rodent_field %>%
  group_by(site_id, round) %>%
  summarise(num_burrows = n(), .groups = "drop") ##drop the groupings

#Add number of burrows to surveys dataframe
surveys <- surveys %>%
  left_join(rodent_summary, by = c("site_id", "round")) %>%
   mutate(num_burrows = replace_na(num_burrows, 0))

#remove sandhu ladnertrunk and lbf ladnertrunk since they were only recorded in r2
surveys <- surveys %>%
  filter(!(site_id == "sandhu_ladnertrunk"))
surveys <- surveys %>%
  filter(!(site_id == "lbf_ladnertrunk"))

#create bare vs grass surveys dfs
bare_surveys <- surveys %>%
  filter(cover_type == "bare")
  
grass_surveys <- surveys %>%
  filter(cover_type == "herbaceous")
```

## 3. Plot

```{r}
# Plotting the trajectories
ggplot(surveys, aes(x = round, y = num_burrows, group = site_id)) +
  # 1. Draw the connecting lines first so they sit behind the points
  geom_line(color = "grey50", alpha = 0.6, size = 0.8) +
  
  # 2. Draw the points on top (using shape = 21 so they can take a 'fill' color)
  geom_point(shape = 21, size = 5, fill = "black", stroke = 1) +
  labs(
    x = "Sampling round",
    y = "Within-field burrow abundance",
    fill = "Cover type"
  ) +
  # 3. Split into panels by cover type
  facet_wrap(~cover_type) +
  theme_classic(base_size = 20)

```

## 4. Stats test

```{r}
#All sites
#check significance using paired wilcoxon signed-rank test
#Reshape data to wide format (one row per site, columns for R1 and R4)
rodent_wide <- surveys %>%
  pivot_wider(
    id_cols = c(site_id, cover_type, area), 
    names_from = round, 
    values_from = num_burrows,
    values_fill = 0 # If a site had no burrows in a round, it's a 0
  )

# 2. Run the paired test
wilcox.test(rodent_wide$`2024_r1`, rodent_wide$`2024_r2`, paired = TRUE)
##p=0.6236. We fail to reject the null hypothesisi that true location shift is equal to zero. 
##Therefore, we did not find a significant change in rodent burrow abundance over the course of the season.

#bare sites
rodent_wide_bare <- rodent_wide %>%
  filter(cover_type == "bare")
# 2. Run the paired test
wilcox.test(rodent_wide_bare$`2024_r1`, rodent_wide_bare$`2024_r2`, paired = TRUE)
##p=0.5839. We fail to reject the null hypothesisi that true location shift is equal to zero. 
##Therefore, we did not find a significant change in rodent burrow abundance over the course of the season.

#herbaceous sites
rodent_wide_herb <- rodent_wide %>%
  filter(cover_type == "herbaceous")
# 2. Run the paired test
wilcox.test(rodent_wide_herb$`2024_r1`, rodent_wide_herb$`2024_r2`, paired = TRUE)
##p=0.8551. We fail to reject the null hypothesisi that true location shift is equal to zero. 
##Therefore, we did not find a significant change in rodent burrow abundance over the course of the season.

```

# Plot the difference between ditch and field over time.

## 1. Create dataframe for all rodent surveys and plot ids in r1

```{r}
#calculate number of burrows in ditch and average number of burrows in field plots at each site 
#create vector for site names not "sandhu_ladnertrunk" and "lbf_ladnertrunk" since they were not sampled during that round
site_names2 <- c("bates_57b", "bks_56", "chong_68", "bks_68", "chong_96", "bks_96", "gwb_neaves", "bks_neaves", "avtar_168", "surjit_168")

#create vector for plot type
plot_type <-c("ditch", "field")


#combine vectors into dataframe 
samples_plot2 <- tibble(site_id = site_names2) %>%
  expand_grid(plot_type)

#add area and cover type
samples_plot2 <- samples_plot2 %>%
  left_join(site_info %>% select(site_id, area, cover_type), by = c("site_id"))

#fix cover type formatting
surveys2 <- samples_plot2 %>%
   mutate(
     cover_type = as.factor(cover_type),
     cover_type = fct_recode(cover_type, "bare" = "bare", "herbaceous" = "grass")
   )

#add sampling date
site_dates <- rodent_dates_dVf %>%
  group_by(site_id, plot_type) %>%
  summarise(
    date = first(date), #select first date for each site/plot type combo so that there is one date for field and one for bare for each site within this round
    .groups = "drop"
  )
surveys2 <- surveys2 %>%
  # Pull the dates in by matching site_id
  left_join(site_dates, by = c("site_id", "plot_type"))

```

## 2. Calculate avg rodent abundance by plot type

```{r}
#filter to round 1 data (since that has ditch and within field measurements)
rodent_r1 <- rodent %>%
  filter(round == "2024_r1")

#create summary dataframe for rodent data
rodent_summary2 <- rodent_r1 %>%
  group_by(site_id, plot_id, plot_type) %>% ##first calculate number of burrows in each plot (p1, p2, d)
  summarise(num_burrows = n(), .groups = "drop") %>%
  group_by(site_id, plot_type) %>% ##then calculate average burrows with each plot type (f vs d)
  summarise(avg_burrows = mean(num_burrows), .groups = "drop")

#Add number of burrows to surveys dataframe
surveys2 <- surveys2 %>%
  left_join(rodent_summary2, by = c("site_id", "plot_type")) %>%
   mutate(avg_burrows = replace_na(avg_burrows, 0))


```

## 3. Calculate difference in burrow abundance between ditch and field for each site

```{r}
burrow_differences <- surveys2 %>%
  # Keep only the columns needed for this calculation
  select(site_id, cover_type, plot_type, avg_burrows) %>%
  
  # Pivot plot_type values (ditch, field) into their own columns
  pivot_wider(
    names_from = plot_type, 
    values_from = avg_burrows
  ) %>%
  
  # Calculate the difference (e.g., Ditch minus Field)
  mutate(
    ditch_field_diff = ditch - field
  )

#add date to dataframe
# Pivot your dates wide so each site has exactly one row for each site
site_dates_wide <- site_dates %>%
  mutate(plot_type = paste0("date_", plot_type)) %>% # turns into "date_ditch" and "date_field"
  pivot_wider(names_from = plot_type, values_from = date)

# Join it to your differences dataframe
burrow_differences <- burrow_differences %>%
  left_join(site_dates_wide, by = "site_id") %>%
  mutate(date_ditch = as.Date(date_ditch),
         date_field = as.Date(date_field)) %>%
  mutate(avg_date = pmin(date_ditch, date_field) + (abs(date_ditch - date_field) / 2))

range(burrow_differences$date_ditch) ##april 17-may 6
range(burrow_differences$date_field) ##april 1 - may 23
```

## 4. Plot

```{r}
# Plotting the trajectories
ggplot(burrow_differences, aes(x = avg_date, y = ditch_field_diff)) +
  geom_line(color = "grey50", alpha = 0.6, size = 0.8) +
  geom_point(shape = 21, size = 5, fill = "black", stroke = 1) +
  labs(
    x = "Survey date\n(averaged between field and ditch surveys)",
    y = "Ditch burrow - field burrow",
    fill = "Cover type"
  ) +
  geom_smooth(method = "lm", color = "blue", fill = "blue", alpha = 0.15, se = TRUE) +
  # 3. Split into panels by cover type
  facet_wrap(~cover_type) +
  theme_classic(base_size = 20) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## 5. Stats

```{r}
#filter for each cover type
bare_diffs <- burrow_differences %>% filter(cover_type == "bare")
grass_diffs <- burrow_differences %>% filter(cover_type == "herbaceous")

# Non-parametric trend test
cor.test(~ ditch_field_diff + as.numeric(avg_date), data = bare_diffs, method = "kendall")
##tau = 0.53, p = 0.2065
cor.test(~ ditch_field_diff + as.numeric(avg_date), data = grass_diffs, method = "kendall")
##tau = -0.4, p = 0.4833
```

# Plot the difference between grass and bare vs round

Use within-field date (not ditch data) only since floralVSburrow paper did not include ditch data.

## 1. Create dataframe for rodent surveys for all sites and rounds

```{r}
#create vector for site names not "sandhu_ladnertrunk" and "lbf_ladnertrunk" since they were not sampled during round 1
site_names2 <- c("bates_57b", "bks_56", "chong_68", "bks_68", "chong_96", "bks_96", "gwb_neaves", "bks_neaves", "avtar_168", "surjit_168")

#create vector for round
round <-c("2024_r1", "2024_r2")


#combine vectors into dataframe 
samples_plot3 <- tibble(site_id = site_names2) %>%
  expand_grid(round)

#add area and cover type
samples_plot3 <- samples_plot3 %>%
  left_join(site_info %>% select(site_id, area, cover_type), by = c("site_id")) %>%
  mutate(area = ifelse(area == "pittmeadows_1", "pittmeadows", area)) %>%
  mutate(area = ifelse(area == "surrey_1", "surrey", area))

#correct cover type formatting
surveys3 <- samples_plot3 %>%
   mutate(
     cover_type = as.factor(cover_type),
     cover_type = fct_recode(cover_type, "bare" = "bare", "herbaceous" = "grass")
   )
```

## 2. Calculate burrow abundance in each area/site/covertype/round

```{r}
#filter to within-field data 
rodent_field <- rodent %>%
  filter(plot_type == "field")

#create summary dataframe for rodent data
rodent_summary3 <- rodent_field %>%
  group_by(site_id, cover_type, area, round) %>%
  summarise(num_burrows = n(), .groups = "drop")

#fix covertype formatting
rodent_summary3 <- rodent_summary3 %>%
   mutate(
     cover_type = as.factor(cover_type),
     cover_type = fct_recode(cover_type, "bare" = "bare", "herbaceous" = "grass")
   )

#Add number of burrows to surveys dataframe
surveys3 <- surveys3 %>%
  left_join(rodent_summary3, by = c("site_id", "area", "cover_type", "round")) %>%
   mutate(num_burrows = replace_na(num_burrows, 0))
```

## 3. Calculate difference between bare and grassy sites within each area

```{r}
burrow_differences_covertype <- surveys3 %>%
  # Keep only the columns needed for this calculation
  select(area, cover_type, round, num_burrows) %>%
  
  # Pivot plot_type values (ditch, field) into their own columns
  pivot_wider(
    names_from = cover_type, 
    values_from = num_burrows
  ) %>%
  
  # Calculate the difference (grass minus bare)
  mutate(
    herb_bare_diff = herbaceous - bare
  )
```

## 4. Plot

```{r}
ggplot(burrow_differences_covertype, aes(x = round, y = herb_bare_diff)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.5) +
  # 2. Add jittered points so you can see the individual 'area' blocks
  geom_jitter(width = 0.1, size = 3, alpha = 0.8, color = "black", shape = 21, fill = "white") +
  # 3. Add a reference line at 0 (where herbaceous count == bare count)
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.8) +
  labs(
    x = "Sampling round",
    y = "Burrow abundance difference\nbetween cover types\n(herbaceous - bare)"
  ) +
  # Using a clean theme and removing the redundant legend
  theme_classic(base_size = 18) 
```

# Plot difference in burrow abundance between grassy and bare sites over date (only in R2)

Only include 2024_r2 data since floralVSburrow paper only used rodent r2 data in analyses.

## 1. Calculate burrow abundance in each site in r2

```{r}
#filter to r2
rodent_r2 <- rodent %>%
  filter(round == "2024_r2")

#calculate number of burrows per site
rodent_summary4 <- rodent_r2 %>%
  group_by(site_id) %>%
  summarise(num_burrows = n(), .groups = "drop")
```

## 2. Create dataframe for all sites

```{r}
samples_plot4 <- tibble(site_id = site_names)

#add area and cover type
samples_plot4 <- samples_plot4 %>%
  left_join(site_info %>% select(site_id, area, cover_type), by = c("site_id")) %>%
  mutate(area = ifelse(area == "pittmeadows_1", "pittmeadows", area)) %>%
  mutate(area = ifelse(area == "surrey_1", "surrey", area))


#join burrow numbers to site names df
rodent_summary4 <- samples_plot4 %>%
  left_join(rodent_summary4, by = c("site_id")) %>%
   mutate(num_burrows = replace_na(num_burrows, 0))

#correct cover type formatting
rodent_summary4 <- rodent_summary4 %>%
   mutate(
     cover_type = as.factor(cover_type),
     cover_type = fct_recode(cover_type, "bare" = "bare", "herbaceous" = "grass")
   )

```

## 3. Calculate difference between grassy and bare sites within each area

```{r}
burrow_differences_covertype_time <- rodent_summary4 %>%
  select(area, cover_type, num_burrows) %>%
  # Pivot cover_type values into their own columns
  pivot_wider(
    names_from = cover_type, 
    values_from = num_burrows
  ) %>%
  
  # Calculate the difference (e.g., Ditch minus Field)
  mutate(
    herb_bare_diff = herbaceous - bare
  )
```

## 4. Add average date between grassy and bare surveys to dataframe

```{r}
#find average date between grassy and bare sites
site_dates2 <- rodent_dates_fVb %>% #use floral vs burrow dataframe
  group_by(site_id) %>%
  summarise(
    date = first(date), #select first date for each site (p1)
    .groups = "drop"
  )

#add area and cover type
site_dates2 <- site_dates2 %>%
  left_join(site_info %>% select(site_id, area, cover_type), by = c("site_id")) %>%
  mutate(area = ifelse(area == "pittmeadows_1", "pittmeadows", area)) %>%
  mutate(area = ifelse(area == "surrey_1", "surrey", area))

#correct cover type formatting
site_dates2 <- site_dates2 %>%
   mutate(
     cover_type = as.factor(cover_type),
     cover_type = fct_recode(cover_type, "bare" = "bare", "herbaceous" = "grass")
   )

# Pivot your dates wide so each site has exactly one row for each area
site_dates_wide2 <- site_dates2 %>%
  select(area, cover_type, date) %>%
  mutate(cover_type = paste0("date_", cover_type)) %>% # turns into "date_ditch" and "date_field"
  pivot_wider(names_from = cover_type, values_from = date)

#join to burrow differences df
burrow_differences_covertype_time <- burrow_differences_covertype_time %>%
  # Pull the dates in by matching site_id
  left_join(site_dates_wide2, by = c("area")) %>%
  mutate(date_herbaceous = as.Date(date_herbaceous),
         date_bare = as.Date(date_bare)) %>%
  mutate(avg_date = pmin(date_herbaceous, date_bare) + (abs(date_herbaceous - date_bare) / 2))
```

## 5. Plot

```{r}
ggplot(burrow_differences_covertype_time, aes(x = avg_date, y = herb_bare_diff, group = 1)) +
  geom_line(color = "grey50", alpha = 0.6, size = 0.8) +
  geom_point(shape = 21, size = 5, fill = "black", stroke = 1) +
  labs(
    x = "Survey date\n(averaged between herb and bare surveys)",
    y = "Burrow count difference\nbetween cover types\n(herbaceous - bare)"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40", linewidth = 0.8) +
  geom_smooth(method = "lm", color = "blue", fill = "blue", alpha = 0.15, se = TRUE) +
  theme_classic(base_size = 20) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## 6. Stats

```{r}
# Non-parametric trend test
cor.test(~ herb_bare_diff + as.numeric(as.Date(avg_date)), data = burrow_differences_covertype_time, method = "kendall")
##tau = -0.33, p = 0.4694
```
