---
title: "survey timing"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load packages

```{r}
#load package
library(tidyverse)   #data manipulation, visualization etc. (dplyr, ggplot2, tidyr)
library(lubridate) #date <-> julian date

```

# Create lists of dates

```{r}
dates <- read.csv("06_cleandata/survey_dates.csv")

#create unique row for each plot id at each site for each survey
dates_long <- dates %>%
  pivot_longer(
    cols = c(p1, p2, d),
    names_to = "plot_id",
    values_to = "plot_value"
  ) %>%
  rename(area = Region.name, ##rename columns
         date = Date,
         site_id = Site.name,
         survey_type = Survey.type,
         round = Round,
         dVf = ditchVSfield_analysis,
         fVb = floralVSburrow_analysis,
         ) %>%
  filter(plot_value != "n") %>% ##filter to only rows where plot_type = y
  select(date, area, site_id, plot_id, round, survey_type, dVf, fVb) %>% #select columns to keep 
  mutate(plot_type = if_else(plot_id == "d", "ditch", "field")) %>% #add plot type column 
  mutate(plot_type = as.factor(plot_type)) %>% #make it a factor 
  mutate(date = as.Date(date)) %>% #fix date formatting
  mutate(julianDate = julian.Date(date)) #create column for julian date

#create dataframe for just rodent dates
##dates for surveys included in ditchVSfield (round 1) analyses:
rodent_dates_dVf <- dates_long %>%
  filter(survey_type == "rodent" & dVf == "y") %>%
  select(date, area, site_id, plot_id, round, plot_type, julianDate)

rodent_dates_dVf %>%
  count(area, site_id, round, plot_id) %>% #looks good

write_csv(rodent_dates_dVf, file = file.path("06_cleandata","rodent_dates_dVf.csv"))

##dates for surveys included in floralVSburrow (round 2) analyses:
rodent_dates_fVb <- dates_long %>%
  filter(survey_type == "rodent" & fVb == "y") %>%
  select(date, area, site_id, plot_id, round, plot_type, julianDate)

#check
rodent_dates_fVb %>%
  count(area, site_id, round, plot_id) %>% #looks good

#save
write_csv(rodent_dates_fVb, file = file.path("06_cleandata","rodent_dates_fVb.csv"))

#create dataframe for just bee dates
bee_dates <- dates_long %>%
  filter(survey_type == "bombus") %>%
  select(date, area, site_id, plot_id, round, plot_type, julianDate)

#check
bee_dates %>%
  count(area, site_id, round, plot_id) %>%
  View() ##looks good

#save
write_csv(bee_dates, file = file.path("06_cleandata","bee_dates.csv"))

#create dataframe for just veg dates
veg_dates <- dates_long %>%
  filter(survey_type == "vegetation") %>%
  select(date, area, site_id, plot_id, round, plot_type, julianDate)

#check
veg_dates %>%
  count(area, site_id, round, plot_id) %>%
  View() ##looks good

#save
write_csv(veg_dates, file = file.path("06_cleandata","veg_dates.csv"))


```

# Check survey timing

## Rodents

### Check dates apart for ditch vs field surveys.

```{r}
rod_days_apart_dVf <- rodent_dates_dVf %>%
  #1. choose one within-field plot so that there is only one date for field
  filter(plot_id != "p2") %>%
  # 2. Pivot plot_type wide so ditch and field dates are on the same row
  pivot_wider(
    id_cols = c(area, site_id, round),
    names_from = plot_type,
    values_from = julianDate
  ) %>%
  
  # 3. Calculate the absolute number of days apart
  mutate(
    days_apart = abs(as.numeric(ditch - field))
  ) %>%
  
  # 4. Clean up column names for presentation
  rename(
    `Ditch Date` = ditch,
    `Field Date` = field,
    `Days Apart` = days_apart
  )

View(rod_days_apart_dVf)

#save
write_csv(rod_days_apart_dVf, file = file.path("07_outputs","rodent_days_apart_dVf.csv"))
```

### Check dates apart for grass vs bare in each area

```{r}
rod_days_apart_fVb <- rodent_dates_fVb %>%
  filter(plot_id == "p1") %>%
  #add cover type to dataframe
  left_join(site_info %>% dplyr::select(cover_type, site_id), by = c("site_id"))%>%
  # 2. Pivot plot_type wide so ditch and field dates are on the same row
  pivot_wider(
    id_cols = c(area, round),
    names_from = cover_type,
    values_from = julianDate
  ) %>%
  
  # 3. Calculate the absolute number of days apart
  mutate(
    days_apart = abs(as.numeric(grass - bare))
  ) %>%
  
  # 4. Clean up column names for presentation
  rename(
    `Herb Date` = grass,
    `Bare Date` = bare,
    `Days Apart` = days_apart
  )

View(rod_days_apart_fVb)

#save
write_csv(rod_days_apart_fVb, file = file.path("07_outputs","rodent_days_apart_fVb.csv"))
```

## Bees

### Check dates apart for ditch vs field surveys.

```{r}
bee_days_apart_dVf <- bee_dates %>%
  #1. choose one within-field plot so that there is only one date for field
  filter(plot_id != "p2") %>%
  # 2. Pivot plot_type wide so ditch and field dates are on the same row
  pivot_wider(
    id_cols = c(area, site_id, round),
    names_from = plot_type,
    values_from = julianDate
  ) %>%
  
  # 3. Calculate the absolute number of days apart
  mutate(
    days_apart = abs(as.numeric(ditch - field))
  ) %>%
  
  # 4. Clean up column names for presentation
  rename(
    `Ditch Date` = ditch,
    `Field Date` = field,
    `Days Apart` = days_apart
  )

View(bee_days_apart_dVf)

#save
write_csv(bee_days_apart_dVf, file = file.path("07_outputs","bee_days_apart_dVf.csv"))
```

### Check dates apart for grass vs bare in each area

```{r}
bee_days_apart_fVb <- bee_dates %>%
  filter(plot_id == "p1") %>% #choose one plot id
  #add cover type to dataframe
  left_join(site_info %>% dplyr::select(cover_type, site_id), by = c("site_id"))%>%
  # 2. Pivot plot_type wide so ditch and field dates are on the same row
  pivot_wider(
    id_cols = c(area, round),
    names_from = cover_type,
    values_from = julianDate
  ) %>%
  
  # 3. Calculate the absolute number of days apart
  mutate(
    days_apart = abs(as.numeric(grass - bare))
  ) %>%
  
  # 4. Clean up column names for presentation
  rename(
    `Herb Date` = grass,
    `Bare Date` = bare,
    `Days Apart` = days_apart
  )

View(bee_days_apart_fVb)

#save
write_csv(bee_days_apart_fVb, file = file.path("07_outputs","bee_days_apart_fVb.csv"))
```

## Veg

### Check dates apart for ditch vs field surveys.

```{r}
veg_days_apart_dVf <- veg_dates %>%
  #1. choose one within-field plot so that there is only one date for field
  filter(plot_id != "p2") %>%
  # 2. Pivot plot_type wide so ditch and field dates are on the same row 
  pivot_wider(
    id_cols = c(area, site_id, round),
    names_from = plot_type,
    values_from = julianDate
  ) %>%
  
  # 3. Calculate the absolute number of days apart
  mutate(
    days_apart = abs(as.numeric(ditch - field))
  ) %>%
  
  # 4. Clean up column names for presentation
  rename(
    `Ditch Date` = ditch,
    `Field Date` = field,
    `Days Apart` = days_apart
  )

View(veg_days_apart_dVf)

#save
write_csv(veg_days_apart_dVf, file = file.path("07_outputs","veg_days_apart_dVf.csv"))
```

### Check dates apart for grass vs bare in each area

```{r}
veg_days_apart_fVb <- veg_dates %>%
  filter(plot_id == "p1") %>% #choose one plot id
  #add cover type to dataframe
  left_join(site_info %>% dplyr::select(cover_type, site_id), by = c("site_id"))%>%
  # 2. Pivot plot_type wide so ditch and field dates are on the same row
  pivot_wider(
    id_cols = c(area, round),
    names_from = cover_type,
    values_from = julianDate
  ) %>%
  
  # 3. Calculate the absolute number of days apart
  mutate(
    days_apart = abs(as.numeric(grass - bare))
  ) %>%
  
  # 4. Clean up column names for presentation
  rename(
    `Herb Date` = grass,
    `Bare Date` = bare,
    `Days Apart` = days_apart
  )

View(veg_days_apart_fVb)

#save
write_csv(veg_days_apart_fVb, file = file.path("07_outputs","veg_days_apart_fVb.csv"))
```

## Check dates apart for p1 and p2

Rodent

R1

```{r}
rod_days_apart_r1_p1p2 <- rodent_dates_dVf %>%
  filter(plot_id != "d") %>%
  # 2. Pivot plot_type wide so p1 and p2 dates are on the same row
  pivot_wider(
    id_cols = c(area, site_id, round),
    names_from = plot_id,
    values_from = julianDate
  ) %>%
  
  # 3. Calculate the absolute number of days apart
  mutate(
    days_apart = abs(as.numeric(p1 - p2))
  ) %>%
  
  # 4. Clean up column names for presentation
  rename(
    `P1 Date` = p1,
    `P2 Date` = p2,
    `Days Apart` = days_apart
  )

View(rod_days_apart_r1_p1p2) ##always done on same day!



#save
write_csv(rod_days_apart_r1_p1p2, file = file.path("07_outputs","rodent_days_apart_r1_p1p2.csv"))
```

R2

```{r}
rod_days_apart_r2_p1p2 <- rodent_dates_fVb %>%
  filter(plot_id != "d") %>%
  # 2. Pivot plot_type wide so p1 and p2 dates are on the same row
  pivot_wider(
    id_cols = c(area, site_id, round),
    names_from = plot_id,
    values_from = julianDate
  ) %>%
  
  # 3. Calculate the absolute number of days apart
  mutate(
    days_apart = abs(as.numeric(p1 - p2))
  ) %>%
  
  # 4. Clean up column names for presentation
  rename(
    `P1 Date` = p1,
    `P2 Date` = p2,
    `Days Apart` = days_apart
  )

View(rod_days_apart_r2_p1p2) ##always done on same day!



#save
write_csv(rod_days_apart_r2_p1p2, file = file.path("07_outputs","rodent_days_apart_r2_p1p2.csv"))
```

Bee

```{r}
bee_days_apart_p1p2 <- bee_dates %>%
  filter(plot_id != "d") %>%
  # 2. Pivot plot_type wide so p1 and p2 dates are on the same row
  pivot_wider(
    id_cols = c(area, site_id, round),
    names_from = plot_id,
    values_from = julianDate
  ) %>%
  
  # 3. Calculate the absolute number of days apart
  mutate(
    days_apart = abs(as.numeric(p1 - p2))
  ) %>%
  
  # 4. Clean up column names for presentation
  rename(
    `P1 Date` = p1,
    `P2 Date` = p2,
    `Days Apart` = days_apart
  )

View(bee_days_apart_p1p2) ##always done on same day!


#save
write_csv(bee_days_apart_p1p2, file = file.path("07_outputs","bee_days_apart_p1p2.csv"))
```

Veg

```{r}
veg_days_apart_p1p2 <- veg_dates %>%
  filter(plot_id != "d") %>%
  # 2. Pivot plot_type wide so p1 and p2 dates are on the same row
  pivot_wider(
    id_cols = c(area, site_id, round),
    names_from = plot_id,
    values_from = julianDate
  ) %>%
  
  # 3. Calculate the absolute number of days apart
  mutate(
    days_apart = abs(as.numeric(p1 - p2))
  ) %>%
  
  # 4. Clean up column names for presentation
  rename(
    `P1 Date` = p1,
    `P2 Date` = p2,
    `Days Apart` = days_apart
  )

View(veg_days_apart_p1p2) ##almost always done on same day. In R1, p1 and p2 at one site were surveyed 5 days apart and p1 and p2 at another site were surveyed 3 days apart.

#save
write_csv(veg_days_apart_p1p2, file = file.path("07_outputs","veg_days_apart_p1p2.csv"))
```

Calculate days between veg survey and bee survey at each site/round

```{r}
veg_bee_gap <- bind_rows(
  bee_dates %>% mutate(survey_type = "Bee"),
  veg_dates %>% mutate(survey_type = "Veg")
) %>%
  
  # 2. Strict filter for p1 to secure exactly one date per site visit
  filter(plot_id == "p1") %>%
  
  # 3. Keep only the core tracking variables
  select(area, site_id, round, survey_type, julianDate) %>%
  
  # 4. Pivot wide so 'Bee' and 'Veg' julian dates sit side-by-side on the same row
  pivot_wider(
    id_cols = c(area, site_id, round),
    names_from = survey_type,
    values_from = julianDate
  ) %>%
  
  # 5. Calculate the absolute number of days apart between the two surveys
  mutate(
    days_apart = abs(as.numeric(Veg - Bee))
  ) %>%
  
  # 6. Clean up column names for clean presentation/saving
  rename(
    `Bee Date (Julian)` = Bee,
    `Veg Date (Julian)` = Veg,
    `Days Apart` = days_apart
  )

# View the output dataframe
View(veg_bee_gap)
```
