---
title: "exploratory_ditch_vs_field"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Library/CloudStorage/OneDrive-UBC/MSc/12_analysis/rodent-bee-blueberry")
```

# Load data

```{r}
#load packages
library(ggplot2) #for plots
library(dplyr) #for working with dataframes
library(tidyr) #for working with datasets
library(lme4) #for mixed effects models
library(ggpubr) #for adding significance asterisks to graph
library(stringr)

#read in data
data <- read.csv("06_cleandata/combined_plot_data.csv")
rodent <- read.csv("06_cleandata/rodent_clean.csv")

#check that variables are in appropriate format
str(data)

#change character columns to factors
data <- data %>%
   mutate(
     site_id = as.factor(site_id),
     round = as.factor(round),
     plot_id = as.factor(plot_id),
     bee_date = as.Date(bee_date),
     area = as.factor(area))

#add column for ditch vs field
data <- data %>%
  mutate(field_ditch = case_when(
    str_detect(plot_id, "p1|p2") ~ "field",
    str_detect(plot_id, "d") ~ "ditch",
    TRUE ~ NA_character_
  ))

#add column for ditch vs field
rodent <- rodent %>%
  mutate(field_ditch = case_when(
    str_detect(plot_id, "p1|p2") ~ "field",
    str_detect(plot_id, "d") ~ "ditch",
    TRUE ~ NA_character_
  ))

#add column for small holes (dm)
rodent <- rodent %>%
  mutate(burrow_size = case_when(
    diameter_cm <= 1.5 ~ "small",
    diameter_cm > 1.5 ~ "big",
    TRUE ~ NA_character_
  ))
```

# Compare ditch vs field

## Num NSQ

```{r}
#aggregate across site and round to account for 2 field plots at each site vs 1 ditch plot at each site
data_site_avg <- data %>%
  group_by(site_id, round, field_ditch) %>%
  summarise(mean_nsq = mean(nest_searching_queens, na.rm = TRUE), .groups = "drop")

#plot
ggplot(data_site_avg, aes(x = field_ditch, y = mean_nsq)) +
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Plot Type") +
  ylab("Mean NSQ per Site") +
  stat_compare_means(method = "t.test", label = "p.format", size = 3.7)
```

## Num workers

```{r}
data_site_avg2 <- data %>%
  group_by(site_id, round, field_ditch) %>%
  summarise(mean_workers = mean(workers, na.rm = TRUE), .groups = "drop")

ggplot(data_site_avg2, aes(x = field_ditch, y = mean_workers)) +
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Plot Type") +
  ylab("Mean Workers per Site") +
  stat_compare_means(method = "t.test", label = "p.format", size = 3.7, vjust = 0)
```

## Num burrows

```{r}
data_site_avg3 <- data %>%
  group_by(site_id, round, field_ditch) %>%
  summarise(mean_burrows = mean(num_burrows, na.rm = TRUE), .groups = "drop")

ggplot(data_site_avg3, aes(x = field_ditch, y = mean_burrows)) +
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Plot Type") +
  ylab("Mean burrow number per Site") +
  stat_compare_means(method = "t.test", label = "p.format", size = 3.7, vjust = 0)
```

## Burrow diameter

```{r}
data_site_avg4 <- rodent %>%
  group_by(site_id, round, field_ditch) %>%
  summarise(mean_burrow_dia = mean(diameter_cm, na.rm = TRUE), .groups = "drop")

ggplot(data_site_avg4, aes(x = field_ditch, y = mean_burrow_dia)) +
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Plot Type") +
  ylab("Mean burrow diameter per Site") +
  stat_compare_means(method = "t.test", label = "p.format", size = 3.7, vjust = 0)
```

Burrow size (vole vs dm) vs plot type

```{r}
ggplot(rodent, aes(fill = burrow_size, x = field_ditch)) +
  geom_bar(position = "dodge") +
  labs(,
    x = "Plot Type",
    y = "Count",
    fill = "Burrow size"
  ) +
  theme_minimal()
```

Num dm holes (\<1.5cm) across plot types

```{r}
data_site_avg5 <- data %>%
  group_by(site_id, round, field_ditch) %>%
  summarise(mean_dm = mean(num_dm_size, na.rm = TRUE), .groups = "drop")

ggplot(data_site_avg5, aes(x = field_ditch, y = mean_dm)) +
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Plot Type") +
  ylab("Mean dm burrows (1.5cm) per Site") +
  stat_compare_means(method = "t.test", label = "p.format", size = 3.7, vjust = 0)
```

## Bombus-relevant floral abundance

```{r}
data_site_avg6 <- data %>%
  group_by(site_id, round, field_ditch) %>%
  summarise(mean_floral_abun = mean(bombus_floral_abun, na.rm = TRUE), .groups = "drop")

ggplot(data_site_avg6, aes(x = field_ditch, y = mean_floral_abun)) +
  geom_boxplot() +
  theme_classic(base_size = 15) +
  xlab("Plot Type") +
  ylab("Mean floral abundance per Site") +
  stat_compare_means(method = "t.test", label = "p.format", size = 3.7, vjust = 0)
```
